---
title: "Why functional fine-mapping (versus scalar SuSiE on summaries)"
author: "Gao Wang and contributors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Why functional fine-mapping (versus scalar SuSiE on summaries)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.align = "center", dpi = 120
)
```

Two common shortcuts for fine-mapping a functional response with
scalar SuSiE both lose information. This vignette demonstrates the
loss on a small simulation and shows that `fsusie()` recovers the
causal SNP that the scalar shortcuts miss.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## The two scalar shortcuts

1. **Per-position SuSiE, then aggregate**: fit `susie(X, y_t)`
   independently for each position `t = 1, ..., T`, then take the
   per-SNP maximum or sum of PIPs. Loses the smoothness of the effect
   profile and inflates false positives because each fit is
   under-powered.
2. **Top-PC SuSiE**: project `Y` onto its top principal component and
   fit `susie(X, Y %*% v_1)`. Keeps the strongest mode of variation in
   `Y` but discards the SNP-driven signal whenever the SNP effect
   loads on a different direction.

## Simulation

Two causal SNPs whose effects are localised to non-overlapping
position windows. The combined signal in any single position is weak;
the combined signal across positions is strong.

```{r simulate}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); J <- ncol(X); T_m <- 64

positions <- seq_len(T_m)
shape_left  <- exp(-((positions - 16)^2) / (2 * 4^2))
shape_right <- exp(-((positions - 48)^2) / (2 * 4^2))

beta <- matrix(0, J, T_m)
beta[42, ] <- 0.6 * shape_left
beta[97, ] <- 0.6 * shape_right

# Add nuisance variation in a direction orthogonal to the SNP effect
# so the top PC is dominated by noise, not by the SNP signal.
nuisance <- rnorm(n) %o% sin(2 * pi * positions / T_m)
Y <- X %*% beta + 1.5 * nuisance + matrix(rnorm(n * T_m, sd = 0.4), n)
```

## Per-position SuSiE

```{r per-position}
pip_per_pos <- vapply(seq_len(T_m), function(t) {
  fit_t <- susie(X, Y[, t], L = 5,
                 estimate_prior_variance    = TRUE,
                 estimate_residual_variance = TRUE,
                 verbose = FALSE)
  fit_t$pip
}, numeric(J))
pip_per_snp <- apply(pip_per_pos, 1, max)
which(pip_per_snp > 0.9)
```

## Top-PC SuSiE

```{r top-pc}
v1 <- svd(scale(Y, center = TRUE, scale = FALSE))$v[, 1]
y_pc <- as.numeric(scale(Y, center = TRUE, scale = FALSE) %*% v1)
fit_pc <- susie(X, y_pc, L = 5,
                estimate_prior_variance    = TRUE,
                estimate_residual_variance = TRUE,
                verbose = FALSE)
which(fit_pc$pip > 0.9)
```

## Functional SuSiE (the right answer)

```{r fsusie}
fit_f <- fsusie(Y, X, pos = positions, L = 5, verbose = FALSE)
which(fit_f$pip > 0.9)
length(fit_f$sets$cs)
```

## Why fsusie wins

`fsusie()` integrates over the position dimension under a wavelet-
basis prior that adapts to the actual smoothness of the effect curve.
Two causal SNPs that each affect a localised window get the full
benefit of within-window pooling without bleeding into SNPs whose
effects live in a different window. The scalar shortcuts above either
average effects out (top-PC) or fragment them (per-position).

The same argument scales up: if a SNP affects multiple correlated
modalities (e.g., methylation at several nearby CpGs and expression
of a single gene), the joint mfsusie fit pools across modalities for
the same reason. See [the multi-modality intro](mfsusie_intro.html).
