---
title: "Single-outcome functional fine-mapping"
author: "Gao Wang and contributors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Single-outcome functional fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 3, fig.align = "center", dpi = 120
)
```

## Overview

`fsusie()` fits the Functional Sum of Single Effects (fSuSiE) model to
a single functional response. It extends `susieR::susie` to outcomes
sampled at multiple positions, such as DNA methylation profiles, ATAC-
seq tracks, or expression along a transcript. The wavelet basis lets
the model share information across nearby positions while keeping the
SuSiE per-effect prior structure intact.

`fsusie()` is a thin wrapper around `mfsusie()` for the special case
of one outcome. The argument order is `(Y, X, pos, ...)` so the
response comes first, mirroring how single-response models are
typically called.

## Bundled fixture

The package ships `rnaseq_example`, a simulated cis-eQTL fixture:
200 SNPs over a real LD scaffold (`susieR::N3finemapping$X`),
128 evenly-spaced exon-body positions, one causal SNP with a
peaked Gaussian effect over ~30 positions.

```{r setup, message = FALSE}
library(mfsusieR)
set.seed(1)
data(rnaseq_example)
str(rnaseq_example, max.level = 1)
```

A few sample expression-coverage curves:

```{r sample-curves, fig.cap = "Five randomly-chosen sample curves from `rnaseq_example$Y`."}
matplot(rnaseq_example$pos,
        t(rnaseq_example$Y[1:5, ]), type = "l", lty = 1,
        col = c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2"),
        xlab = "position", ylab = "y")
```

## Fitting

```{r fit}
fit <- fsusie(rnaseq_example$Y, rnaseq_example$X,
              pos = rnaseq_example$pos, L = 5, verbose = FALSE)
fit$niter; fit$converged
length(fit$sets$cs)
fit$pip[rnaseq_example$causal_snps]
```

## Inspecting the fit

`mfsusie_plot()` is the unified plot function for `fsusie()` and
`mfsusie()` fits. On a single-outcome fit it produces a 2-panel
column: PIPs on top (CS-colored points) and per-effect curves on
the bottom. Use `susie_plot(fit, y = "PIP")` (re-exported from
susieR) when only the PIP panel is wanted.

```{r mfsusie-plot, fig.cap = "PIP and per-effect curves. Top: PIPs colored by CS membership. Bottom: posterior effect curves on the position grid."}
mfsusie_plot(fit)
```

Credible sets:

```{r cs}
fit$sets$cs
fit$sets$purity
```

## Smoothed effect curves and credible bands

The default plot above shows the raw inverse-DWT curves from
`coef(fit)`. After `mf_post_smooth()`, the same plot call shows
the smoothed curve and a pointwise credible band:

```{r post-smooth, fig.cap = "After mf_post_smooth(method = 'TI'): cycle-spinning translation-invariant denoising with 95% pointwise credible bands. Affected regions (band excludes zero) are marked with thick bars at the bottom."}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

See `vignette("post_processing")` for the full set of methods
(`"scalewise"`, `"TI"`, `"HMM"`) and the lfsr overlay produced by
the HMM smoother.

## Working with the fit

`coef(fit)` returns a list of `M` outcome matrices, each `L x T_m`,
one row per single effect in position space.
`predict(fit, newx = X)` reconstructs the fitted curves on new
genotypes; `fitted(fit)` returns the in-sample fits.

## Reading the fit object

```{r fields}
str(fit, max.level = 1)
```

The list of class `c("mfsusie", "susie")` carries the standard SuSiE
fields plus per-outcome wavelet metadata used by `predict.mfsusie`,
`coef.mfsusie`, and `fitted.mfsusie`.

