---
title: "GTEx-style RNA-seq cis-eQTL fine-mapping"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{GTEx-style RNA-seq cis-eQTL fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.align = "center", dpi = 120
)
```

This vignette walks through a GTEx-style cis-eQTL fine-mapping
analysis using `fsusie()`. The phenotype is RNA-seq read coverage
along a gene body; `fsusie()` treats it as a single functional
response and returns credible sets of variants that drive
coordinated changes in expression at distinct positions of the
gene.

The case studies in `fsusie-experiments/scripts_plot/GeTX/`
(AHCYL1, SCD, HSP90AA1, GALNT6, ABHD17A, STMN2) all share this
shape: a multi-CS fit on log-transformed RNA-seq counts over a
cis window. We reproduce the same workflow here on a bundled
simulated fixture (`gtex_example`) shaped after those examples.

## Bundled GTEx fixture

```{r setup, message = FALSE}
library(mfsusieR)
data(gtex_example)
str(gtex_example, max.level = 1)
```

200 SNPs over a real LD scaffold (`susieR::N3finemapping$X`),
256 gene-body positions, three causal SNPs at distinct positions
(5', mid-body, 3') with localized peak effects:

```{r show-truth, fig.width = 7, fig.height = 3, fig.cap = "Three true per-position effects on the gene-body grid."}
matplot(gtex_example$pos, t(gtex_example$causal_betas),
        type = "l", lty = 1, lwd = 1.5,
        col = c("#E69F00", "#56B4E9", "#009E73"),
        xlab = "gene-body position", ylab = "true effect",
        main = "Ground-truth effect curves")
abline(h = 0, lty = 2, col = "grey60")
legend("topright",
       legend = paste0("SNP ", gtex_example$causal_snps),
       col = c("#E69F00", "#56B4E9", "#009E73"),
       lwd = 1.5, bty = "n")
```

## Fit

We use `L = 15` candidate single effects and `L_greedy = 5`. The
`L_greedy` knob (added on the susieR side of the migration plan)
runs a greedy filter pass on the full-`L` IBSS and is the
recommended default for most cis-windows.

```{r fit}
fit <- fsusie(gtex_example$Y, gtex_example$X,
              pos = gtex_example$pos,
              L = 15, L_greedy = 5,
              verbose = TRUE)
fit$niter; fit$converged
length(fit$sets$cs)
fit$sets$cs
fit$pip[gtex_example$causal_snps]
```

All three causal SNPs receive PIP = 1 and land in three separate
credible sets.

## Inspecting the fit

The unsmoothed `mfsusie_plot()` shows PIPs (top panel, points
colored by CS membership) and the per-CS effect curves recovered
by inverse DWT (bottom panel):

```{r plot-raw, fig.width = 6, fig.height = 5, fig.cap = "Raw fSuSiE fit. Top: PIPs across SNPs colored by credible set. Bottom: posterior effect curves on the gene-body grid."}
mfsusie_plot(fit)
```

Adding `mf_post_smooth(method = "TI")` denoises the wavelet
posterior with cycle-spinning translation-invariant denoising
and adds 95% pointwise credible bands. The three CS effect
curves now show clean peaks at distinct positions, mirroring
the multi-CS structure highlighted in the AHCYL1 / SCD figures
of the fsusie-experiments paper:

```{r plot-smoothed, fig.width = 6, fig.height = 5, fig.cap = "TI-smoothed fSuSiE fit on the simulated GTEx-style fixture. Each CS captures a distinct gene-body region; bands and bottom-bar affected regions surface where each effect is significantly different from zero."}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

## Why functional fine-mapping helps here

The fine-mapping signal is spread across many positions of the
gene. A scalar SuSiE fit on `rowSums(Y)` (total expression)
mixes every per-position effect into one number and loses the
per-region structure. The functional model preserves it:

```{r susie-comparison}
library(susieR)
fit_scalar <- susie(gtex_example$X, rowSums(gtex_example$Y),
                    L = 15, verbose = FALSE)
fit_scalar$sets$cs
fit_scalar$pip[gtex_example$causal_snps]
```

Per-position effects (the GTEx case studies' value add) are not
recoverable from the scalar fit; they are the natural output of
`fsusie()`.

## Genome-browser overlay (optional)

For a publication-grade genome-browser figure with ideogram,
genome axis, fitted effect with credible band, allele-stratified
observed coverage, and a gene-region track, `mfsusieR` ships
worked-example scripts in
`system.file("scripts/gtex_case_study", package = "mfsusieR")`
(AHCYL1, SCD, HSP90AA1, STMN2, ABHD17A). They layer an
`fsusie()` fit onto `Gviz` tracks. The Bioconductor packages
they depend on (`Gviz`, `GenomicFeatures`, `AnnotationDbi`,
`TxDb.Hsapiens.UCSC.hg38.knownGene`, `org.Hs.eg.db`, `smashr`)
are listed as `Suggests`; install them with
`BiocManager::install(...)` if you want the genome-browser
overlay. The base-R `mfsusie_plot()` shown above is the lighter
default and is what most users need.
