---
title: "Single-modality functional fine-mapping with fsusie()"
author: "Gao Wang and contributors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Single-modality functional fine-mapping with fsusie()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 3, fig.align = "center", dpi = 120
)
```

## Overview

`fsusie()` fits the Functional Sum of Single Effects (fSuSiE) model to
a single functional response. It extends `susieR::susie` to traits
sampled at multiple positions, such as DNA methylation profiles, ATAC-
seq tracks, or expression along a transcript. The wavelet basis lets
the model share information across nearby positions while keeping the
SuSiE per-effect prior structure intact.

`fsusie()` is a thin wrapper around `mfsusie()` for the special case
of one modality. The argument order is `(Y, X, pos, ...)` so the
response comes first, mirroring how single-response models are
typically called.

## Simulating a fixture

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

Use the `N3finemapping` SNP genotype matrix from susieR for a
realistic LD pattern; simulate a wavelet-smoothable functional effect
on a length-128 grid.

```{r simulate}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(200)]
n <- nrow(X); J <- ncol(X)
T_m <- 128

# Two causal SNPs, smooth bell-shaped effect curve.
positions <- seq_len(T_m)
shape <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 8)^2))
beta <- matrix(0, nrow = J, ncol = T_m)
beta[37, ] <-  1.5 * shape
beta[112, ] <- -1.0 * shape

Y <- X %*% beta + matrix(rnorm(n * T_m, sd = 0.5), nrow = n)
matplot(t(Y[1:5, ]), type = "l", lty = 1, col = 1:5,
        xlab = "position", ylab = "y", main = "Five sample curves")
```

## Fitting

```{r fit}
fit <- fsusie(Y, X, pos = positions, L = 5, verbose = FALSE)
fit$niter; fit$converged
length(fit$sets$cs)
```

Top PIP SNPs:

```{r pip-plot}
plot(fit$pip, pch = 16, col = ifelse(fit$pip > 0.9, "red", "grey50"),
     xlab = "SNP index", ylab = "PIP")
abline(v = c(37, 112), lty = 2, col = "blue")
legend("topright", legend = c("True causal", "PIP > 0.9"),
       lty = c(2, NA), pch = c(NA, 16),
       col = c("blue", "red"), bty = "n", cex = 0.8)
```

Credible sets:

```{r cs}
fit$sets$cs
fit$sets$purity
```

## Inspecting effect curves

`coef(fit)` returns a list of `M` modality matrices, each `L x T_m`,
one row per single effect in position space.

```{r effects}
B_hat <- coef(fit)[[1]]
matplot(t(B_hat), type = "l", lty = 1,
        xlab = "position", ylab = "effect",
        main = "Estimated single-effect curves")
matlines(t(beta[c(37, 112), ]), col = "black", lty = 2)
legend("topright", legend = c("Estimated", "True"),
       lty = c(1, 2), col = c(1, 1), bty = "n", cex = 0.8)
```

`predict(fit, newx = X)` reconstructs the fitted curves on new
genotypes; `fitted(fit)` returns the in-sample fits.

## Reading the fit object

```{r fields}
str(fit, max.level = 1)
```

The list of class `c("mfsusie", "susie")` carries the standard SuSiE
fields plus per-modality wavelet metadata used by `predict.mfsusie`,
`coef.mfsusie`, and `fitted.mfsusie`.

