---
title: "Colocalization across SuSiE-suite fits"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Colocalization across SuSiE-suite fits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

This vignette runs a formal colocalization analysis on a pair of
SuSiE-suite fine-mapping fits using `coloc::coloc.susie()`. The
recipe applies to any pair of susie-class fits — `susieR::susie()`,
`mfsusieR::fsusie()`, `mfsusieR::mfsusie()`, `mvsusieR::mvsusie()`,
and `susieR::susie_rss()` — because each returns the
`(class = "susie", $sets$cs_index, $lbf_variable)` triple that
`coloc.susie()` reads. `susie_rss()` is particularly useful for
GWAS summary-statistics fine-mapping fits.

For pre-fit covariate adjustment, see
[the covariates vignette](fsusie_covariates_adjustment.html).

For illustration we use `fsusie()` (single-outcome functional
fine-mapping); `mfsusie()` works the same way and is the natural
choice when both fits cover multiple outcomes.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
library(coloc)
set.seed(1)
```

## Data

```{r coloc-data}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X); T_m <- 64
colnames(X) <- paste0("SNP", seq_len(p))
positions <- seq_len(T_m)
shape  <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 6)^2))
beta1  <- matrix(0, p, T_m); beta1[42, ] <- 1.2 * shape
beta2  <- matrix(0, p, T_m); beta2[42, ] <- 0.9 * shape         # shared causal
beta3  <- matrix(0, p, T_m); beta3[88, ] <- 0.9 * shape         # distinct
Y1 <- X %*% beta1 + matrix(rnorm(n * T_m, sd = 0.4), n)
Y2 <- X %*% beta2 + matrix(rnorm(n * T_m, sd = 0.4), n)
Y3 <- X %*% beta3 + matrix(rnorm(n * T_m, sd = 0.4), n)
```

## Fits

```{r coloc-fit}
fit1 <- fsusie(Y1, X, pos = positions, L = 15, L_greedy = 5, verbose = TRUE)
fit2 <- fsusie(Y2, X, pos = positions, L = 15, L_greedy = 5, verbose = TRUE)
fit3 <- fsusie(Y3, X, pos = positions, L = 15, L_greedy = 5, verbose = TRUE)
```

`coloc.susie()` reads the SuSiE-class slots (`$sets$cs_index` to
identify retained effects, `$lbf_variable` for the per-effect log
Bayes factors over variants) and returns the standard
PP.H0..PP.H4 posterior table per pair of credible sets.
`coloc.susie()` expects `colnames(X)` to be present so it can
label posteriors by variant; both fits run on the same `X`, so
the labels align by construction.

## Shared causal

```{r coloc-shared}
res_shared <- coloc.susie(fit1, fit2)
res_shared$summary
```

`PP.H4.abf` is close to 1 for the (`fit1`, `fit2`) pair: a single
shared causal variant `SNP42` drives both outcomes.

## Distinct causals

```{r coloc-distinct}
res_distinct <- coloc.susie(fit1, fit3)
res_distinct$summary
```

`PP.H3.abf` is close to 1 for (`fit1`, `fit3`): both outcomes have
a causal in the locus, but the causal variants are different
(`SNP42` vs `SNP88`).

The same call works on `mfsusie()` fits, on `susieR::susie()`
fits, on `mvsusieR::mvsusie()` fits, and on `susieR::susie_rss()`
fits; the inputs `coloc.susie` reads are part of the SuSiE-class
contract.

## Notes

- This vignette assumes the user already has a pair of
  fine-mapping fits, possibly from different SuSiE-suite
  packages, and wants to test for colocalization between them.
  The natural alternative when subject-level data is available
  for all outcomes is to run `mfsusie()` jointly across them;
  the multi-outcome fit shares signal across modalities and
  identifies per-outcome and shared credible sets in one pass
  without a post-hoc combination step. `coloc.susie()` becomes
  essential when one or more inputs are summary-statistics fits
  (e.g., a GWAS `susie_rss()` fit colocalized against a
  molecular-QTL `fsusie()` fit), where the underlying
  individual-level data is not available.
- Joint `mfsusie()` fine-mapping that takes summary statistics
  from a GWAS as one of its outcomes is future work.
- For multi-outcome colocalization across more than two
  outcomes, see [the multi-outcome intro](mfsusie_intro.html),
  which fits all outcomes jointly and avoids the post-hoc
  combine.
