---
title: "Functional fine-mapping with fSuSiE"
author: "William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Functional fine-mapping with fSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

## Overview

fSuSiE extends the SuSiE Bayesian variable selection model to
functional traits. The functional response is sampled at multiple
positions of a regular grid (e.g., methylation along a CpG
island, RNA-seq coverage along a gene body, ATAC-seq across a
peak, ChIP-seq read counts on a window). fSuSiE places a
wavelet-basis prior on the per-position effect of each variant
and performs sparse variable selection by the SuSiE iterative
Bayesian stepwise selection (IBSS) algorithm.

The model is

$$
{\bf Y} = {\bf X} {\bf B} + {\bf E},
$$

where ${\bf Y}$ is `n x T` (functional response), ${\bf X}$ is
`n x p` (genotypes), and ${\bf B}$ is `p x T` with most rows zero
(few causal variants). fSuSiE writes ${\bf B}$ as a sum of
single-effect matrices ${\bf B} = \sum_{l=1}^L {\bf B}^{(l)}$
where each ${\bf B}^{(l)}$ has at most one non-zero row, then
fits the model in the wavelet domain.

`fsusie()` is the single-outcome wrapper around `mfsusie()`. The
argument order is `(Y, X, pos, ...)`.

## Example data

```{r setup, message = FALSE}
library(mfsusieR)
data(coverage_example)
str(coverage_example, max.level = 1)
coverage_example$causal_snps
```

The data are simulated cis-QTL coverage on a power-of-two grid
(`n = 574`, `p = 100`, `T = 128`). The genotype matrix is the
real LD scaffold from `susieR::N3finemapping$X[, 1:100]`. Two
causal SNPs at positions 25 and 75 carry smooth random
per-position effects sampled from the IBSS wavelet prior, the
same prior class fSuSiE assumes. The example is assay-agnostic:
the per-position response could be RNA-seq exon-body coverage,
ATAC-seq peak coverage, WGBS / ChIP-seq read counts on a window,
or any per-position coverage / read-count assay.

```{r causal-shapes, fig.height = 3.5}
plot(coverage_example$causal_betas[1L, ], type = "l",
     col = "#1f78b4", lwd = 2,
     ylim = range(coverage_example$causal_betas),
     xlab = "position", ylab = "effect")
abline(h = 0, lty = 3, col = "grey50")
lines(coverage_example$causal_betas[2L, ], col = "#33a02c", lwd = 2)
legend("topright",
       legend = paste0("SNP ", coverage_example$causal_snps),
       col = c("#1f78b4", "#33a02c"), lwd = 2, bty = "n")
```

## Fitting fSuSiE

```{r fit, message = FALSE}
fit <- fsusie(coverage_example$Y, coverage_example$X,
              pos = coverage_example$pos,
              L = 15, L_greedy = 5, verbose = TRUE)
fit$niter; fit$converged
fit$sets$cs
fit$pip[coverage_example$causal_snps]
```

Both causal SNPs receive PIP near 1 and land in two separate
credible sets.

```{r cs-correlations}
cs1 <- fit$sets$cs[[1]]
cs2 <- fit$sets$cs[[2]]
range(cor(coverage_example$X[, cs1])[upper.tri(diag(length(cs1)))])
```

## Visualizing the fit

`mfsusie_plot()` returns a two-panel layout for single-outcome
fits: PIPs on top, per-CS effect curves on the bottom. With
`facet_cs = "auto"` (the default) two CSes whose affected
regions are disjoint are stacked into separate rows.

```{r plot-fit}
mfsusie_plot(fit)
```

The raw inverse-DWT effect curves track the true shapes but
carry high-frequency content at the smaller wavelet scales.
After post-smoothing with the default method (`"TI"`), the
curves are denoised and pointwise 95% credible bands are
available.

```{r plot-smoothed}
fit_s <- mf_post_smooth(fit)
mfsusie_plot(fit_s)
```

`mf_post_smooth()` exposes other smoothers
(`"scalewise"`, `"HMM"`, `"smash"`) and their tunable
parameters; see `vignette("post_processing")` for the
comparison.

The errorbar style draws a per-position dot at the posterior
mean and a vertical bar to the credible band.

```{r plot-errorbar}
mfsusie_plot(fit_s, effect_style = "errorbar", facet_cs = "stack")
```

## Recovering the true effect

The smoothed posterior tracks the simulated per-position effects.
Overlay the credible band against the truth to verify recovery
on each causal SNP:

```{r truth-overlay, fig.height = 5, fig.width = 8}
truth <- coverage_example$causal_betas
band  <- fit_s$smoothed$TI$credible_bands

# Match each fitted CS to its causal SNP by lead PIP.
cs_lead <- vapply(fit$sets$cs, function(idx) {
  idx[which.max(fit$pip[idx])]
}, integer(1L))

par(mfrow = c(1L, length(cs_lead)),
    mar = c(4, 4, 2.5, 1))
for (i in seq_along(cs_lead)) {
  truth_idx <- match(cs_lead[i], coverage_example$causal_snps)
  pos_grid  <- coverage_example$pos
  b_i  <- band[[1L]][[i]]
  yrng <- range(b_i, truth, na.rm = TRUE)
  plot(pos_grid, truth[truth_idx, ], type = "l",
       col = "grey60", lwd = 1.5, lty = 2,
       xlab = "position", ylab = "effect",
       ylim = yrng,
       main = sprintf("CS %d (SNP %d)", i, cs_lead[i]))
  polygon(c(pos_grid, rev(pos_grid)),
          c(b_i[, 1L], rev(b_i[, 2L])),
          col = adjustcolor("#1f78b4", alpha.f = 0.25),
          border = NA)
  lines(pos_grid, fit_s$smoothed$TI$effect_curves[[1L]][[i]],
        col = "#1f78b4", lwd = 2)
  abline(h = 0, lty = 3, col = "grey50")
  legend("topright", legend = c("truth", "TI smooth", "95% band"),
         col = c("grey60", "#1f78b4", adjustcolor("#1f78b4", alpha.f = 0.5)),
         lty = c(2, 1, NA), lwd = c(1.5, 2, NA),
         pch = c(NA, NA, 15), bty = "n", cex = 0.75)
}
```

Both effect curves stay inside the band on the bulk of the
support, with the truth crossing the band only at the small
high-frequency wiggles the smoother trims.

## Uneven sampling positions

Wavelets assume a `2^J` evenly-spaced grid. `fsusie()` accepts
unevenly-spaced positions by remapping the data onto a regular
dyadic grid with the Kovac-Silverman procedure. Pass the
positions through the `pos` argument; the response is
interpolated internally and the fit returns the per-position
effect on the remapped grid.

To illustrate, drop a random subset of positions from the
evenly-spaced response and refit on the irregular grid:

```{r uneven-fit, message = FALSE}
set.seed(2)
T_full   <- ncol(coverage_example$Y)
keep     <- sort(sample.int(T_full, size = 100L, replace = FALSE))
Y_uneven <- coverage_example$Y[, keep]
fit_uneven <- fsusie(Y_uneven, coverage_example$X,
                     pos = keep,
                     L = 15, L_greedy = 5, verbose = TRUE)
fit_uneven$sets$cs
fit_uneven$pip[coverage_example$causal_snps]
```

```{r plot-uneven}
fit_uneven_s <- mf_post_smooth(fit_uneven)
mfsusie_plot(fit_uneven_s)
```

The credible bands are wider than in the evenly-sampled fit
because each position now contributes information at a
non-uniform local density.

## Prior choices

The wavelet-domain mixture prior on ${\bf B}^{(l)}$ has two
options for sharing variance components. The default,
`prior_variance_scope = "per_scale"`, fits a separate mixture
weight vector at each wavelet scale per outcome; this is the
standard fSuSiE prior. The alternative,
`prior_variance_scope = "per_outcome"`, collapses across scales
and fits a single mixture weight vector per outcome. The
per-scale prior has higher power and estimation accuracy on
typical functional data; the per-outcome prior is faster on
larger problems.

```{r prior-comparison, message = FALSE}
fit_per_scale <- fsusie(coverage_example$Y, coverage_example$X,
                        pos = coverage_example$pos,
                        L = 15, L_greedy = 5,
                        prior_variance_scope = "per_scale",
                        verbose = FALSE)
fit_per_out   <- fsusie(coverage_example$Y, coverage_example$X,
                        pos = coverage_example$pos,
                        L = 15, L_greedy = 5,
                        prior_variance_scope = "per_outcome",
                        verbose = FALSE)
fit_per_scale$niter
fit_per_out$niter
```

```{r prior-scatter, fig.height = 4, fig.width = 4}
plot(fit_per_scale$pip, fit_per_out$pip,
     pch = 16, cex = 0.8, col = "#1f78b4",
     xlab = "PIP (per-scale prior)",
     ylab = "PIP (per-outcome prior)",
     xlim = c(0, 1), ylim = c(0, 1),
     main = "Prior comparison: PIPs")
abline(0, 1, lty = 3, col = "grey50")
```

The PIPs are nearly identical for the variants in either
credible set. The per-scale prior is the recommended default
for functional fine-mapping problems.

## Wavelet-domain preprocessing

`fsusie()` exposes two preprocessing flags for the wavelet
matrix:

- `low_count_filter` (default `0`): wavelet columns with
  `median(|column|) <= low_count_filter` are masked, contributing
  `(Bhat = 0, Shat = 1)` to the SER step. Use a small positive
  value when the response has sparse non-zero entries (e.g.,
  ATAC-seq read pileups).
- `quantile_norm` (default `FALSE`): applies a column-wise
  rank-based normal quantile transform to the wavelet matrix.
  Use when the per-position residuals are heavy-tailed or
  visibly non-Gaussian.

See `vignette("practical_data_applications")` for an applied
walk-through on a multi-region ATAC-seq example dataset.

## What the fit object carries

```{r fields}
str(fit, max.level = 1)
```

The fit is a list of class `c("mfsusie", "susie")` with the
standard SuSiE slots (`alpha`, `mu`, `mu2`, `lbf`,
`lbf_variable`, `KL`, `pip`, `sets`) plus per-outcome wavelet
metadata (`dwt_meta`) used by `predict.mfsusie`,
`coef.mfsusie`, and `fitted.mfsusie`. After
`mf_post_smooth(fit, method = ...)`, the chosen smoother's
output lives at `fit$smoothed[[method]]` (with `effect_curves`,
`credible_bands`, and an optional `lfsr_curves` slot for HMM).
The fit accumulates per-method results across calls, so
multiple smoothers can coexist on the same fit; see
`vignette("post_processing")` for the comparison workflow.
`mfsusie_plot()` picks a smoother by priority `TI > smash >
HMM > scalewise` when more than one is present.
