---
title: "Adjusting for covariates in functional fine-mapping"
author: "William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Adjusting for covariates in functional fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

This vignette adjusts a functional response `Y` for covariate
effects before fine-mapping. Observed covariates that affect the
trait at every position (age, sex, batch, principal components
of a global expression matrix) inflate the residual variance and
attenuate the genetic signal if left in. The right tool for
removing such effects on a functional response is a wavelet-
domain empirical-Bayes regression that accommodates smooth
position-dependent covariate effects. `mfsusieR` exposes this
through `mf_adjust_for_covariates()`.

## Setup

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
library(wavethresh)
set.seed(2L)
```

## Generating the data

We simulate functional responses with two ingredients:

1. **Genetic signal** — two causal SNPs (positions 25 and 75 in
   the genotype matrix) with smooth random per-position effects
   `f1`, `f2`.
2. **Covariate signal** — three covariates (independent of
   genotype) acting through smooth random per-position
   coefficient curves `f1_cov`, `f2_cov`, `f3_cov`.

All effect curves are drawn from
`mf_simu_ibss_per_level(7)$sim_func`, which samples from the
IBSS wavelet prior (the same prior class fSuSiE infers).

```{r sim-effects, fig.height = 3.5}
rsnr    <- 1
pos1    <- 25L
pos2    <- 75L
lev_res <- 7L
T_m     <- 2L^lev_res

f1     <- mf_simu_ibss_per_level(lev_res)$sim_func
f2     <- mf_simu_ibss_per_level(lev_res)$sim_func
f1_cov <- mf_simu_ibss_per_level(lev_res)$sim_func
f2_cov <- mf_simu_ibss_per_level(lev_res)$sim_func
f3_cov <- mf_simu_ibss_per_level(lev_res)$sim_func

plot(f1, type = "l", col = "royalblue", lwd = 2,
     xlab = "position", ylab = "effect")
abline(h = 0, lty = 3, col = "grey50")
lines(f2, type = "l", col = "limegreen", lwd = 2)
legend("topright", legend = c("f1 (SNP 25)", "f2 (SNP 75)"),
       col = c("royalblue", "limegreen"), lwd = 2, bty = "n")
```

The observed data is a mixture of covariate-driven confounding
and genotype signal. `genetic_signal` is the part we want to
recover; `covariate_signal` is the part we want to remove.

```{r sim-data}
data(N3finemapping)
Geno <- N3finemapping$X[, seq_len(100)]
n    <- nrow(Geno)

# Three covariates, sd = 2; independent of Geno.
Cov <- matrix(rnorm(3L * n, sd = 2), ncol = 3L)

genetic_signal   <- matrix(0, n, T_m)
covariate_signal <- matrix(0, n, T_m)
for (i in seq_len(n)) {
  noise_i <- rnorm(T_m, sd = (1 / rsnr) * var(f1))
  genetic_signal[i, ] <-
    Geno[i, pos1] * f1 + Geno[i, pos2] * f2 + noise_i
  covariate_signal[i, ] <-
    Cov[i, 1L] * f1_cov + Cov[i, 2L] * f2_cov + Cov[i, 3L] * f3_cov
}
Y <- covariate_signal + genetic_signal
```

The covariate signal dominates the per-position scale of `Y`,
so without adjustment fSuSiE's recovered effect curves are
biased toward the covariate-induced confounding rather than the
true genetic effect.

## Account for covariates

`mf_adjust_for_covariates(Y, Cov, method = "wavelet_eb")` fits
a wavelet-domain empirical-Bayes regression of `Y` on the
covariates `Cov` and returns the residualised response
`Y_adjusted = Y - Cov %*% fitted_func` along with the fitted
covariate effect curves.

```{r adjust, message = FALSE}
adj <- mf_adjust_for_covariates(Y, Cov)
str(adj, max.level = 1L)
```

The first row of `adj$fitted_func` is the recovered effect of
the first covariate on the response curve; overlay it on the
truth `f1_cov`:

```{r plot-fitted-cov, fig.height = 3.5}
plot(adj$fitted_func[1L, ], type = "l", col = "blue",
     lty = 2, lwd = 2,
     xlab = "position", ylab = "covariate effect",
     main = "Covariate 1: estimated vs true")
lines(f1_cov, lwd = 1.5)
legend("topright", lwd = c(1.5, 2), lty = c(1, 2),
       legend = c("true f1_cov", "estimated"), bty = "n")
```

`adj$Y_adjusted` is the residualised response on the original
position grid, ready to feed into `fsusie()`.

## Recovering the genetic signal

A two-panel scatter shows that the adjusted response tracks the
genetic-only signal far more tightly than the raw response:

```{r recovery-scatter, fig.height = 4, fig.width = 8}
par(mfrow = c(1L, 2L))
plot(Y, genetic_signal, pch = 19L, cex = 0.4,
     xlab = "Observed Y", ylab = "Genetic signal",
     main = "Before adjustment")
abline(0, 1, lty = 3, col = "grey50")
plot(adj$Y_adjusted, genetic_signal, pch = 19L, cex = 0.4,
     xlab = "Y adjusted", ylab = "Genetic signal",
     main = "After adjustment")
abline(0, 1, lty = 3, col = "grey50")
par(mfrow = c(1L, 1L))
```

## Fine-mapping the adjusted response

```{r fit-adj, message = FALSE}
fit_adj <- fsusie(adj$Y_adjusted, Geno)
fit_adj$sets$cs
fit_adj$pip[c(pos1, pos2)]
```

```{r plot-fit-adj-dims}
plot_dims <- mfsusie_plot_dimensions(fit_adj)
```

```{r plot-fit-adj, fig.width = plot_dims$width, fig.height = plot_dims$height}
mfsusie_plot(fit_adj, effect_variables = c(pos1, pos2))
```

## Comparison with no adjustment

For contrast, fit `fsusie()` directly on `Y` without
adjustment:

```{r fit-unadj, message = FALSE}
fit_unadj <- fsusie(Y, Geno)
fit_unadj$sets$cs
fit_unadj$pip[c(pos1, pos2)]
```

`fsusie()` returns raw inverse-DWT effect curves via `coef()`;
position-space smoothing with a 95% credible band is an opt-in
post-processing step. Run `mf_post_smooth(fit, method = "TI")`
on each fit to get translation-invariant cycle-spun curves,
then read them back via `coef(fit_s, smooth_method = "TI")`.
Match each fitted CS to its causal SNP by lead-PIP and overlay
the smoothed recovered curve from each fit against the
ground-truth `f1` and `f2`.

```{r post-smooth, message = FALSE}
fit_adj_s   <- mf_post_smooth(fit_adj,   method = "TI")
fit_unadj_s <- mf_post_smooth(fit_unadj, method = "TI")
cf_adj   <- coef(fit_adj_s,   smooth_method = "TI")[[1L]]
cf_unadj <- coef(fit_unadj_s, smooth_method = "TI")[[1L]]

match_cs <- function(fit, target_snp) {
  if (length(fit$sets$cs) == 0L) return(NA_integer_)
  leads <- vapply(fit$sets$cs, function(idx) {
    idx[which.max(fit$pip[idx])]
  }, integer(1L))
  hit <- which(leads == target_snp)
  if (length(hit) == 0L) NA_integer_ else hit[1L]
}

plot_effect <- function(snp, truth, col, label) {
  l_adj   <- match_cs(fit_adj,   snp)
  l_unadj <- match_cs(fit_unadj, snp)
  rec_adj   <- if (!is.na(l_adj))   cf_adj[l_adj, ]     else NULL
  rec_unadj <- if (!is.na(l_unadj)) cf_unadj[l_unadj, ] else NULL
  yrng <- range(c(truth, rec_adj, rec_unadj), na.rm = TRUE)
  plot(truth, pch = 20L, col = "grey30", cex = 0.7,
       xlab = "position", ylab = "effect", ylim = yrng,
       main = paste("Estimated effect:", label))
  abline(h = 0, lty = 3, col = "grey50")
  if (!is.null(rec_adj))   lines(rec_adj,   col = col, lwd = 1.8)
  if (!is.null(rec_unadj)) lines(rec_unadj, col = col, lwd = 1.5, lty = 2)
  legend("topright",
         legend = c("truth", "fSuSiE adjusted", "fSuSiE unadjusted"),
         col    = c("grey30", col, col),
         lty    = c(NA, 1, 2), lwd = c(NA, 1.8, 1.5),
         pch    = c(20, NA, NA), bty = "n", cex = 0.85)
}
```

Estimated effect 1 (true causal at SNP 25):

```{r effect-1, fig.height = 4, fig.width = 8}
plot_effect(pos1, f1, "blue", "f1 (SNP 25)")
```

Estimated effect 2 (true causal at SNP 75):

```{r effect-2, fig.height = 4, fig.width = 8}
plot_effect(pos2, f2, "darkgreen", "f2 (SNP 75)")
```

The TI-smoothed adjusted-fit curves track the truth on both
signals at near-perfect correlation. The unadjusted fit is
biased toward the covariate-induced confounding and the
recovered curves drift away from the truth.

## Scalar traits: OLS adjustment with the FWL correction

When the response is a scalar (`T = 1`) or smoothness across
positions is irrelevant, the closed-form OLS residualization
is the right tool. Pass `method = "ols"` to use it. When the
covariates are correlated with genotype, also pass `X` so the
function returns the Frisch-Waugh-Lovell-corrected
`X_adjusted`:

```{r ols-scalar, message = FALSE}
# Scalar response derived from the same covariate-confounded
# data: take the per-individual mean of Y as the scalar trait
# and wrap it as a single-column matrix.
Y_scalar <- matrix(rowMeans(Y), ncol = 1L)

adj_ols <- mf_adjust_for_covariates(Y_scalar, Cov, X = Geno,
                                    method = "ols")
str(adj_ols, max.level = 1L)
```

`adj_ols` returns:

- `Y_adjusted`: the scalar response after projecting out the
  covariate column space.
- `X_adjusted`: the genotype matrix after the same projection
  (the FWL correction). When `Cov` is correlated with `Geno`,
  using `X_adjusted` instead of the original `Geno` removes
  the bias in the downstream fit.

A scalar `susie()` fit on the OLS-adjusted data:

```{r ols-fit, message = FALSE}
fit_scalar <- susieR::susie(adj_ols$X_adjusted,
                            as.numeric(adj_ols$Y_adjusted),
                            L = 5L, verbose = FALSE)
fit_scalar$sets$cs
fit_scalar$pip[c(pos1, pos2)]
```

For `T > 1` use `method = "wavelet_eb"` (the default
demonstrated above). Use `method = "ols"` for scalar traits or
when smooth covariate effects across positions are not
expected.

## Next steps

For colocalization across two functional fits, see
[the colocalization vignette](fsusie_colocalization.html). For
post-fit smoothing of effect curves, see
[the post-processing vignette](post_processing.html).

## Session info

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
