---
title: "DNA methylation QTL fine-mapping"
author: "Gao Wang, William Denault, Peter Carbonetto"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{DNA methylation QTL fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", results = "hold",
  fig.align = "center", dpi = 120
)
```

This vignette walks through a DNA methylation QTL analysis using
`fsusie()`. The methylation phenotype is a vector of CpG-level beta
values along a region; `fsusie()` treats it as a single functional
response sampled at the CpG positions, returns credible sets of variables
that drive coordinated changes across CpGs, and provides the per-CpG
effect curves for each credible set.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Simulated DNAm region

Replace this block with your own genotype matrix and CpG-level
methylation matrix in practice.

```{r simulate}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X)

# Five CpGs in the region, irregularly spaced.
cpg_pos <- c(120, 320, 480, 540, 1200)
T_m <- length(cpg_pos)

# Two causal variables with localised effects on subsets of CpGs.
beta <- matrix(0, nrow = p, ncol = T_m)
beta[42, ] <-  c(0.20, 0.18, 0.05, 0.02, 0.00)   # 5'-side cluster
beta[97, ] <-  c(0.00, 0.00, 0.03, 0.10, 0.12)   # 3'-side cluster

methyl_logit <- X %*% beta + matrix(rnorm(n * T_m, sd = 0.05), n)
# Beta values are bounded in [0, 1]; logit on a baseline of 0.5.
Y <- 1 / (1 + exp(-(methyl_logit + qlogis(0.5))))
Y_df <- as.data.frame(cbind(Y, X[, c(42, 97)]))
colnames(Y_df) <- c(paste0("CpG_", seq_len(T_m)), "snp_42", "snp_97")
head(Y_df, 3)
```

The CpG positions are irregular; `fsusie()` (via `mfsusie()`) remaps
the response onto a power-of-two grid before the wavelet step (see
`?mfsusieR:::remap_data`).

## Fit

```{r fit}
fit <- fsusie(Y, X, pos = cpg_pos, L = 3, verbose = FALSE)
fit$niter; fit$converged
fit$sets$cs
```

## Inspecting the fit

`mfsusie_plot()` shows PIPs (top, CS-colored) and per-effect
methylation curves on the post-remap grid. After `mf_post_smooth()`
the curves are denoised and credible bands are drawn as ribbons.

```{r plot-smoothed, fig.width = 6, fig.height = 5, fig.cap = "TI-smoothed methylation curves with 95% pointwise credible bands."}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

`coef(fit)` returns a length-`M` list; element `m` is an `L x T_basis`
matrix on the wavelet-remapped grid (the irregular CpG positions are
interpolated to a regular dyadic grid before the DWT).

```{r effect-curves}
B_hat <- coef(fit)[[1]]
sapply(coef(fit), dim)
```

## Predictions on new genotypes

```{r predict}
fitted_curves <- predict(fit, newx = X[1:3, ])[[1]]
pos_remap <- fit$dwt_meta$pos[[1]]
matplot(pos_remap, t(fitted_curves), type = "l", lty = 1,
        col = c("#E69F00", "#56B4E9", "#009E73"), lwd = 1.5,
        xlab = "position (post-remap grid)",
        ylab = "predicted methylation",
        main = "Predicted profiles for three individuals")
legend("topright", legend = paste("subject", 1:3),
       col = c("#E69F00", "#56B4E9", "#009E73"),
       lwd = 1.5, bty = "n")
```

## What this gets you

- Per-CpG resolution: each credible set comes with the full
  methylation profile of its effect, not just a scalar.
- Joint information: a variable that nudges multiple correlated CpGs is
  prioritized over a variable that nudges one in isolation.
- A drop-in API: this is `susie()` with `Y` replaced by a CpG matrix
  and a `pos` vector of positions.

For a multi-region or multi-omics version of the same workflow (e.g.,
methylation + RNA-seq jointly), see
[the multi-outcome intro](mfsusie_intro.html).
