---
title: "Functional fine-mapping with fSuSiE"
author: "William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Functional fine-mapping with fSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

## Overview

fSuSiE extends the SuSiE Bayesian variable selection model to
functional traits. The functional response is sampled at multiple
positions of a regular grid (e.g., methylation along a CpG
island, RNA-seq coverage along a gene body, ATAC-seq across a
peak). fSuSiE places a wavelet-basis prior on the per-position
effect of each variant and performs sparse variable selection by
the SuSiE iterative Bayesian stepwise selection (IBSS) algorithm.

The model is

$$
{\bf Y} = {\bf X} {\bf B} + {\bf E},
$$

where ${\bf Y}$ is `n x T` (functional response), ${\bf X}$ is
`n x p` (genotypes), and ${\bf B}$ is `p x T` with most rows zero
(few causal variants). fSuSiE writes ${\bf B}$ as a sum of
single-effect matrices ${\bf B} = \sum_{l=1}^L {\bf B}^{(l)}$
where each ${\bf B}^{(l)}$ has at most one non-zero row, then
fits the model in the wavelet domain.

`fsusie()` is the single-outcome wrapper around `mfsusie()`. The
argument order is `(Y, X, pos, ...)`.

## Example data

```{r setup, message = FALSE}
library(mfsusieR)
data(rnaseq_example)
str(rnaseq_example, max.level = 1)
rnaseq_example$causal_snps
```

The data are simulated cis-eQTL coverage on a power-of-two grid
(`n = 574`, `p = 100`, `T = 128`). Two causal SNPs at positions
25 and 75 carry smooth random per-position effects generated by
wavelet-coefficient shrinkage. The simulation script is
`data-raw/make_data.R`.

```{r causal-shapes, fig.height = 3.5}
plot(rnaseq_example$causal_betas[1L, ], type = "l",
     col = "#1f78b4", lwd = 2,
     ylim = range(rnaseq_example$causal_betas),
     xlab = "position", ylab = "effect")
abline(h = 0, lty = 3, col = "grey50")
lines(rnaseq_example$causal_betas[2L, ], col = "#33a02c", lwd = 2)
legend("topright",
       legend = paste0("SNP ", rnaseq_example$causal_snps),
       col = c("#1f78b4", "#33a02c"), lwd = 2, bty = "n")
```

## Fitting fSuSiE

```{r fit, message = FALSE}
fit <- fsusie(rnaseq_example$Y, rnaseq_example$X,
              pos = rnaseq_example$pos,
              L = 15, L_greedy = 5, verbose = TRUE)
fit$niter; fit$converged
fit$sets$cs
fit$pip[rnaseq_example$causal_snps]
```

Both causal SNPs receive PIP near 1 and land in two separate
credible sets.

```{r cs-correlations}
cs1 <- fit$sets$cs[[1]]
cs2 <- fit$sets$cs[[2]]
range(cor(rnaseq_example$X[, cs1])[upper.tri(diag(length(cs1)))])
```

## Visualizing the fit

`mfsusie_plot()` returns a two-panel layout for single-outcome
fits: PIPs on top, per-CS effect curves on the bottom. With
`facet_cs = "auto"` (the default) two CSes whose affected
regions are disjoint are stacked into separate rows.

```{r plot-fit}
mfsusie_plot(fit)
```

The raw inverse-DWT effect curves track the true shapes but
carry high-frequency content at the smaller wavelet scales.
After post-smoothing, the curves are denoised and pointwise 95%
credible bands are available.

```{r plot-smoothed}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

The errorbar style draws a per-position dot at the posterior
mean and a vertical bar to the credible band.

```{r plot-errorbar}
mfsusie_plot(fit_s, effect_style = "errorbar", facet_cs = "stack")
```

## Uneven sampling positions

Wavelets assume a `2^J` evenly-spaced grid. `fsusie()` accepts
unevenly-spaced positions by remapping the data onto a regular
dyadic grid with the Kovac-Silverman procedure. Pass the
positions through the `pos` argument; the response is
interpolated internally and the fit returns the per-position
effect on the remapped grid.

To illustrate, drop a random subset of positions from the
evenly-spaced response and refit on the irregular grid:

```{r uneven-fit, message = FALSE}
set.seed(2)
T_full   <- ncol(rnaseq_example$Y)
keep     <- sort(sample.int(T_full, size = 100L, replace = FALSE))
Y_uneven <- rnaseq_example$Y[, keep]
fit_uneven <- fsusie(Y_uneven, rnaseq_example$X,
                     pos = keep,
                     L = 15, L_greedy = 5, verbose = TRUE)
fit_uneven$sets$cs
fit_uneven$pip[rnaseq_example$causal_snps]
```

```{r plot-uneven}
fit_uneven_s <- mf_post_smooth(fit_uneven, method = "TI",
                               wavelet_filter = 1L,
                               wavelet_family = "DaubExPhase")
mfsusie_plot(fit_uneven_s)
```

The credible bands are wider than in the evenly-sampled fit
because each position now contributes information at a
non-uniform local density.

## Prior choices

The wavelet-domain mixture prior on ${\bf B}^{(l)}$ has two
options for sharing variance components. The default,
`prior_variance_scope = "per_scale"`, fits a separate mixture
weight vector at each wavelet scale per outcome; this is the
standard fSuSiE prior. The alternative,
`prior_variance_scope = "per_outcome"`, collapses across scales
and fits a single mixture weight vector per outcome. The
per-scale prior has higher power and estimation accuracy on
typical functional data; the per-outcome prior is faster on
larger problems.

```{r prior-comparison, message = FALSE}
fit_per_scale <- fsusie(rnaseq_example$Y, rnaseq_example$X,
                        pos = rnaseq_example$pos,
                        L = 15, L_greedy = 5,
                        prior_variance_scope = "per_scale",
                        verbose = FALSE)
fit_per_out   <- fsusie(rnaseq_example$Y, rnaseq_example$X,
                        pos = rnaseq_example$pos,
                        L = 15, L_greedy = 5,
                        prior_variance_scope = "per_outcome",
                        verbose = FALSE)
fit_per_scale$niter
fit_per_out$niter
```

```{r prior-scatter, fig.height = 4, fig.width = 4}
plot(fit_per_scale$pip, fit_per_out$pip,
     pch = 16, cex = 0.8, col = "#1f78b4",
     xlab = "PIP (per-scale prior)",
     ylab = "PIP (per-outcome prior)",
     xlim = c(0, 1), ylim = c(0, 1),
     main = "Prior comparison: PIPs")
abline(0, 1, lty = 3, col = "grey50")
```

The PIPs are nearly identical for the variants in either
credible set. The per-scale prior is the recommended default
for functional fine-mapping problems.

## What the fit object carries

```{r fields}
str(fit, max.level = 1)
```

The fit is a list of class `c("mfsusie", "susie")` with the
standard SuSiE slots (`alpha`, `mu`, `mu2`, `lbf`,
`lbf_variable`, `KL`, `pip`, `sets`) plus per-outcome wavelet
metadata (`dwt_meta`) used by `predict.mfsusie`,
`coef.mfsusie`, and `fitted.mfsusie`. After `mf_post_smooth()`,
`effect_curves` and `credible_bands` are populated and
`mfsusie_plot()` reads them.
