---
title: "GTEx-style RNA-seq cis-eQTL fine-mapping"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{GTEx-style RNA-seq cis-eQTL fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

This vignette walks through a GTEx-style cis-eQTL fine-mapping
analysis using `fsusie()`. The phenotype is RNA-seq read coverage
along a gene body; `fsusie()` treats it as a single functional
response and returns credible sets of variants that drive
coordinated changes in expression at distinct positions of the
gene.

The published case studies (AHCYL1, SCD, HSP90AA1, STMN2,
ABHD17A) all share this shape: a multi-CS fit on log1p-transformed
RNA-seq counts over a cis window. **The underlying GTEx
individual-level data is protected and cannot be redistributed
with this package**, so we simulate data (`gtex_example`) that
reproduces the multi-CS structure of those figures and walk
through both the standard plot and a publication-grade
genome-browser overlay on this simulated data.

## Example dataset

```{r setup, message = FALSE}
library(mfsusieR)
data(gtex_example)
str(gtex_example, max.level = 1)
```

200 SNPs over a real LD scaffold (`susieR::N3finemapping$X`),
256 gene-body positions in a 32 kb window on chr1
(109.99 Mb - 110.02 Mb), three causal SNPs at distinct positions
(5', mid-body, 3') with localized peak effects.

## Fit

```{r fit}
fit <- fsusie(gtex_example$Y, gtex_example$X,
              pos = gtex_example$pos,
              L = 15, L_greedy = 5,
              verbose = TRUE)
fit$niter; fit$converged
length(fit$sets$cs)
fit$pip[gtex_example$causal_snps]
```

All three causal SNPs receive PIP = 1 and land in three separate
credible sets.

## Standard fsusie figure

`mfsusie_plot()` shows PIPs colored by CS membership and the
TI-smoothed per-CS effect curves with 95% credible bands plus
affected-region bars at the bottom of the effect panel.

```{r plot-smoothed}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

The three CS effect curves capture three separate gene-body
regions, mirroring the multi-CS structure in published GTEx
fSuSiE figures.

## Genome-browser overlay (Gviz)

For a publication-grade figure with chromosome ideogram, genome
axis, fitted effect track, observed allele-stratified coverage,
and read-count tracks, `mfsusieR` ships a worked example using
`Gviz`. The Bioconductor packages it depends on (`Gviz`,
`GenomicRanges`, `IRanges`) are listed as `Suggests` only.

```{r gviz, eval = requireNamespace("Gviz", quietly = TRUE), fig.height = 7}
suppressPackageStartupMessages({
  library(Gviz); library(GenomicRanges); library(IRanges)
})

cs <- 1L                         # which credible set to feature
markers <- fit_s$sets$cs[[cs]]
lead    <- markers[which.max(fit_s$pip[markers])]
x_lead  <- gtex_example$X[, lead]

# Effect curve + 95% band on the gene-body grid.
band  <- fit_s$credible_bands[[1L]][[cs]]
effect <- fit_s$effect_curves[[1L]][[cs]]

# Allele-stratified mean log1p coverage at the lead variant.
# The genotype matrix is mean-centered, so a 0/1 split on the
# sign of the dosage approximates a reference / carrier split.
g0 <- which(x_lead < 0)   # reference (below-mean dosage)
g1 <- which(x_lead >= 0)  # carrier (above-mean dosage)
mu0 <- colMeans(gtex_example$Y[g0, , drop = FALSE])
mu1 <- colMeans(gtex_example$Y[g1, , drop = FALSE])

chrom <- gtex_example$chrom
pos   <- gtex_example$pos

gr     <- GRanges(chrom, IRanges(pos, pos + 1L))
ylim_e <- range(c(effect, band)) * c(1.1, 1.1)

# Effect track: solid mean line with two dashed credible-band
# lines, layered via OverlayTrack so all three render on the
# same panel.
e_main <- DataTrack(range = gr, data = effect, type = "l", lwd = 2,
                    col = "#1f78b4", name = sprintf("Effect CS %d", cs),
                    ylim = ylim_e,
                    background.title = "white", col.axis = "black",
                    col.title = "black", fontface.title = 1)
e_lo   <- DataTrack(range = gr, data = band[, 1L], type = "l",
                    lty = 2, col = "#1f78b4", ylim = ylim_e)
e_hi   <- DataTrack(range = gr, data = band[, 2L], type = "l",
                    lty = 2, col = "#1f78b4", ylim = ylim_e)
effect_track <- OverlayTrack(trackList = list(e_main, e_lo, e_hi),
                             background.title = "white")

obs_diff_track <- DataTrack(range = gr, data = mu1 - mu0,
                            type = "p", pch = 16, cex = 0.5,
                            col = "#1f78b4",
                            name = "Observed difference",
                            background.title = "white",
                            col.axis = "black", col.title = "black",
                            fontface.title = 1)

ylim_c <- range(c(mu0, mu1))
c_ref  <- DataTrack(range = gr, data = mu0, type = "p", pch = 16,
                    cex = 0.5, col = "navy",
                    name = "Avg. log1p count", ylim = ylim_c,
                    background.title = "white", col.axis = "black",
                    col.title = "black", fontface.title = 1)
c_alt  <- DataTrack(range = gr, data = mu1, type = "p", pch = 16,
                    cex = 0.5, col = "turquoise", ylim = ylim_c)
count_track <- OverlayTrack(trackList = list(c_ref, c_alt),
                            background.title = "white")

axis_track <- GenomeAxisTrack(col = "black", fontcolor = "black",
                              col.title = "black",
                              background.title = "white")

# Gene-region track (simulated exon / intron structure shipped
# with `gtex_example`): two isoforms of the AHCYL1-like gene plus
# a 5' L1 element, drawn as exon boxes connected by intron lines.
gene_track <- GeneRegionTrack(gtex_example$gene_track,
                              chromosome = chrom,
                              name = "", showId = TRUE,
                              geneSymbol = TRUE,
                              col = "salmon", fill = "salmon",
                              background.title = "white",
                              col.axis = "black", col.title = "black")

plotTracks(list(axis_track, effect_track, obs_diff_track,
                count_track, gene_track),
           from = gtex_example$locus[1L],
           to   = gtex_example$locus[2L],
           sizes = c(1, 3, 3, 3, 2),
           main  = sprintf("%s cis-eQTL: CS %d (lead SNP rank %d, %d ref / %d carrier)",
                           gtex_example$gene, cs, lead,
                           length(g0), length(g1)),
           cex.main = 1.0)
```

The Gviz figure layout consists of
top axis with positions in the gene-body window, the fSuSiE
effect curve with dashed credible band, the observed
allele-stratified coverage difference at each gene-body position,
and the per-position log1p counts split by genotype class.
Adding an `IdeogramTrack` from a real chromosome and a
`GeneRegionTrack` from a `TxDb` package (e.g.
`TxDb.Hsapiens.UCSC.hg38.knownGene`) extends this to a full
ideogram + transcript layout; the worked-example scripts in
`system.file("scripts/gtex_case_study", package = "mfsusieR")`
show the wiring on a real `(chrom, gene)` pair when the
underlying RNA-seq data is available locally.

## Why functional fine-mapping helps here

The fine-mapping signal is spread across many positions of the
gene. A scalar SuSiE fit on `rowSums(Y)` (total expression)
mixes every per-position effect into one number and loses the
per-region structure. The functional model preserves it:

```{r susie-comparison}
library(susieR)
fit_scalar <- susie(gtex_example$X, rowSums(gtex_example$Y),
                    L = 15, verbose = FALSE)
fit_scalar$sets$cs
fit_scalar$pip[gtex_example$causal_snps]
```

Per-position effects are not recoverable from the scalar fit;
they are the natural output of `fsusie()`.
