---
title: "DNA methylation QTL fine-mapping"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{DNA methylation QTL fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", results = "hold",
  fig.align = "center", dpi = 120
)
```

This vignette walks through a DNA methylation QTL analysis using
`fsusie()`. The methylation phenotype is a vector of CpG-level beta
values along a region; `fsusie()` treats it as a single functional
response sampled at the CpG positions, returns credible sets of variables
that drive coordinated changes across CpGs, and provides the per-CpG
effect curves for each credible set.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Bundled DNAm fixture

The package bundles `dnam_example`, a simulated cis-mQTL fixture
shaped after a real DNA methylation profile: 200 SNPs over a real
LD scaffold (`susieR::N3finemapping$X` columns 101-300), 64
irregular CpG positions across a ~10-kb region, two causal SNPs
with overlapping but oppositely-signed Gaussian methylation
effects. Replace this with your own `(X, Y, pos)` in practice.

```{r load-data}
data(dnam_example)
str(dnam_example, max.level = 1)
dnam_example$causal_snps
```

The CpG positions are irregular; `fsusie()` remaps the response
onto a power-of-two grid before the wavelet step.

## Fit

```{r fit}
fit <- fsusie(dnam_example$Y, dnam_example$X,
              pos = dnam_example$pos, L = 15, L_greedy = 5, verbose = TRUE)
fit$niter; fit$converged
fit$sets$cs
fit$pip[dnam_example$causal_snps]
```

## Inspecting the fit

`mfsusie_plot()` shows PIPs (top, CS-colored) and per-effect
methylation curves on the post-remap grid. After `mf_post_smooth()`
the curves are denoised and credible bands are drawn as ribbons.

```{r plot-smoothed, fig.width = 6, fig.height = 5, fig.cap = "TI-smoothed methylation curves with 95% pointwise credible bands."}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

`coef(fit)` returns a length-`M` list; element `m` is an `L x T_basis`
matrix on the wavelet-remapped grid (the irregular CpG positions are
interpolated to a regular dyadic grid before the DWT).

```{r effect-curves}
B_hat <- coef(fit)[[1]]
sapply(coef(fit), dim)
```

## Predictions on new genotypes

```{r predict}
fitted_curves <- predict(fit, newx = dnam_example$X[1:3, ])[[1]]
pos_remap <- fit$dwt_meta$pos[[1]]
matplot(pos_remap, t(fitted_curves), type = "l", lty = 1,
        col = c("#E69F00", "#56B4E9", "#009E73"), lwd = 1.5,
        xlab = "position (post-remap grid)",
        ylab = "predicted methylation",
        main = "Predicted profiles for three individuals")
legend("topright", legend = paste("subject", 1:3),
       col = c("#E69F00", "#56B4E9", "#009E73"),
       lwd = 1.5, bty = "n")
```

## What this gets you

- Per-CpG resolution: each credible set comes with the full
  methylation profile of its effect, not just a scalar.
- Joint information: a variable that nudges multiple correlated CpGs is
  prioritized over a variable that nudges one in isolation.
- A drop-in API: this is `susie()` with `Y` replaced by a CpG matrix
  and a `pos` vector of positions.

For a multi-region or multi-omics version of the same workflow (e.g.,
methylation + RNA-seq jointly), see
[the multi-outcome intro](mfsusie_intro.html).
