---
title: "DNA methylation QTL fine-mapping with fsusie()"
author: "Gao Wang, William Denault, Peter Carbonetto"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{DNA methylation QTL fine-mapping with fsusie()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", results = "hold",
  fig.align = "center", dpi = 120
)
```

This vignette walks through a DNA methylation QTL analysis using
`fsusie()`. The methylation phenotype is a vector of CpG-level beta
values along a region; `fsusie()` treats it as a single functional
response sampled at the CpG positions, returns credible sets of SNPs
that drive coordinated changes across CpGs, and provides the per-CpG
effect curves for each credible set.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
library(ggplot2)
library(reshape2)
set.seed(1)
```

## Simulated DNAm region

Replace this block with your own genotype matrix and CpG-level
methylation matrix in practice.

```{r simulate}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X)

# Five CpGs in the region, irregularly spaced.
cpg_pos <- c(120, 320, 480, 540, 1200)
T_m <- length(cpg_pos)

# Two causal SNPs with localised effects on subsets of CpGs.
beta <- matrix(0, nrow = p, ncol = T_m)
beta[42, ] <-  c(0.20, 0.18, 0.05, 0.02, 0.00)   # 5'-side cluster
beta[97, ] <-  c(0.00, 0.00, 0.03, 0.10, 0.12)   # 3'-side cluster

methyl_logit <- X %*% beta + matrix(rnorm(n * T_m, sd = 0.05), n)
# Beta values are bounded in [0, 1]; logit on a baseline of 0.5.
Y <- 1 / (1 + exp(-(methyl_logit + qlogis(0.5))))
Y_df <- as.data.frame(cbind(Y, X[, c(42, 97)]))
colnames(Y_df) <- c(paste0("CpG_", seq_len(T_m)), "snp_42", "snp_97")
head(Y_df, 3)
```

The CpG positions are irregular; `fsusie()` (via `mfsusie()`) remaps
the response onto a power-of-two grid before the wavelet step (see
`?mfsusieR:::remap_data`).

## Fit

```{r fit}
fit <- fsusie(Y, X, pos = cpg_pos, L = 3, verbose = FALSE)
fit$niter; fit$converged
fit$sets$cs
```

PIPs:

```{r pip}
plot(fit$pip, pch = 16, col = ifelse(fit$pip > 0.9, "red", "grey60"),
     xlab = "SNP index", ylab = "PIP")
abline(v = c(42, 97), lty = 2, col = "blue")
```

## Per-effect methylation curves

`coef(fit)` returns a length-`M` list; element `m` is an `L x T_padded`
matrix on the wavelet-remapped grid (the irregular CpG positions are
interpolated to a regular dyadic grid before the DWT). Plot against
`fit$mf_meta$pos[[m]]` to land on the right x-axis.

```{r effect-curves}
B_hat <- coef(fit)[[1]]
pos_remap <- fit$mf_meta$pos[[1]]
T_pad <- ncol(B_hat)

eff_df <- as.data.frame(t(B_hat))
eff_df$pos <- pos_remap
eff_long <- melt(eff_df, id.vars = "pos",
                 variable.name = "effect", value.name = "estimate")
ggplot(eff_long,
       aes(x = pos, y = estimate, colour = effect, group = effect)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "position (post-remap grid)", y = "effect on methylation",
       title = "Per-effect methylation curves")
```

## Predictions on new genotypes

```{r predict}
fitted_curves <- predict(fit, newx = X[1:3, ])[[1]]
pos_remap <- fit$mf_meta$pos[[1]]
matplot(pos_remap, t(fitted_curves), type = "b", lty = 1, pch = 16,
        xlab = "position (post-remap grid)",
        ylab = "predicted methylation",
        main = "Predicted profiles for three individuals")
```

## What this gets you

- Per-CpG resolution: each credible set comes with the full
  methylation profile of its effect, not just a scalar.
- Joint information: a SNP that nudges multiple correlated CpGs is
  prioritised over a SNP that nudges one in isolation.
- A drop-in API: this is `susie()` with `Y` replaced by a CpG matrix
  and a `pos` vector of positions.

For a multi-region or multi-omics version of the same workflow (e.g.,
methylation + RNA-seq jointly), see
[the multi-modality intro](mfsusie_intro.html).
