---
title: "Why functional fine-mapping"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Why functional fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 5, fig.align = "center", dpi = 110,
  out.width = "100%"
)
has_mvsusie <- requireNamespace("mvsusieR", quietly = TRUE)
```

Three common alternatives for fine-mapping a functional response
all lose information relative to fSuSiE: per-position SuSiE,
top-PC SuSiE, and naive multivariate SuSiE. This vignette runs
the same simulation that ships with the broader SuSiE-suite
diagnostics and shows the four methods side by side. The
canonical setup is a single causal SNP whose effect spans 88
out of 128 positions; we report PIPs and recovered effect
curves for each method.

## Simulation

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
data(N3finemapping)
X_full <- N3finemapping$X[seq_len(100), ]
T_m    <- 128L

effect <- 1.2 * cos(seq_len(T_m) / T_m * 3 * pi)
effect[effect < 0] <- 0
effect[seq_len(40)] <- 0     # zero on positions 1..40

true_pos <- 700L
y <- matrix(0, 100L, T_m)
for (i in seq_len(100L)) {
  y[i, ] <- X_full[i, true_pos] * effect + rnorm(T_m)
}
plot(effect, pch = 19, xlab = "position", ylab = "true effect")
```

```{r filter-zero-var}
col_var <- apply(X_full, 2, var)
keep    <- which(col_var > 0)
X       <- X_full[, keep]
true_pos_filt <- which(keep == true_pos)
cat("N =", nrow(X), "  J =", ncol(X), "  T =", T_m,
    "  true SNP filtered index =", true_pos_filt, "\n")
```

## Method 1: Per-position SuSiE

For each affected position run `susie(X, y[, t], L = 1)` and
collect the credible sets. Two failure modes appear: (i)
SuSiE returns no credible set at most affected positions; (ii)
when it does return a CS, the lead variant is inconsistent
across nearby positions.

```{r per-position, message = FALSE, warning = FALSE}
affected_pos <- which(effect > 0)
susie_est <- vector("list", length(affected_pos))
for (k in seq_along(affected_pos)) {
  susie_est[[k]] <- susie(X, y[, affected_pos[k]], L = 1L)$sets
}
n_cs_per_pos <- vapply(susie_est, function(s) length(s$cs), integer(1L))
table(n_cs_per_pos)
```

Most positions (the `0` column above) yield no CS at all.
Compare positions 88 and 91 to see the second failure: both
positions carry similar effect amplitude, but the per-position
SuSiE fits land on different lead variants, so reasoning about
"the causal SNP" from per-position fits is not straightforward.

```{r per-position-illustration, fig.height = 4.5, fig.width = 8, message = FALSE, warning = FALSE}
illustrate_pos <- function(t) {
  sf <- susie(X, y[, t], L = 1L)
  list(t = t, pip = sf$pip, cs = sf$sets$cs)
}
ill <- list(illustrate_pos(88L), illustrate_pos(91L))

par(mfrow = c(2L, 2L), mar = c(3.5, 4, 2.5, 1))
# Top row: effect curve with the highlighted position.
for (i in seq_along(ill)) {
  plot(effect, pch = 20L, col = "grey60", cex = 0.7,
       xlab = "position", ylab = "true effect",
       main = sprintf("Position %d", ill[[i]]$t),
       cex.main = 1.05, font.main = 2L)
  abline(h = 0, lty = 3, col = "grey50")
  points(ill[[i]]$t, effect[ill[[i]]$t], pch = 21L,
         col = "firebrick", bg = "firebrick", cex = 1.4)
}
# Bottom row: per-position SuSiE PIP.
for (i in seq_along(ill)) {
  plot(ill[[i]]$pip, pch = 20L, col = "grey60", cex = 0.7,
       xlab = "SNP", ylab = "PIP", ylim = c(0, 1),
       main = sprintf("Per-position SuSiE PIP at position %d",
                      ill[[i]]$t),
       cex.main = 1.0, font.main = 2L)
  abline(v = true_pos_filt, lty = 2, col = "grey50")
  cs_l1 <- ill[[i]]$cs$L1
  if (!is.null(cs_l1)) {
    points(cs_l1, ill[[i]]$pip[cs_l1], pch = 19L,
           col = "dodgerblue", cex = 1.2)
  }
  points(true_pos_filt, ill[[i]]$pip[true_pos_filt], pch = 21L,
         col = "firebrick", bg = NA, cex = 1.6, lwd = 1.8)
}
par(mfrow = c(1L, 1L))
```

Top row: the per-position truth, with the highlighted position
in red. Bottom row: the SuSiE PIP at that position; blue points
mark the credible set, red ring marks the true causal SNP. The
two positions sit a few CpGs apart yet pick different lead
variants, illustrating the per-position-fit incoherence.

## Method 2: Top-PC SuSiE for PC = 1..5

Project `Y` onto its top `pc` principal component and fit
`susie()` on each principal component score:

```{r top-pc, message = FALSE, warning = FALSE}
y_centered <- scale(y, center = TRUE, scale = FALSE)
svd_y <- svd(y_centered)
pc_pip_at_truth <- numeric(5L)
pc_n_cs         <- integer(5L)
for (pc in seq_len(5L)) {
  y_pc <- y_centered %*% svd_y$v[, pc]
  fit_pc <- susie(X, as.numeric(y_pc), L = 1L)
  pc_pip_at_truth[pc] <- fit_pc$pip[true_pos_filt]
  pc_n_cs[pc]         <- length(fit_pc$sets$cs)
}
data.frame(PC = seq_len(5L),
           pip_at_true_snp = pc_pip_at_truth,
           n_cs            = pc_n_cs)
```

PC 1 captures the trait-driven direction cleanly and SuSiE on
the PC 1 score returns a single, high-PIP credible set on the
true SNP. That fixes the per-position-SuSiE recall problem:
collapsing to one scalar derived from the curve gives a
coherent fine-mapping signal. But PCA discards the per-position
effect direction. The fit tells us a signal exists at one
variant; it does not tell us where on the curve that variant
acts. That target information is what fSuSiE preserves and
PCA loses.

## Method 3: mvSuSiE under default and tuned configurations

`mvSuSiE` treats `Y` as a multi-condition response without a
wavelet-basis assumption. Two configurations matter for this
simulation: the default (canonical mixture prior, estimated
residual variance) and the tuned (true outer-product prior,
fixed identity residual variance). The default fails on this
data; the tuned configuration recovers the correct credible
set. The PIP-zero failure mode of the default mvSuSiE on this
simulation is the subject of [mvsusieR issue 61](https://github.com/stephenslab/mvsusieR/issues/61).

```{r mvsusie, eval = has_mvsusie, message = FALSE, warning = FALSE}
library(mvsusieR)

# Tuned configuration: true outer-product prior + fixed identity
# residual variance. This is the configuration that recovers the
# correct CS.
U_true     <- outer(effect, effect)
prior_true <- create_mixture_prior(
  mixture_prior = list(matrices = list(U_true), weights = 1))
V_true     <- diag(T_m)

fit_mv_tuned <- mvsusie(X, y, L = 1L,
                        prior_variance            = prior_true,
                        residual_variance         = V_true,
                        estimate_residual_variance = FALSE,
                        estimate_prior_variance    = TRUE,
                        max_iter = 100L, verbose = FALSE)
cat("PIP at true SNP (tuned):", fit_mv_tuned$pip[true_pos_filt], "\n")
cat("CS count (tuned):", length(fit_mv_tuned$sets$cs), "\n")

# Default configuration: canonical mixture prior, estimated
# residual variance. The PIPs collapse to zero on this data
# (mvsusieR issue 61).
prior_canon <- create_mixture_prior(R = T_m)
fit_mv_default <- mvsusie(X, y, L = 1L,
                          prior_variance = prior_canon,
                          max_iter = 100L, verbose = FALSE)
cat("Max PIP (default):", max(fit_mv_default$pip), "\n")
cat("CS count (default):", length(fit_mv_default$sets$cs), "\n")
```

```{r mvsusie-effect-curve, eval = has_mvsusie, fig.height = 4, fig.width = 8}
# Recovered effect curves from both mvSuSiE configurations.
# `coef()` on an mvsusie fit returns a (J + 1) x R matrix; row
# j + 1 is the fitted effect curve of SNP j across positions.
lead_tuned   <- which.max(fit_mv_tuned$pip)
lead_default <- which.max(fit_mv_default$pip)
mv_eff_tuned   <- as.numeric(coef(fit_mv_tuned)[lead_tuned + 1L, ])
mv_eff_default <- as.numeric(coef(fit_mv_default)[lead_default + 1L, ])

par(mfrow = c(1L, 2L), mar = c(4, 4, 2.5, 1))
yrng <- range(c(effect, mv_eff_tuned, mv_eff_default))
plot(effect, pch = 20L, col = "grey30", cex = 0.7, ylim = yrng,
     xlab = "position", ylab = "effect",
     main = "mvSuSiE tuned (truth-informed prior)",
     cex.main = 1.05, font.main = 2L)
abline(h = 0, lty = 3, col = "grey50")
lines(mv_eff_tuned, col = "firebrick", lwd = 2)
legend("topright", legend = c("truth", "mvSuSiE tuned"),
       col = c("grey30", "firebrick"), lwd = c(NA, 2),
       lty = c(NA, 1), pch = c(20, NA), bty = "n", cex = 0.8)

plot(effect, pch = 20L, col = "grey30", cex = 0.7, ylim = yrng,
     xlab = "position", ylab = "effect",
     main = "mvSuSiE default (canonical prior)",
     cex.main = 1.05, font.main = 2L)
abline(h = 0, lty = 3, col = "grey50")
lines(mv_eff_default, col = "firebrick", lwd = 2)
legend("topright", legend = c("truth", "mvSuSiE default"),
       col = c("grey30", "firebrick"), lwd = c(NA, 2),
       lty = c(NA, 1), pch = c(20, NA), bty = "n", cex = 0.8)
par(mfrow = c(1L, 1L))
```

The tuned configuration recovers the effect shape closely. The
default configuration's recovered curve is essentially flat at
zero: with no truth-informed prior the canonical mixture cannot
distinguish the signal from noise, the PIPs collapse, and the
fitted effect carries no information. The signal is recoverable
only when the true effect direction is supplied as the prior.

## Method 4: fSuSiE

A single `fsusie()` call recovers the correct credible set and
a smooth recovered effect curve, no truth-informed configuration
required.

```{r fsusie, message = FALSE, warning = FALSE}
fit_f <- fsusie(y, X, L = 15L, L_greedy = 5L, verbose = FALSE)
fit_f$sets$cs
fit_f$pip[true_pos_filt]
```

The default `mfsusie_plot()` shows PIPs in the top panel and the
TI-smoothed effect curve with 95% credible band in the bottom
panel. Pass `truth = list(effect)` to overlay the per-position
truth as dark-grey dots; the true causal SNP is highlighted on
the PIP panel with a red ring and a vertical dashed reference.

```{r fsusie-plot-default-dims}
fit_f_s <- mf_post_smooth(fit_f)
plot_dims <- mfsusie_plot_dimensions(fit_f_s)
```

```{r fsusie-plot-default, fig.width = plot_dims$width, fig.height = plot_dims$height}
mfsusie_plot(fit_f_s, truth = list(effect))
# Highlight the true causal SNP on the PIP panel.
abline(v = true_pos_filt, lty = 2, col = "firebrick", lwd = 1.2)
points(true_pos_filt, fit_f_s$pip[true_pos_filt], pch = 21L,
       col = "firebrick", bg = NA, cex = 1.6, lwd = 2)
```

The credible set lands on the true SNP, and the smoothed effect
curve tracks the truth across the affected support.

For a publication-style view, build PIP and per-effect panels
directly from `fit_f`'s posterior fields:

```{r fsusie-plot-ggplot2, message = FALSE, fig.height = 3, fig.width = 8}
library(ggplot2)
library(cowplot)

ti          <- fit_f_s$smoothed$TI
mean_curve  <- ti$effect_curves[[1L]][[1L]]
band        <- ti$credible_bands[[1L]][[1L]]
df_truth    <- data.frame(x = seq_along(effect),     y = effect)
df_curve    <- data.frame(x = seq_along(mean_curve), y = mean_curve)
df_band     <- data.frame(x  = seq_along(mean_curve),
                          lo = band[, 1L], hi = band[, 2L])

P_effect <- ggplot() +
  geom_ribbon(data = df_band,
              mapping = aes(x = x, ymin = lo, ymax = hi),
              fill = "skyblue", alpha = 0.45) +
  geom_point(data = df_truth, mapping = aes(x = x, y = y),
             color = "grey30", size = 1.2) +
  geom_line(data = df_curve, mapping = aes(x = x, y = y),
            color = "firebrick", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  labs(x = "position", y = "effect",
       title = "fSuSiE recovered effect curve") +
  theme_classic()

df_pip <- data.frame(x = seq_along(fit_f$pip), y = fit_f$pip)
P_pip <- ggplot(df_pip, aes(x = x, y = y)) +
  geom_point(color = "grey60", size = 0.9) +
  geom_point(data = df_pip[true_pos_filt, , drop = FALSE],
             color = "firebrick", size = 2.2) +
  geom_vline(xintercept = true_pos_filt,
             linetype = "dashed", color = "firebrick", alpha = 0.6) +
  ylim(0, 1) +
  labs(x = "SNP", y = "PIP", title = "fSuSiE PIPs") +
  theme_classic()

plot_grid(P_pip, P_effect, nrow = 1L, ncol = 2L,
          rel_widths = c(1, 1.2))
```

The PIP scatter on the left puts a single point at PIP near 1 on
the true causal SNP (red); the effect-curve on the right has the
credible band hugging the truth across the affected positions.
The shape, the zero-effect prefix on positions 1..40, and the
peak around the cosine maximum are all recovered cleanly without
any truth-informed prior or pre-specified number of effects.

## When to use which tool in the SuSiE suite

Per-position SuSiE is appropriate when only one position is
expected to carry signal and that position is known a priori.
Top-PC SuSiE is appropriate when the principal-component
direction aligns with the variable-driven effect (rare in
practice). mvSuSiE is appropriate when the response has a
small, fixed number of conditions and a meaningful between-
condition covariance prior is available. fSuSiE is appropriate
when the response is a smooth function of position and the
position grid is power-of-two (or remappable to one), which
covers methylation, accessibility, and coverage-style traits.

## Session info

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
