---
title: "Covariate adjustment and colocalization with fsusie()"
author: "Gao Wang and contributors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Covariate adjustment and colocalization with fsusie()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 3, fig.align = "center", dpi = 120
)
```

This vignette covers two adjacent workflows that wrap a single
`fsusie()` fit: (i) regressing out covariates before fine-mapping, and
(ii) colocalizing two functional outcomes via per-outcome fits and a
shared posterior.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Adjusting for covariates

If a confounder `Z` (e.g., age, sex, batch indicators, principal
components of a global expression matrix) affects the response, fit
the residualised response. The conventional recipe is to project out
`Z` from both `Y` and `X` and fit `fsusie()` on the residuals.

```{r covariate-data}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X); T_m <- 64

Z <- cbind(1, scale(rnorm(n)), scale(rnorm(n)))   # intercept + 2 covariates
gamma <- matrix(rnorm(ncol(Z) * T_m), nrow = ncol(Z))
positions <- seq_len(T_m)
shape <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 6)^2))
beta <- matrix(0, p, T_m); beta[42, ] <- 1.2 * shape
Y <- Z %*% gamma + X %*% beta + matrix(rnorm(n * T_m, sd = 0.4), n)
```

Project out `Z` from `X` and `Y` with the same hat matrix.

```{r residualize}
H_resid <- diag(n) - Z %*% solve(crossprod(Z), t(Z))
X_r <- H_resid %*% X
Y_r <- H_resid %*% Y
fit <- fsusie(Y_r, X_r, pos = positions, L = 3, verbose = FALSE)
which(fit$pip > 0.9)
length(fit$sets$cs)
```

Equivalent options when `Z` is large: regress `Y` on `Z` once with
`stats::lm.fit` and pass the residual matrix; or include `Z` in `X`
with `prior_weights` set to `0` on the covariate columns.

## Colocalisation across two functional outcomes

The standard recipe: fit `fsusie()` on each outcome separately, then
combine the per-effect posterior probabilities of association across
the two CSes into a coloc-style PP.H4 statistic. Because both fits
return per-effect `alpha` matrices on the same `X`, the combination is
a straightforward outer product.

```{r coloc-data}
p <- ncol(X)
beta1 <- matrix(0, p, T_m); beta1[42, ] <- 1.2 * shape
beta2 <- matrix(0, p, T_m); beta2[42, ] <- 0.9 * shape         # shared causal
beta3 <- matrix(0, p, T_m); beta3[88, ] <- 0.9 * shape         # distinct
Y1 <- X %*% beta1 + matrix(rnorm(n * T_m, sd = 0.4), n)
Y2 <- X %*% beta2 + matrix(rnorm(n * T_m, sd = 0.4), n)
Y3 <- X %*% beta3 + matrix(rnorm(n * T_m, sd = 0.4), n)
```

Fit both outcomes and a non-colocalizing third for comparison.

```{r coloc-fit}
fit1 <- fsusie(Y1, X, pos = positions, L = 3, verbose = FALSE)
fit2 <- fsusie(Y2, X, pos = positions, L = 3, verbose = FALSE)
fit3 <- fsusie(Y3, X, pos = positions, L = 3, verbose = FALSE)

# Per-effect colocalization: outer product of alpha rows, summed
# over effects, summed over the diagonal (same variable causal in both).
coloc_pp <- function(a, b) {
  L <- min(nrow(a), nrow(b))
  sum(vapply(seq_len(L), function(l) sum(a[l, ] * b[l, ]), numeric(1)))
}
coloc_pp(fit1$alpha, fit2$alpha)   # high: shared causal
coloc_pp(fit1$alpha, fit3$alpha)   # low: distinct causals
```

For full coloc-style PP.H0..H4 with separate priors, plug the
per-effect alpha matrices into the standard
`coloc::coloc.bf_bf` / `coloc.susie` machinery; mfsusieR's `alpha` and
`lbf_variable` slots have the shape `coloc` expects.

## Notes

- `fsusie()` does not mutate `X` and `Y` in place; the residualisation
  pattern above is the recommended idiom.
- The colocalization snippet here is intentionally minimal. For
  multi-outcome colocalization across more than two outcomes, see
  [the multi-outcome intro](mfsusie_intro.html), which fits all
  outcomes jointly and avoids the post-hoc combine.
