---
title: "Methylation QTL fine-mapping with fSuSiE"
author: "William Denault and Gao Wang"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Methylation QTL fine-mapping with fSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 5, fig.align = "center", dpi = 110,
  out.width = "100%"
)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE) &&
              requireNamespace("cowplot",  quietly = TRUE) &&
              requireNamespace("reshape2", quietly = TRUE)
```

This vignette compares fSuSiE with two non-functional approaches
(per-CpG association tests and per-CpG SuSiE) on a simulated
methylation QTL dataset. fSuSiE returns a single set of
fine-mapping signals together with per-CpG effect estimates;
the per-CpG approaches return one signal per CpG, leaving the
user to integrate them by hand.

## Example data

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
data(dnam_example)
str(dnam_example, max.level = 1)
n   <- nrow(dnam_example$X)
p   <- ncol(dnam_example$X)
T_m <- ncol(dnam_example$Y)
dnam_example$causal_snps
```

The dataset has `n = `r n`` samples, `p = `r p`` SNPs, and
`T = `r T_m`` CpGs. Three causal SNPs (1, 9, 3) act on two
CpG clusters: SNPs 1 and 9 act on CpGs 9-16 with opposite
signs, and SNP 3 acts on CpGs 25-32. SNP 4 is a high-LD
near-clone of SNP 3, which the per-CpG association tests will
not be able to disambiguate. The simulation script is
`data-raw/make_data.R`.

To illustrate, the plot below shows the effect of SNP 3 on
the methylation levels. Each point is one individual at one
CpG, colored by SNP-3 genotype (0/1/2). The bold lines show
the per-genotype mean methylation level at each CpG.

```{r data-by-genotype, fig.height = 2.5, fig.width = 6, eval = has_ggplot}
suppressPackageStartupMessages({
  library(ggplot2); library(cowplot); library(reshape2)
})
X <- dnam_example$X; Y <- dnam_example$Y
geno_chr <- as.character(X[, 3L])
pdat_pts <- data.frame(
  cpg        = rep(seq_len(T_m), each = n) +
                runif(n * T_m, min = -0.2, max = 0.2),
  meth_level = as.vector(Y),
  geno       = factor(rep(geno_chr, T_m)))
mean_by_geno <- aggregate(Y, by = list(geno = geno_chr), FUN = mean)
pdat_lines <- do.call(rbind, lapply(seq_len(nrow(mean_by_geno)),
  function(k) data.frame(
    cpg        = seq_len(T_m),
    meth_level = as.numeric(mean_by_geno[k, -1L]),
    geno       = factor(mean_by_geno[k, "geno"]))))
p_data <- ggplot(pdat_pts,
                 aes(x = cpg, y = meth_level, color = geno)) +
  geom_point(shape = 20, size = 1.0) +
  scale_color_manual(values = c("darkblue", "limegreen", "darkorange")) +
  scale_x_continuous(breaks = c(0, seq(4, T_m, 4))) +
  geom_line(data = pdat_lines,
            aes(x = cpg, y = meth_level, group = geno),
            color = "black", linewidth = 1.6) +
  geom_line(data = pdat_lines,
            aes(x = cpg, y = meth_level, group = geno),
            linewidth = 1.0) +
  labs(x = "CpG", y = "methylation", color = "SNP 3 genotype") +
  theme_cowplot(font_size = 10)
print(p_data)
```

The lines diverge over CpGs 25-32: each additional copy of the
SNP 3 alternate allele lifts the methylation level there. At
the other CpGs the lines run together; SNP 3 has no direct
effect outside its causal cluster.

## Per-CpG association tests

A common QTL-mapping recipe runs a separate association test
for every CpG-SNP pair. Below we use the linear regression
function `lm()`. Plotting the resulting `-log10(p-value)`
matrix in two views (one per SNP, one per CpG) gives a
SNP-centred and a CpG-centred picture of the signal.

```{r per-cpg-lm, fig.height = 2.5, fig.width = 6, eval = has_ggplot}
assoc <- matrix(0, T_m, p)
for (i in seq_len(T_m)) {
  for (j in seq_len(p)) {
    assoc[i, j] <- summary(lm(Y[, i] ~ X[, j]))$coefficients[2L, 4L]
  }
}
threshold <- -log10(0.05 / (T_m * p))
pdat_a <- data.frame(
  cpg    = rep(seq_len(T_m), times = p),
  snp    = rep(seq_len(p),   each = T_m),
  effect = as.vector(t(dnam_example$causal_betas[
              match(seq_len(p), dnam_example$causal_snps), ,
              drop = FALSE]) != 0),
  pval   = as.vector(assoc))
pdat_a <- transform(pdat_a, pval = -log10(pval))
pdat_a$effect[is.na(pdat_a$effect)] <- FALSE
p_snp <- ggplot(pdat_a, aes(x = snp, y = pval,
                            shape = effect, color = effect)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = threshold, color = "black",
             linetype = "dotted") +
  scale_x_continuous(breaks = seq_len(p)) +
  scale_shape_manual(values = c(1, 4)) +
  scale_color_manual(values = c("darkblue", "darkorange")) +
  labs(x = "SNP", y = "-log10 p-value") +
  theme_cowplot(font_size = 10)
p_cpg <- ggplot(pdat_a, aes(x = cpg, y = pval,
                            shape = effect, color = effect)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = threshold, color = "black",
             linetype = "dotted") +
  scale_x_continuous(breaks = c(0, seq(4, T_m, 4))) +
  scale_shape_manual(values = c(1, 4)) +
  scale_color_manual(values = c("darkblue", "darkorange")) +
  labs(x = "CpG", y = "-log10 p-value") +
  theme_cowplot(font_size = 10)
plot_grid(p_snp, p_cpg, nrow = 1L)
```

The SNP-centred view (left) hits Bonferroni significance
(`-log10(0.05 / (T*p))`) at two of the three causal SNPs and
is unable to separate SNP 3 from its high-LD partner SNP 4.
*But more importantly,* this view alone does not tell us which
CpGs are affected by the SNPs.

The CpG-centred view (right) shows that the affected CpGs
cluster in two contiguous regions, but it is less clear which
CpG sites in these clusters are actually affected. *And this
view does not tell us which SNPs drive the changes in
methylation at those CpGs.*

## Per-CpG SuSiE

A natural follow-up is to run SuSiE separately on each
associated CpG. This produces one fine-mapping signal per
CpG, but leaves the user to integrate them.

```{r per-cpg-susie, fig.height = 2, fig.width = 6, eval = has_ggplot}
assoc_cpgs <- sort(unique(subset(pdat_a, pval > threshold)$cpg))
susie_fits <- lapply(assoc_cpgs, function(i)
  susieR::susie(X, Y[, i], L = 3L))
pdat_s <- do.call(rbind, lapply(seq_along(assoc_cpgs), function(k) {
  data.frame(cpg = assoc_cpgs[k], snp = seq_len(p),
             pip = susie_fits[[k]]$pip)
}))
# Color by cluster membership, with one entry per associated
# CpG. CpGs in cluster A (positions <= 16) are red; cluster B
# (positions > 16) is darkblue. The shape cycles through three
# pchs per cluster so multiple CpGs at the same SNP are
# visually separable.
cpg_levels <- sort(unique(pdat_s$cpg))
pal_levels <- ifelse(cpg_levels < 17L, "red", "darkblue")
shape_cycle <- rep(c(1L, 20L, 4L),
                   length.out = length(cpg_levels))
pdat_s$cpg <- factor(pdat_s$cpg, levels = cpg_levels)
p_susie <- ggplot(pdat_s, aes(x = snp, y = pip,
                              color = cpg, shape = cpg)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = pal_levels) +
  scale_shape_manual(values = shape_cycle) +
  scale_x_continuous(breaks = seq_len(p)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "SNP", y = "PIP", color = "CpG", shape = "CpG") +
  theme_cowplot(font_size = 10)
print(p_susie)
```

Running `susie()` on each associated CpG independently gives
one fine-mapping signal per CpG. Two takeaways:

- The signals do not all agree. SNP 1 is identified for the
  CpGs in cluster A with PIP near 1 (consistent recovery), but
  for cluster B some CpGs prefer SNP 3 and others SNP 4 — the
  high-LD partner ambiguity is unresolved in any single fit.
- More importantly, none of the per-CpG SuSiE fits tells us
  *which CpGs* are affected by each causal SNP. The output is
  per-CpG PIPs, not per-SNP affected-CpG sets. fSuSiE solves
  this directly.

## fSuSiE

A single `fsusie()` call recovers all three causal SNPs in
three credible sets. HMM post-processing produces the per-CpG
local false sign rate (lfsr), which the bubble plot below
uses.

```{r run-fsusie, message = FALSE}
fit <- fsusie(Y, X, L = 3L,
              prior_variance_scope = "per_outcome",
              L_greedy = 5L, verbose = TRUE)
fit$sets$cs
fit$pip[dnam_example$causal_snps]
fit_h <- mf_post_smooth(fit, method = "HMM")
```

```{r mfsusie-plot}
mfsusie_plot(fit_h)
```

The PIP panel shows three sharp peaks at the causal SNPs. The
effect panel rows decompose the per-CS effects across CpGs
under HMM post-processing.

## Local-false sign rate of effect sizes

`mfsusie_plot_lfsr()` shows the per-CpG local false sign rate
as a bubble grid. Rows are credible sets, columns are CpGs,
dot size is `-log10(lfsr)` clamped to `[0, cex_max]`. Color
behavior depends on whether `truth` is supplied:

- Without `truth`: dots at CpGs where `lfsr <= lfsr_threshold`
  (default 0.01) get the CS color; the rest are grey.
- With `truth`: dots at the truly-affected CpGs get the CS
  color; the rest are grey, regardless of lfsr.

The simulated `truth_mask` is indexed by causal SNP (SNPs 1,
3, 9). The fit's CS order is determined by IBSS and may not
match. The mapping below pulls the right truth mask for each
CS by matching its lead variant to the causal SNPs:

```{r lfsr-bubble, fig.height = 3, fig.width = 7}
truth_per_cs <- lapply(fit_h$sets$cs, function(cs_vars) {
  hit <- intersect(cs_vars, dnam_example$causal_snps)
  if (length(hit) == 0L) return(rep(FALSE, T_m))
  k <- match(hit[1L], dnam_example$causal_snps)
  dnam_example$truth_mask[[k]]
})
mfsusie_plot_lfsr(fit_h, truth = truth_per_cs)
```

CSes 1 and 2 cover CpGs 9-16; CS 3 covers CpGs 25-32. Both
clusters are recovered cleanly. The legend in the upper-right
shows the dot-size scale; the lower-right indicates the color
rule.

## TI-smoothed effects with errorbars

A second post-processing run with the translation-invariant
smoother gives pointwise credible bands for each per-CpG
effect:

```{r ti-smooth, message = FALSE}
fit_TI <- mf_post_smooth(fit, method = "TI",
                         wavelet_filter = 1L,
                         wavelet_family = "DaubExPhase")
range(fit_h$pip - fit_TI$pip)
```

```{r mfsusie-plot-errorbar, fig.height = 6, fig.width = 7}
mfsusie_plot(fit_TI, effect_style = "errorbar", facet_cs = "stack")
```

Each row is one CS. The dot at each CpG is the posterior mean
effect; the vertical bar is the 95% credible band. Black dots
on the zero line mark CpGs where the band excludes zero (the
estimated affected CpGs). These align with the ground-truth
clusters from the simulation.

## Session info

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```