---
title: "Colocalization across SuSiE-suite fits"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Colocalization across SuSiE-suite fits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette runs a formal colocalization analysis on a pair of
SuSiE-suite fine-mapping fits using
`susieR::susie_post_outcome_configuration(method = "coloc_pairwise")`.
That function is a port of the `coloc::coloc.bf_bf` pairwise-ABF
recipe maintained inside susieR for convenience, so this workflow
adds no soft dependency on the `coloc` package. The recipe
applies to any pair of susie-class fits (`susieR::susie()`,
`mfsusieR::fsusie()`, `mfsusieR::mfsusie()`, `mvsusieR::mvsusie()`,
`susieR::susie_rss()`) because each returns the
`(class = "susie", $sets$cs_index, $lbf_variable)` triple that
`susie_post_outcome_configuration()` reads. `susie_rss()` is
particularly useful for GWAS summary-statistics fine-mapping
fits.

For pre-fit covariate adjustment, see
[the covariates vignette](fsusie_covariates_adjustment.html).

For illustration we use `fsusie()` (single-outcome functional
fine-mapping); `mfsusie()` works the same way and is the natural
choice when both fits cover multiple outcomes.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Data

```{r coloc-data}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X); T_m <- 64
colnames(X) <- paste0("SNP", seq_len(p))
positions <- seq_len(T_m)
shape  <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 6)^2))
beta1  <- matrix(0, p, T_m); beta1[42, ] <- 1.2 * shape
beta2  <- matrix(0, p, T_m); beta2[42, ] <- 0.9 * shape         # shared causal
beta3  <- matrix(0, p, T_m); beta3[88, ] <- 0.9 * shape         # distinct
Y1 <- X %*% beta1 + matrix(rnorm(n * T_m, sd = 0.4), n)
Y2 <- X %*% beta2 + matrix(rnorm(n * T_m, sd = 0.4), n)
Y3 <- X %*% beta3 + matrix(rnorm(n * T_m, sd = 0.4), n)
```

## Fine-mapping

```{r coloc-fit}
fit1 <- fsusie(Y1, X, pos = positions, verbose = FALSE)
fit2 <- fsusie(Y2, X, pos = positions, verbose = FALSE)
fit3 <- fsusie(Y3, X, pos = positions, verbose = FALSE)
```

## Colocalization

`susie_post_outcome_configuration(list(fit1, fit2),
method = "coloc_pairwise")` reads the credible-set indices from
each fit's `$sets$cs` and the per-effect log-Bayes-factor matrix
from `$lbf_variable`. Both are populated by `fsusie()` /
`mfsusie()` on the standard fit path, so the call is a one-liner
per pair of fits. The returned `$coloc_pairwise` data.frame
reports the posterior probabilities of five hypotheses for each
pair of credible sets:

| | meaning |
|---|---|
| `PP.H0` | no causal in either trait |
| `PP.H1` | causal in trait 1 only |
| `PP.H2` | causal in trait 2 only |
| `PP.H3` | causal in both, but at distinct variants |
| `PP.H4` | shared causal variant |

`hit1` and `hit2` name the lead variant in each fit; `l1` and
`l2` are the credible-set indices being compared; `trait1` and
`trait2` index the input fits.

### Shared causal signals

`fit1` and `fit2` were simulated with the same causal SNP at
`SNP42`. The pairwise summary should put nearly all posterior
mass on `PP.H4` (shared causal):

```{r coloc-shared}
res_shared <- susieR::susie_post_outcome_configuration(
  list(fit1, fit2), method = "coloc_pairwise")
res_shared$coloc_pairwise
```

`PP.H4` is essentially 1 and `hit1 == hit2 == SNP42`,
matching the simulation.

### Distinct causal signals

`fit1` carries a signal at SNP42; `fit3` carries one at SNP88.
The pairwise summary should put posterior mass on `PP.H3`
(both have a causal, but at different variants):

```{r coloc-distinct}
res_distinct <- susieR::susie_post_outcome_configuration(
  list(fit1, fit3), method = "coloc_pairwise")
res_distinct$coloc_pairwise
```

`PP.H3` is essentially 1, with `hit1 = SNP42` and
`hit2 = SNP88`.

## Notes

- This vignette assumes the user already has a pair of
  fine-mapping fits, possibly from different SuSiE Suite
  packages, and wants to test for colocalization between them.
  The natural alternative when subject-level data is available
  for all outcomes is to run `mfsusie()` jointly across them;
  the multi-outcome fit shares signal across modalities and
  identifies per-outcome and shared credible sets in one pass
  without a post-hoc combination step. The pairwise-ABF
  recipe shown here becomes essential when one or more inputs
  are summary-statistics fits (e.g., a GWAS `susie_rss()` fit
  colocalized against a molecular-QTL `fsusie()` fit), where
  the underlying individual-level data is not available.
- Joint `mfsusie()` fine-mapping that takes summary statistics
  from a GWAS as one of its outcomes is future work.
- For multi-outcome colocalization across more than two
  outcomes, see [the multi-outcome intro](mfsusie_intro.html),
  which fits all outcomes jointly and avoids the post-hoc
  combine. The disadvantage is that it would subject to mfSuSiE modeling
  assumptions and not as modular as integrating SuSiE Suite fits.

## Session info

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
