---
title: "Adjusting for covariates in functional fine-mapping"
author: "William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Adjusting for covariates in functional fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

This vignette adjusts a functional response `Y` for covariate
effects before fine-mapping. Observed covariates that affect the
trait at every position (age, sex, batch, principal components
of a global expression matrix) inflate the residual variance and
attenuate the genetic signal if left in. The right tool for
removing such effects on a functional response is a wavelet-
domain empirical-Bayes regression that accommodates smooth
position-dependent covariate effects. mfsusieR exposes this
through `mf_adjust_for_covariates()`.

## Setup

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
library(wavethresh)
set.seed(2)
```

## Simulated data

```{r sim}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(100)]
n <- nrow(X); p <- ncol(X); T_m <- 128L
positions <- seq_len(T_m)

# Three covariate functions, each smooth in position. These act
# like global confounders that influence the response across all
# CpGs / positions.
smooth_func <- function() {
  raw <- rnorm(T_m)
  w   <- wd(raw, filter.number = 1L, family = "DaubExPhase")
  for (s in 0L:(log2(T_m) - 1L)) {
    coefs_s <- accessD(w, level = s)
    w <- putD(w, level = s,
              v = coefs_s / (1.5^(log2(T_m) - s)))
  }
  wr(w)
}
f1_cov <- smooth_func()
f2_cov <- smooth_func()
f3_cov <- smooth_func()
Cov <- cbind(scale(rnorm(n)), scale(rnorm(n)), scale(rnorm(n)))

# Genetic effect: a single causal SNP at position 42 with a
# peaked Gaussian effect.
shape <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 12)^2))
beta  <- matrix(0, p, T_m); beta[42L, ] <- 1.2 * shape

# The response: covariate signal + genetic signal + noise.
Y <- Cov %*% rbind(f1_cov, f2_cov, f3_cov) +
     X   %*% beta +
     matrix(rnorm(n * T_m, sd = 0.4), n)
```

## Naive fit, ignoring covariates

```{r naive-fit, message = FALSE}
fit_naive <- fsusie(Y, X, pos = positions,
                    L = 15, L_greedy = 5, verbose = TRUE)
fit_naive$sets$cs
fit_naive$pip[42L]
```

The covariate-driven baseline shifts dilute the signal. The
true causal SNP at position 42 may still be selected but its PIP
is degraded.

## Adjusting for covariates

`mf_adjust_for_covariates()` fits a wavelet-domain empirical-
Bayes regression of `Y` on the covariates `Z = Cov` and returns
the residualised response `Y_adjusted = Y - Z %*% fitted_func`.
The default method `"wavelet_eb"` is appropriate when covariate
effects vary smoothly across positions; the closed-form OLS
alternative `"ols"` is the no-shrinkage limit.

```{r adjust, message = FALSE}
adj <- mf_adjust_for_covariates(Y, Cov,
                                method = "wavelet_eb")
str(adj, max.level = 1)
```

`adj$Y_adjusted` is the residualised response on the original
position grid, ready to feed into `fsusie()`.

## Fine-mapping the adjusted response

```{r adjusted-fit, message = FALSE}
fit_adj <- fsusie(adj$Y_adjusted, X, pos = positions,
                  L = 15, L_greedy = 5, verbose = TRUE)
fit_adj$sets$cs
fit_adj$pip[42L]
```

PIP at the causal SNP is sharper after adjustment. The
recovered effect curve from `mfsusie_plot()`:

```{r adj-plot}
fit_adj_s <- mf_post_smooth(fit_adj, method = "TI",
                            wavelet_filter = 1L,
                            wavelet_family = "DaubExPhase")
mfsusie_plot(fit_adj_s)
```

## OLS as a closed-form sanity check

When covariates are few, well-conditioned, and known to act
linearly per position, the closed-form OLS adjustment is faster
and shrinkage-free:

```{r ols, message = FALSE}
adj_ols <- mf_adjust_for_covariates(Y, Cov, method = "ols")
fit_ols <- fsusie(adj_ols$Y_adjusted, X, pos = positions,
                  L = 15, L_greedy = 5, verbose = FALSE)
fit_ols$sets$cs
fit_ols$pip[42L]
```

The OLS path projects out the covariate column space without
modelling smoothness. For three uncorrelated covariates with
n = `r n` and T = `r T_m`, OLS and the wavelet-EB shrinkage give
qualitatively similar results.

## Adjusting the genotype matrix as well

When covariates are correlated with genotype, the Frisch-Waugh-
Lovell correction also residualises `X`. Pass `X` to
`mf_adjust_for_covariates()` and the function returns
`X_adjusted = (I - Z (Z'Z)^{-1} Z') X`:

```{r fwl}
adj_fwl <- mf_adjust_for_covariates(Y, Cov, X = X,
                                    method = "wavelet_eb")
fit_fwl <- fsusie(adj_fwl$Y_adjusted, adj_fwl$X_adjusted,
                  pos = positions, L = 15, L_greedy = 5,
                  verbose = FALSE)
fit_fwl$sets$cs
fit_fwl$pip[42L]
```

## Next steps

For colocalization across two functional fits, see
[the colocalization vignette](fsusie_colocalization.html). For
post-fit smoothing of effect curves, see
[the post-processing vignette](post_processing.html).
