---
title: "Adjusting for covariates and colocalization analysis"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Adjusting for covariates and colocalization analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

This vignette covers two adjacent workflows: (i) regressing out
covariates before fine-mapping with `fsusie()`, and (ii) running a
formal colocalization analysis on a pair of fine-mapping fits using
`coloc::coloc.susie()`. The colocalization recipe applies to **any
pair of susie-class fits** in the SuSiE suite — `susieR::susie()`,
`mfsusieR::fsusie()`, `mfsusieR::mfsusie()`, `mvsusieR::mvsusie()`,
and `susieR::susie_rss()` — because each returns the
`(class = "susie", $sets$cs_index, $lbf_variable)` triple that
`coloc.susie()` reads. `susie_rss()` is particularly useful for
GWAS summary-statistics fine-mapping fits.

For illustration we use `fsusie()` (single-outcome functional
fine-mapping); `mfsusie()` works the same way and is the natural
choice when both fits cover multiple outcomes.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
library(coloc)
set.seed(1)
```

## Adjusting for covariates

If a confounder `Z` (e.g., age, sex, batch indicators, principal
components of a global expression matrix) affects the response, fit
the residualised response. The conventional recipe is to project out
`Z` from both `Y` and `X` and fit `fsusie()` on the residuals.

```{r covariate-data}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X); T_m <- 64
colnames(X) <- paste0("SNP", seq_len(p))

Z <- cbind(1, scale(rnorm(n)), scale(rnorm(n)))   # intercept + 2 covariates
gamma <- matrix(rnorm(ncol(Z) * T_m), nrow = ncol(Z))
positions <- seq_len(T_m)
shape <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 6)^2))
beta <- matrix(0, p, T_m); beta[42, ] <- 1.2 * shape
Y <- Z %*% gamma + X %*% beta + matrix(rnorm(n * T_m, sd = 0.4), n)
```

Project out `Z` from `X` and `Y` with the same hat matrix.

```{r residualize}
H_resid <- diag(n) - Z %*% solve(crossprod(Z), t(Z))
X_r <- H_resid %*% X
Y_r <- H_resid %*% Y
fit <- fsusie(Y_r, X_r, pos = positions, L = 15, L_greedy = 5, verbose = TRUE)
which(fit$pip > 0.9)
length(fit$sets$cs)
```

Equivalent options when `Z` is large: regress `Y` on `Z` once with
`stats::lm.fit` and pass the residual matrix; or include `Z` in `X`
with `prior_weights` set to `0` on the covariate columns.

## Colocalization with `coloc::coloc.susie()`

For each outcome we fit `fsusie()` separately, then pass both fits
to `coloc::coloc.susie()`. The function reads the SuSiE-class
slots (`$sets$cs_index` to identify retained effects, `$lbf_variable`
for the per-effect log Bayes factors over variants) and returns
the standard PP.H0..PP.H4 posterior table per pair of credible
sets.

```{r coloc-data}
beta1 <- matrix(0, p, T_m); beta1[42, ] <- 1.2 * shape
beta2 <- matrix(0, p, T_m); beta2[42, ] <- 0.9 * shape         # shared causal
beta3 <- matrix(0, p, T_m); beta3[88, ] <- 0.9 * shape         # distinct
Y1 <- X %*% beta1 + matrix(rnorm(n * T_m, sd = 0.4), n)
Y2 <- X %*% beta2 + matrix(rnorm(n * T_m, sd = 0.4), n)
Y3 <- X %*% beta3 + matrix(rnorm(n * T_m, sd = 0.4), n)
```

Fit each outcome:

```{r coloc-fit}
fit1 <- fsusie(Y1, X, pos = positions, L = 15, L_greedy = 5, verbose = TRUE)
fit2 <- fsusie(Y2, X, pos = positions, L = 15, L_greedy = 5, verbose = TRUE)
fit3 <- fsusie(Y3, X, pos = positions, L = 15, L_greedy = 5, verbose = TRUE)
```

`coloc.susie()` expects `colnames(X)` to be present so it can label
posteriors by variant. Both fits run on the same `X`, so the labels
align by construction.

```{r coloc-shared}
res_shared <- coloc.susie(fit1, fit2)
res_shared$summary
```

`PP.H4.abf` is close to 1 for the (`fit1`, `fit2`) pair: a single
shared causal variant `SNP42` drives both outcomes.

```{r coloc-distinct}
res_distinct <- coloc.susie(fit1, fit3)
res_distinct$summary
```

`PP.H3.abf` is close to 1 for (`fit1`, `fit3`): both outcomes have
a causal in the locus, but the causal variants are different
(`SNP42` vs `SNP88`).

The same call works on `mfsusie()` fits, on `susieR::susie()` fits,
on `mvsusieR::mvsusie()` fits, and on `susieR::susie_rss()` fits;
the inputs `coloc.susie` reads (`class`, `$sets$cs_index`,
`$lbf_variable`) are part of the SuSiE-class contract.

## Notes

- This vignette assumes the user already has a pair of
  fine-mapping fits, possibly from different SuSiE-suite
  packages, and wants to test for colocalization between them.
  The natural alternative when subject-level data is available
  for all outcomes is to run `mfsusie()` jointly across them; the
  multi-outcome fit shares signal across modalities and identifies
  per-outcome and shared credible sets in one pass without a
  post-hoc combination step. `coloc.susie()` becomes essential
  when one or more inputs are summary-statistics fits (e.g., a
  GWAS `susie_rss()` fit colocalized against a molecular-QTL
  `fsusie()` fit), where the underlying individual-level data is
  not available.
- Joint `mfsusie()` fine-mapping that takes summary statistics
  from a GWAS as one of its outcomes (so it can colocalize and
  fine-map jointly with molecular-QTL outcomes) is future work.
- The colocalization snippet here is intentionally minimal. For
  multi-outcome colocalization across more than two outcomes, see
  [the multi-outcome intro](mfsusie_intro.html), which fits all
  outcomes jointly and avoids the post-hoc combine.
