---
title: "Methylation QTL fine-mapping with fSuSiE"
author: "William Denault and Gao Wang"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Methylation QTL fine-mapping with fSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 5, fig.align = "center", dpi = 110,
  out.width = "100%"
)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE) &&
              requireNamespace("cowplot",  quietly = TRUE) &&
              requireNamespace("reshape2", quietly = TRUE)
```

This vignette compares fSuSiE with two non-functional approaches
(per-CpG association tests and per-CpG SuSiE) on a simulated
methylation QTL dataset. fSuSiE returns a single set of
fine-mapping signals and per-CpG effect estimates; the
per-CpG approaches return one signal per CpG, leaving the user
to integrate them by hand.

## Example data

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
data(dnam_example)
str(dnam_example, max.level = 1)
n <- nrow(dnam_example$X)
p <- ncol(dnam_example$X)
T_m <- ncol(dnam_example$Y)
dnam_example$causal_snps
```

The dataset has `n = `r n`` samples, `p = `r p`` SNPs, and
`T = `r T_m`` CpGs. Three causal SNPs (1, 9, 3) act on two
CpG clusters: SNPs 1 and 9 act on CpGs 9-16 with opposite
signs, and SNP 3 acts on CpGs 25-32. SNP 4 is a high-LD
near-clone of SNP 3, which the association tests will not be
able to disambiguate. The simulation script is
`data-raw/make_data.R`.

```{r data-by-genotype, fig.height = 2.5, fig.width = 6, eval = has_ggplot}
suppressPackageStartupMessages({
  library(ggplot2); library(cowplot); library(reshape2)
})
X <- dnam_example$X; Y <- dnam_example$Y
pdat <- data.frame(
  cpg        = rep(seq_len(T_m), each = n) +
                runif(n * T_m, min = -0.2, max = 0.2),
  meth_level = as.vector(Y),
  geno       = factor(X[, 3]))
p_data <- ggplot(pdat, aes(x = cpg, y = meth_level, color = geno)) +
  geom_point(shape = 20, size = 1.0) +
  scale_color_manual(values = c("darkblue", "limegreen", "darkorange")) +
  scale_x_continuous(breaks = c(0, seq(4, T_m, 4))) +
  labs(x = "CpG", y = "methylation", color = "SNP 3 genotype") +
  theme_cowplot(font_size = 10)
print(p_data)
```

## Per-CpG association tests

```{r per-cpg-lm, fig.height = 2.5, fig.width = 6, eval = has_ggplot}
assoc <- matrix(0, T_m, p)
for (i in seq_len(T_m)) {
  for (j in seq_len(p)) {
    assoc[i, j] <- summary(lm(Y[, i] ~ X[, j]))$coefficients[2, 4]
  }
}
threshold <- -log10(0.05 / (T_m * p))
pdat_a <- data.frame(
  cpg    = rep(seq_len(T_m), times = p),
  snp    = rep(seq_len(p),   each = T_m),
  effect = as.vector(t(dnam_example$causal_betas[
              match(seq_len(p), dnam_example$causal_snps), ,
              drop = FALSE]) != 0),
  pval   = as.vector(assoc))
pdat_a <- transform(pdat_a, pval = -log10(pval))
pdat_a$effect[is.na(pdat_a$effect)] <- FALSE
p_snp <- ggplot(pdat_a, aes(x = snp, y = pval,
                            shape = effect, color = effect)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = threshold, color = "black",
             linetype = "dotted") +
  scale_x_continuous(breaks = seq_len(p)) +
  scale_shape_manual(values = c(1, 4)) +
  scale_color_manual(values = c("darkblue", "darkorange")) +
  labs(x = "SNP", y = "-log10 p-value") +
  theme_cowplot(font_size = 10)
p_cpg <- ggplot(pdat_a, aes(x = cpg, y = pval,
                            shape = effect, color = effect)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = threshold, color = "black",
             linetype = "dotted") +
  scale_x_continuous(breaks = c(0, seq(4, T_m, 4))) +
  scale_shape_manual(values = c(1, 4)) +
  scale_color_manual(values = c("darkblue", "darkorange")) +
  labs(x = "CpG", y = "-log10 p-value") +
  theme_cowplot(font_size = 10)
plot_grid(p_snp, p_cpg, nrow = 1)
```

The per-SNP view identifies two of the three causal SNPs at
the Bonferroni threshold and does not separate SNP 3 from its
high-LD partner SNP 4. The per-CpG view shows the affected
CpGs cluster in two regions but does not assign causal SNPs to
them.

## Per-CpG SuSiE

```{r per-cpg-susie, fig.height = 2, fig.width = 6, eval = has_ggplot}
assoc_cpgs <- sort(unique(subset(pdat_a, pval > threshold)$cpg))
susie_fits <- lapply(assoc_cpgs, function(i)
  susie(X, Y[, i], L = 3))
pdat_s <- do.call(rbind, lapply(seq_along(assoc_cpgs), function(k) {
  data.frame(cpg = assoc_cpgs[k], snp = seq_len(p),
             pip = susie_fits[[k]]$pip)
}))
pal <- ifelse(pdat_s$cpg < 17, "red", "darkblue")
p_susie <- ggplot(pdat_s,
                  aes(x = snp, y = pip,
                      color = factor(cpg), shape = factor(cpg))) +
  geom_point(size = 1.5) +
  scale_color_manual(values = unique(pal)[
                       as.integer(factor(unique(pdat_s$cpg)))]) +
  scale_shape_manual(values = rep(c(1, 20, 4),
                       length.out = length(unique(pdat_s$cpg)))) +
  scale_x_continuous(breaks = seq_len(p)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "SNP", y = "PIP", color = "CpG", shape = "CpG") +
  theme_cowplot(font_size = 10)
print(p_susie)
```

Each per-CpG SuSiE fit produces its own set of fine-mapping
signals. Some CpGs in cluster B prefer SNP 3, others prefer
SNP 4. There is no principled way to combine these per-CpG
PIPs into a single per-SNP fine-mapping signal.

## fSuSiE

A single `fsusie()` call recovers all three causal SNPs in
three separate credible sets. HMM post-processing produces the
per-position lfsr, which the bubble plot below uses.

```{r run-fsusie, message = FALSE}
fit <- fsusie(Y, X, L = 3,
              prior_variance_scope = "per_outcome",
              L_greedy = 5, verbose = TRUE)
fit$sets$cs
fit$pip[dnam_example$causal_snps]
fit_h <- mf_post_smooth(fit, method = "HMM")
```

```{r mfsusie-plot}
mfsusie_plot(fit_h)
```

The PIP panel shows three sharp peaks at the causal SNPs. The
effect panel rows decompose the per-CS effects across CpGs
under HMM post-processing.

## lfsr bubble grid

`mfsusie_plot_lfsr()` shows the per-position local false sign
rate as a bubble grid. Dot size is `-log10(lfsr)`; positions
where `lfsr <= 0.01` (the default threshold) are colored by
the CS. Passing the simulated `truth_mask` overlays the known
affected CpGs in CS color and leaves the rest grey.

The simulated `truth_mask` is indexed by causal SNP (SNPs 1,
3, 9). The fit's CS order is determined by IBSS and may not
match. The mapping below pulls the right truth mask for each
CS by matching its lead variant to the causal SNPs:

```{r lfsr-bubble, fig.height = 3, fig.width = 7}
truth_per_cs <- lapply(fit_h$sets$cs, function(cs_vars) {
  hit <- intersect(cs_vars, dnam_example$causal_snps)
  if (length(hit) == 0L) return(rep(FALSE, T_m))
  k <- match(hit[1L], dnam_example$causal_snps)
  dnam_example$truth_mask[[k]]
})
mfsusie_plot_lfsr(fit_h, truth = truth_per_cs)
```

The CS-affected CpGs (large colored dots) align with the
ground-truth clusters: CSes 1 and 2 cover CpGs 9-16, CS 3
covers CpGs 25-32.

## TI-smoothed effects with errorbars

A second post-processing run with the translation-invariant
smoother gives pointwise credible bands for each per-CpG
effect:

```{r ti-smooth, message = FALSE}
fit_TI <- mf_post_smooth(fit, method = "TI",
                         wavelet_filter = 1L,
                         wavelet_family = "DaubExPhase")
range(fit_h$pip - fit_TI$pip)
```

```{r mfsusie-plot-errorbar, fig.height = 6, fig.width = 7}
mfsusie_plot(fit_TI, effect_style = "errorbar", facet_cs = "stack")
```

Each row is one CS. The dot at each CpG is the posterior mean
effect; the vertical bar is the 95% credible band. Black dots
on the zero line mark CpGs where the band excludes zero (the
estimated affected CpGs). These align with the ground-truth
clusters from the simulation.

## Prediction on new genotypes

```{r predict, fig.height = 3.5}
fitted_curves <- predict(fit_TI, newx = X[1:3, ])[[1]]
pos_remap <- fit_TI$dwt_meta$pos[[1]]
matplot(pos_remap, t(fitted_curves), type = "l", lty = 1,
        col = c("#E69F00", "#56B4E9", "#009E73"), lwd = 1.5,
        xlab = "CpG (post-remap grid)",
        ylab = "predicted methylation")
legend("topright", legend = paste("subject", 1:3),
       col = c("#E69F00", "#56B4E9", "#009E73"),
       lwd = 1.5, bty = "n")
```

## Multi-region or multi-omic extensions

For a multi-region or multi-outcome version of the same
workflow (e.g., methylation across multiple regions with
expression at the same locus), see
[the multi-outcome intro](mfsusie_intro.html), which fits
all outcomes jointly under the mfsusie model.
