---
title: "Why functional fine-mapping"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Why functional fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 5, fig.align = "center", dpi = 110,
  out.width = "100%"
)
has_mvsusie <- requireNamespace("mvsusieR", quietly = TRUE)
```

Three common alternatives for fine-mapping a functional response
all lose information relative to fSuSiE: per-position SuSiE,
top-PC SuSiE, and naive multivariate SuSiE. This vignette runs
the same simulation that ships with the broader SuSiE-suite
diagnostics and shows the four methods side by side. The
canonical setup is a single causal SNP whose effect spans 88
out of 128 positions; we report PIPs and recovered effect
curves for each method.

## Simulation

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
data(N3finemapping)
X_full <- N3finemapping$X[seq_len(100), ]
T_m    <- 128L

effect <- 1.2 * cos(seq_len(T_m) / T_m * 3 * pi)
effect[effect < 0] <- 0
effect[seq_len(40)] <- 0     # zero on positions 1..40

true_pos <- 700L
y <- matrix(0, 100L, T_m)
for (i in seq_len(100L)) {
  y[i, ] <- X_full[i, true_pos] * effect + rnorm(T_m)
}
plot(effect, pch = 19, xlab = "position", ylab = "true effect")
```

```{r filter-zero-var}
col_var <- apply(X_full, 2, var)
keep    <- which(col_var > 0)
X       <- X_full[, keep]
true_pos_filt <- which(keep == true_pos)
cat("N =", nrow(X), "  J =", ncol(X), "  T =", T_m,
    "  true SNP filtered index =", true_pos_filt, "\n")
```

## Method 1: Per-position SuSiE

For each affected position run `susie(X, y[, t], L = 1)` and
collect the credible sets. The fit fails to detect a credible
set at most positions, and where it does the credible sets are
inconsistent across nearby positions.

```{r per-position, message = FALSE, warning = FALSE}
affected_pos <- which(effect > 0)
susie_est <- vector("list", length(affected_pos))
for (k in seq_along(affected_pos)) {
  susie_est[[k]] <- susie(X, y[, affected_pos[k]], L = 1L)$sets
}
n_cs_per_pos <- vapply(susie_est, function(s) length(s$cs), integer(1L))
table(n_cs_per_pos)
```

The "table" entry at zero is the count of CpGs where SuSiE
returns no credible set; only a small minority of affected
positions yield a non-empty CS.

## Method 2: Top-PC SuSiE for PC = 1..5

Project `Y` onto its top `pc` principal component and fit
`susie()` on each principal component score:

```{r top-pc, message = FALSE, warning = FALSE}
y_centered <- scale(y, center = TRUE, scale = FALSE)
svd_y <- svd(y_centered)
pc_pip_at_truth <- numeric(5L)
pc_n_cs         <- integer(5L)
for (pc in seq_len(5L)) {
  y_pc <- y_centered %*% svd_y$v[, pc]
  fit_pc <- susie(X, as.numeric(y_pc), L = 1L)
  pc_pip_at_truth[pc] <- fit_pc$pip[true_pos_filt]
  pc_n_cs[pc]         <- length(fit_pc$sets$cs)
}
data.frame(PC = seq_len(5L),
           pip_at_true_snp = pc_pip_at_truth,
           n_cs            = pc_n_cs)
```

Most of the top-PC fits fail to recover the true causal SNP:
the principal-component direction is dominated by orthogonal
noise rather than the variable-driven signal.

## Method 3: mvSuSiE under default and tuned configurations

`mvSuSiE` treats `Y` as a multi-condition response without a
wavelet-basis assumption. Two configurations matter for this
simulation: the default (canonical mixture prior, estimated
residual variance) and the tuned (true outer-product prior,
fixed identity residual variance). The default fails on this
data; the tuned configuration recovers the correct credible
set. The PIP-zero failure mode of the default mvSuSiE on this
simulation is the subject of mvsusieR issue 61.

```{r mvsusie, eval = has_mvsusie, message = FALSE, warning = FALSE}
library(mvsusieR)

# Tuned configuration: true outer-product prior + fixed identity
# residual variance. This is the configuration that recovers the
# correct CS.
U_true     <- outer(effect, effect)
prior_true <- create_mixture_prior(
  mixture_prior = list(matrices = list(U_true), weights = 1))
V_true     <- diag(T_m)

fit_mv_tuned <- mvsusie(X, y, L = 1L,
                        prior_variance            = prior_true,
                        residual_variance         = V_true,
                        estimate_residual_variance = FALSE,
                        estimate_prior_variance    = TRUE,
                        max_iter = 100L, verbose = FALSE)
cat("PIP at true SNP (tuned):", fit_mv_tuned$pip[true_pos_filt], "\n")
cat("CS count (tuned):", length(fit_mv_tuned$sets$cs), "\n")

# Default configuration: canonical mixture prior, estimated
# residual variance. The PIPs collapse to zero on this data
# (mvsusieR issue 61).
prior_canon <- create_mixture_prior(R = T_m)
fit_mv_default <- mvsusie(X, y, L = 1L,
                          prior_variance = prior_canon,
                          max_iter = 100L, verbose = FALSE)
cat("Max PIP (default):", max(fit_mv_default$pip), "\n")
cat("CS count (default):", length(fit_mv_default$sets$cs), "\n")
```

```{r mvsusie-effect-curve, eval = has_mvsusie, fig.height = 3.5}
# Recovered effect curve from the tuned mvSuSiE fit. coef() on
# an mvsusie fit returns a (J + 1) x R matrix; row j + 1 is the
# fitted effect curve of SNP j across positions.
lead_snp <- which.max(fit_mv_tuned$pip)
mv_effect <- as.numeric(coef(fit_mv_tuned)[lead_snp + 1L, ])

plot(effect, type = "l", lwd = 2, col = "black",
     xlab = "position", ylab = "effect",
     main = "mvSuSiE (tuned) recovered effect")
lines(mv_effect, col = "#1f78b4", lwd = 1.5, lty = 2)
legend("topright", legend = c("truth", "mvSuSiE (tuned)"),
       col = c("black", "#1f78b4"), lwd = c(2, 1.5),
       lty = c(1, 2), bty = "n")
```

## Method 4: fSuSiE

A single `fsusie()` call recovers the correct credible set and
a smooth recovered effect curve.

```{r fsusie, message = FALSE, warning = FALSE}
fit_f <- fsusie(y, X, L = 1L, verbose = TRUE)
fit_f$sets$cs
fit_f$pip[true_pos_filt]
```

```{r fsusie-plot}
fit_f_s <- mf_post_smooth(fit_f, method = "TI",
                          wavelet_filter = 1L,
                          wavelet_family = "DaubExPhase")
mfsusie_plot(fit_f_s)
```

## When to use which tool in the SuSiE suite

Per-position SuSiE is appropriate when only one position is
expected to carry signal and that position is known a priori.
Top-PC SuSiE is appropriate when the principal-component
direction aligns with the variable-driven effect (rare in
practice). mvSuSiE is appropriate when the response has a
small, fixed number of conditions and a meaningful between-
condition covariance prior is available. fSuSiE is appropriate
when the response is a smooth function of position and the
position grid is power-of-two (or remappable to one), which
covers methylation, accessibility, and coverage-style traits.
