---
title: "Getting started"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

`mfsusieR` is the multi-outcome, functional regression generalization of
`susieR::susie()`. It provides two main functions:

- `fsusie(Y, X, pos = NULL, ...)` for fine-mapping a **single**
  response. `Y` may be scalar (`T = 1`, recovers
  `susieR::susie()`) or functional (`T > 1`, the single-outcome
  functional SuSiE model).
- `mfsusie(X, Y, pos = NULL, ...)` for fine-mapping **multiple**
  responses jointly. `Y` is a list of length `M` outcomes; each
  element is a matrix `n x T_m` (with `T_m = 1` for scalar
  outcomes).

Both functions return an object of class `c("mfsusie", "susie")` with
the standard SuSiE fields (`alpha`, `mu`, `mu2`, `lbf`, `KL`, `sigma2`,
`elbo`, `pip`, `sets`, `niter`, `converged`).

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

Both fits return raw inverse-DWT effect curves through `coef()`.
For a smoothed position-space curve with a pointwise credible
band (and, for the HMM smoother, a per-position lfsr curve),
post-process with `mf_post_smooth()`: see
[`vignette("post_processing")`](post_processing.html) for the
three available methods (`scalewise`, `TI`, `HMM`).

## A small `fsusie()` example (single outcome, functional)

Simulate one functional response on a length-`32` grid with two causal
variables.

```{r fsusie-fit}
n <- 200; p <- 50; T_m <- 32
X <- matrix(rnorm(n * p), nrow = n)
beta <- numeric(p); beta[c(5, 17)] <- c(1.2, -0.8)
Y <- X %*% matrix(rep(beta, T_m), nrow = p) +
     matrix(rnorm(n * T_m, sd = 0.4), nrow = n)

fit_f <- fsusie(Y, X, L = 15, L_greedy = 5, verbose = TRUE)
fit_f$pip[c(5, 17)]
length(fit_f$sets$cs)
```

`fsusie()` is a wrapper around `mfsusie(X, list(Y), list(pos),
...)` with no numerical difference.
`mfsusie_plot()` works on the result either way:

```{r fsusie-plot, fig.width = 6, fig.height = 5}
mfsusie_plot(fit_f)
```

## A small `mfsusie()` example (multi-outcome)

Same `X` and effect vector, two functional outcomes at different
resolutions plus one scalar outcome.

```{r mfsusie-fit}
T_per <- c(32L, 64L)
Y_func <- lapply(T_per, function(T_m) {
  X %*% matrix(rep(beta, T_m), nrow = p) +
    matrix(rnorm(n * T_m, sd = 0.4), nrow = n)
})
Y_scalar <- as.numeric(X %*% beta + rnorm(n, sd = 0.4))
Y_list <- c(Y_func, list(matrix(Y_scalar, ncol = 1)))

fit_m <- mfsusie(X, Y_list, L = 15, L_greedy = 5,
                 prior_variance_scope = "per_outcome",
                 verbose = TRUE)
fit_m$pip[c(5, 17)]
length(fit_m$sets$cs)
```

`mu[[l]][[m]]` is a `p x T_basis[m]` matrix of per-effect posterior
means on each outcome; `predict()` and `coef()` invert the wavelet
transform to position space transparently. `mfsusie_plot()` tiles
one effect-curve panel per outcome with the PIPs in the top-left
slot:

```{r mfsusie-plot-getting-started, fig.width = 8, fig.height = 6}
mfsusie_plot(fit_m)
```

## Numerical equivalence with `susieR::susie` (degenerate case)

When `Y` is scalar, the prior collapses to a single Gaussian component,
the null component is removed, and the per-(scale, outcome) variance
structure collapses to a scalar, `mfsusie()` reduces **exactly** to
`susieR::susie()`. This is the C1 contract in the unit test suite at
tolerance `<= 1e-13` (see `tests/testthat/test_susier_degeneracy.R`).

```{r c1-degeneracy}
y_scalar <- as.numeric(X %*% beta + rnorm(n, sd = 0.4))
y <- (y_scalar - mean(y_scalar)) / sd(y_scalar)

fit_s <- susie(X, y, L = 5,
               scaled_prior_variance      = 0.2,
               estimate_prior_variance    = FALSE,
               estimate_residual_variance = TRUE,
               max_iter = 100, tol = 1e-8)

fit_d <- mfsusie(X, list(matrix(y, ncol = 1)),
                 L                        = 5,
                 prior_variance_grid      = 0.2,
                 prior_variance_scope     = "per_outcome",
                 null_prior_weight        = 0,
                 residual_variance_scope  = "per_outcome",
                 mixture_weight_method    = "none",
                 L_greedy                 = NULL,
                 max_iter = 100, tol = 1e-8,
                 verbose = FALSE)

# Element-wise agreement on every numeric field.
max(abs(fit_d$alpha - fit_s$alpha))
max(abs(fit_d$pip   - fit_s$pip))
max(abs(fit_d$sigma2[[1]] - fit_s$sigma2))
max(abs(fit_d$lbf   - fit_s$lbf))
max(abs(fit_d$KL    - fit_s$KL))
abs(tail(fit_d$elbo, 1) - tail(fit_s$elbo, 1))
identical(fit_d$niter, fit_s$niter)
```

The remaining differences are at the IEEE-754 double-precision floor
(~`1e-16` for `alpha`, `sigma2`, `pip`; ~`1e-13` for `KL` because the
KL contribution accumulates over `n` residuals in slightly different
order). The credible-set memberships are identical:

```{r c1-cs}
identical(lapply(fit_d$sets$cs, sort),
          lapply(fit_s$sets$cs, sort))
```

This degenerate-case identity is enforced by the test suite on every
build. If a refactor introduces drift above `1e-10`, CI fails.

## Where to next

- Single-outcome applications:
  [Single-outcome intro](fsusie_intro.html),
  [covariate adjustment + colocalization](fsusie_covariates_and_coloc.html),
  [DNAm case study](fsusie_dnam_case_study.html),
  [why functional](fsusie_why_functional.html).
- Multi-outcome applications:
  [Multi-outcome intro](mfsusie_intro.html),
  [long-running fits](mfsusie_long_running_fits.html).
