---
title: "Joint multi-outcome fine-mapping"
author: "Gao Wang and contributors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Joint multi-outcome fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 3, fig.align = "center", dpi = 120
)
```

`mfsusie()` extends `fsusie()` from one response to many. Each
outcome `m` is a matrix `n x T_m`; outcomes can be functional
(`T_m > 1`) or scalar (`T_m = 1`), and `T_m` may differ across
outcomes. The fit returns one credible set list and one PIP vector
per variable combining evidence from every outcome.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Three-outcome fixture

Two functional outcomes (e.g., ChIP-seq and ATAC-seq tracks at
different resolutions) plus three scalar phenotypes (e.g., gene
expression in three tissues). Two causal variables, both shared across all
five outcomes; mfsusie should combine the evidence into tight
credible sets.

```{r simulate}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(120)]
n <- nrow(X); p <- ncol(X)

T_func <- c(32L, 64L)
n_scalar <- 3L

# Causal beta: same two variables, different per-outcome strength.
shape_low  <- exp(-((seq_len(T_func[1]) - T_func[1] / 2)^2) / (2 * 4^2))
shape_high <- exp(-((seq_len(T_func[2]) - T_func[2] / 2)^2) / (2 * 6^2))
beta_func1 <- matrix(0, p, T_func[1]); beta_func1[37, ] <- 1.0 * shape_low
                                       beta_func1[88, ] <- -0.7 * shape_low
beta_func2 <- matrix(0, p, T_func[2]); beta_func2[37, ] <- 0.8 * shape_high
                                       beta_func2[88, ] <- -0.5 * shape_high
beta_scalar <- matrix(0, p, n_scalar)
beta_scalar[37, ] <- c(0.9, 0.8, 0.6)
beta_scalar[88, ] <- c(-0.5, -0.6, -0.4)

Y_func1 <- X %*% beta_func1 + matrix(rnorm(n * T_func[1], sd = 0.3), n)
Y_func2 <- X %*% beta_func2 + matrix(rnorm(n * T_func[2], sd = 0.3), n)
Y_scalar <- X %*% beta_scalar + matrix(rnorm(n * n_scalar, sd = 0.3), n)

# Pack scalar outcomes one at a time so each is a column matrix.
Y_list <- list(
  Y_func1,
  Y_func2,
  matrix(Y_scalar[, 1], ncol = 1),
  matrix(Y_scalar[, 2], ncol = 1),
  matrix(Y_scalar[, 3], ncol = 1)
)
length(Y_list)
sapply(Y_list, ncol)
```

## Fit

```{r fit}
fit <- mfsusie(X, Y_list, L = 5,
               prior_variance_scope = "per_outcome",
               verbose = FALSE)
fit$niter; fit$converged
length(fit$sets$cs)
```

## Inspecting the fit

`mfsusie_plot()` shows everything in one call: PIPs in the
top-left tile (CS-colored), then one effect-curve panel per
outcome. Functional outcomes show position-space effect curves
and credible bands; scalar outcomes (`T_m = 1`) show a per-CS dot
plot. The legend on each panel labels CSs by index.

```{r mfsusie-plot, fig.width = 8, fig.height = 7, fig.cap = "Joint fit across two functional outcomes (T = 32, T = 64) and three scalar outcomes. PIP panel (top-left) combines evidence from all five outcomes."}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

```{r cs}
fit$sets$cs
```

`coef(fit)` returns a list of length `M`. Element `m` is an `L x T_m`
matrix containing the per-effect curve for outcome `m`.

```{r coef}
B_hat <- coef(fit)
sapply(B_hat, dim)
```

## Choosing `prior_variance_scope`

- `"per_outcome"` (used above): one mixture-of-normals prior shared
  across every wavelet scale of a outcome. Faster; appropriate when
  the scales are exchangeable a priori.
- `"per_scale"` (default): one mixture per (outcome, scale),
  letting different scales adapt independently. More flexible at the
  cost of a larger M-step.

Switch between them by changing one argument; the rest of the API and
the fit object structure are identical.

## When to reach for mfsusie over fsusie

Use `mfsusie()` when:

- A variable plausibly affects multiple distinct phenotypes (different
  tissues, different molecular layers).
- The phenotypes have heterogeneous shapes (some scalar, some
  functional, some at different sampling resolutions). mfsusie does
  not require a shared `T`.
- You want a single credible set list combining all the evidence,
  rather than ad-hoc colocalization across separate fsusie fits.

Use `fsusie()` when there is exactly one response (functional or
scalar) and no reason to model joint effects across outcomes.
