---
title: "Two-pass scans with warm-start follow-up"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Two-pass scans with warm-start follow-up}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

Joint multi-outcome fits over many ATAC-seq conditions, dense
position grids, or large `p` (number of predictors) can run beyond
the wall time of a typical cluster job. Genome-wide scans
compound the problem: most loci carry no fine-mapping signal,
yet a uniform iteration budget spends the same effort on them as
on the loci that actually need it.

The pattern this vignette covers is a two-pass scan:

1. **First pass**: run every locus with a low `max_iter` budget
   (e.g. 5). Two outcomes are useful directly. Loci that
   converge with `converged = TRUE` are done. Loci that did not
   converge but whose first-pass PIPs are flat (no SNP near 1,
   no narrow credible set) are unlikely to produce signal at any
   budget and can also be dropped. Filtering on these criteria
   alone usually drops the bulk of a genome-wide scan.
2. **Second pass**: re-fit the remaining candidates (didn't
   converge, but first-pass PIPs already concentrate on a small
   number of SNPs) with a larger budget, warm-started from the
   partial fit so the IBSS loop does not restart from cold and
   just polishes the partial state.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Setup: one signal locus and one null locus

```{r data}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(120)]
n <- nrow(X); p <- ncol(X)
T_func <- c(32L, 32L)

# Locus A: real cis-signal at SNPs 37 and 88.
beta_signal <- numeric(p); beta_signal[c(37L, 88L)] <- c(1, -0.6)
Y_signal <- lapply(T_func, function(T_m) {
  X %*% matrix(rep(beta_signal, T_m), nrow = p) +
    matrix(rnorm(n * T_m, sd = 0.4), nrow = n)
})

# Locus B: no signal (independent noise).
Y_null <- lapply(T_func, function(T_m)
  matrix(rnorm(n * T_m, sd = 0.4), nrow = n))
```

## First pass: short budget, screen everything

`max_iter = 5` is enough for null loci to settle and for signal
loci to expose themselves through `converged = FALSE` plus an
ELBO that is still rising. We use `track_fit = TRUE` only to
inspect the trajectory; in production it is optional.

```{r first-pass}
fit_signal_short <- mfsusie(X, Y_signal, max_iter = 5)
fit_null_short   <- mfsusie(X, Y_null,   max_iter = 5)

c(signal = fit_signal_short$niter,    null = fit_null_short$niter)
c(signal = fit_signal_short$converged, null = fit_null_short$converged)
length(fit_signal_short$sets$cs)
length(fit_null_short$sets$cs)
```

The null locus converges in a couple of iterations and produces
no credible sets. The signal locus reports `converged = FALSE`
at the budget — its ELBO is still rising — and is the candidate
to follow up on.

## Triage: drop converged-no-signal, queue non-converged-with-signal

```{r triage}
loci <- list(signal = fit_signal_short, null = fit_null_short)
followup <- vapply(loci, function(f) {
  !isTRUE(f$converged) || length(f$sets$cs) > 0L
}, logical(1L))
followup
```

Only the signal locus is queued for the second pass. A
production scan would loop this triage over all loci and emit
the followup queue.

## Second pass: warm-start from the partial fit

`model_init` accepts a previously returned mfsusie fit. The IBSS
loop seeds `alpha`, `mu`, `mu2`, `V`, `pi_V`, `G_prior`,
`sigma2`, and `fitted` from the supplied object instead of the
cold zero state. The follow-up then converges in a small number
of additional iterations rather than the cold-start total.

```{r warm-followup}
fit_signal_warm <- mfsusie(X, Y_signal,
                           max_iter = 100,
                           model_init = fit_signal_short,
                           verbose = FALSE)
fit_signal_warm$niter
fit_signal_warm$converged
```

Compare to a cold-start run of the same locus at full budget:

```{r cold-baseline}
fit_signal_cold <- mfsusie(X, Y_signal,
                           max_iter = 100,
                           verbose = FALSE)
fit_signal_cold$niter
```

The cold run converges at iteration `fit_signal_cold$niter`; the
warm follow-up adds only the increment between the partial budget
and convergence, so wall time scales with the residual gap rather
than with the full budget.

## Save and restore

Save the fit object between jobs. Re-running mfsusie with the
identical `(X, Y, L, prior, seed)` reproduces the same trajectory
deterministically.

```{r save-restore, eval = FALSE}
saveRDS(fit_signal_short, "fit_partial.rds")
# ...later, on a new job...
fit_partial <- readRDS("fit_partial.rds")
fit_full    <- mfsusie(X, Y_signal,
                       max_iter = 100,
                       model_init = fit_partial)
```

The supplied `model_init` must have the same `L` as the new call,
or fewer (in which case the warm-loaded effects are expanded with
zero-state slots up to the requested `L`, the same growth pattern
used by `L_greedy` between rounds).

## Verify the warm-start matches the cold-start fit

Two checks: the fitted posteriors should agree, and the warm
trajectory should land on the same ELBO as the cold trajectory.

```{r warm-cold-equality}
# Posterior agreement: PIPs, credible sets, final ELBO.
max_pip_diff <- max(abs(fit_signal_warm$pip - fit_signal_cold$pip))
elbo_diff    <- abs(tail(fit_signal_warm$elbo, 1L) -
                    tail(fit_signal_cold$elbo, 1L))
cs_warm <- vapply(fit_signal_warm$sets$cs,
                  function(idx) idx[which.max(fit_signal_warm$pip[idx])],
                  integer(1L))
cs_cold <- vapply(fit_signal_cold$sets$cs,
                  function(idx) idx[which.max(fit_signal_cold$pip[idx])],
                  integer(1L))
data.frame(
  max_pip_diff   = max_pip_diff,
  elbo_diff      = elbo_diff,
  cs_warm_leads  = paste(sort(cs_warm), collapse = ","),
  cs_cold_leads  = paste(sort(cs_cold), collapse = ",")
)
```

The PIPs match within numerical noise, the final ELBOs agree,
and the credible-set lead variants are identical. The
warm-started run reaches the same fixed point.

```{r elbo-trajectory, fig.height = 4}
# Overlay the ELBO trajectories. The warm trajectory continues
# from where the partial fit left off (iteration `n_partial + 1`).
n_partial <- fit_signal_short$niter
elbo_cold <- fit_signal_cold$elbo
elbo_warm <- c(fit_signal_short$elbo, fit_signal_warm$elbo[-1L])

xrng <- c(1, max(length(elbo_cold), length(elbo_warm)))
yrng <- range(c(elbo_cold, elbo_warm), na.rm = TRUE)
plot(seq_along(elbo_cold), elbo_cold, type = "b", pch = 19L,
     col = "grey30", lwd = 2, xlim = xrng, ylim = yrng,
     xlab = "iteration", ylab = "ELBO",
     main = "Cold-start vs warm-start trajectories",
     cex.main = 1.05, font.main = 2L)
lines(seq_along(elbo_warm), elbo_warm, type = "b", pch = 17L,
      col = "firebrick", lwd = 2)
abline(v = n_partial + 0.5, lty = 3, col = "grey50")
legend("bottomright",
       legend = c("cold start (full run)",
                  paste0("warm continuation (partial+",
                         fit_signal_warm$niter, " iters)"),
                  "partial-fit boundary"),
       col = c("grey30", "firebrick", "grey50"),
       lwd = c(2, 2, 1), lty = c(1, 1, 3),
       pch = c(19, 17, NA), bty = "n", cex = 0.8)
```

Both trajectories converge to the same ELBO. The dashed line
marks where the first-pass partial fit ended; the warm
continuation needs only a small increment past that point.

## Diagnostic tracing inside a long run

For loci that resist convergence, `track_fit = TRUE` populates
`fit$trace` with the compact `susie_track` object that susieR
0.16.1+ emits. It is a single list of stacked tables:
`alpha` (long-format: one row per iteration / effect / variable
with `alpha >= 1/p`, thresholded for compactness), `effect`
(per-(iteration, effect) summaries including `V`,
`top_variable`, `top_alpha`, `effect_lbf`), and `iteration`
(per-iteration scalars: `sigma2`, `max_alpha`, `n_effects_kept`,
...). Iteration is 0-indexed.

```{r track-fit, fig.height = 4.5, fig.width = 8}
fit_traced <- mfsusie(X, Y_signal,
                      max_iter = 20, tol = 1e-3,
                      verbose = FALSE, track_fit = TRUE)
nrow(fit_traced$trace$iteration)
str(fit_traced$trace, max.level = 1L)

# How quickly does the per-iteration posterior at each true
# causal SNP settle? `trace$alpha` stores only rows with
# alpha >= 1/p; below-threshold variables read as 0.
true_snps <- c(37L, 88L)
alpha_df  <- fit_traced$trace$alpha
n_iter    <- nrow(fit_traced$trace$iteration)
alpha_path <- vapply(seq(0L, n_iter - 1L), function(it) {
  rows <- alpha_df[alpha_df$iteration == it, , drop = FALSE]
  vapply(true_snps, function(j) {
    vals <- rows$alpha[rows$variable == j]
    if (length(vals) == 0L) 0 else max(vals)
  }, numeric(1L))
}, numeric(2L))
rownames(alpha_path) <- paste0("SNP ", true_snps)

par(mfrow = c(1L, 2L), mar = c(4, 4, 2.5, 1))
plot(seq_len(ncol(alpha_path)), alpha_path["SNP 37", ],
     type = "b", pch = 19L, col = "#1f78b4", lwd = 2,
     ylim = c(0, 1),
     xlab = "iteration",
     ylab = "max alpha at SNP 37 across effects",
     main = "Posterior trajectory at SNP 37",
     cex.main = 1.05, font.main = 2L)
abline(h = 1, lty = 3, col = "grey50")

plot(seq_len(ncol(alpha_path)), alpha_path["SNP 88", ],
     type = "b", pch = 19L, col = "#33a02c", lwd = 2,
     ylim = c(0, 1),
     xlab = "iteration",
     ylab = "max alpha at SNP 88 across effects",
     main = "Posterior trajectory at SNP 88",
     cex.main = 1.05, font.main = 2L)
abline(h = 1, lty = 3, col = "grey50")
par(mfrow = c(1L, 1L))
```

`fit$trace` makes it possible to see *which* effect lands on
*which* SNP at each iteration. When the IBSS loop is slow to
converge, looking at the alpha trajectories at candidate SNPs
exposes whether the loop is oscillating between competing
configurations or simply tightening posterior mass on a
single one. `track_fit = TRUE` adds compact per-iteration
records via `susieR::make_track_snapshot()` (alpha entries are
thresholded at `1/p`); reserve it for diagnostic runs rather
than the genome-wide screen pass.
## Session info

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
