---
title: "Joint multi-outcome fine-mapping"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Joint multi-outcome fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 3, fig.align = "center", dpi = 120
)
```

`mfsusie()` extends `fsusie()` from one functional response to
many. Each outcome `m` is a matrix `n x T_m`; outcomes are
typically functional (DNAm tracks, RNA-seq coverage, ATAC-seq
profiles) and `T_m` may differ across outcomes. The fit returns
one credible set list and one PIP vector per variable, combining
evidence from every outcome. Scalar outcomes (`T_m = 1`) are
supported through the same interface but `susieR::susie()` is
typically the better tool for that case; the strength of
`mfsusie()` is joint fine-mapping across multiple functional
layers.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Bundled multi-omic fixture

`multiomic_example` is a four-outcome cis-QTL fixture, of which
we use the two functional outcomes here (DNAm with T = 64
irregular CpGs and RNA-seq with T = 32 exon-body positions). Two
causal SNPs are shared across both outcomes with per-outcome
shapes and signs. Genotypes use the `susieR::N3finemapping` LD
scaffold for realistic correlation.

```{r load-data}
data(multiomic_example)
Y_list <- multiomic_example$Y_list[c("dnam", "rna")]
pos_list <- multiomic_example$pos_list[c("dnam", "rna")]
sapply(Y_list, ncol)     # T_m per outcome
multiomic_example$causal_snps
```

## Fit

```{r fit}
fit <- mfsusie(multiomic_example$X, Y_list, pos = pos_list,
               L = 15, L_greedy = 5,
               prior_variance_scope = "per_outcome",
               verbose = TRUE)
fit$niter; fit$converged
length(fit$sets$cs)
fit$pip[multiomic_example$causal_snps]
```

## Inspecting the fit

`mfsusie_plot()` shows everything in one call: PIPs in the
top-left tile (CS-colored), then one effect-curve panel per
outcome with TI-smoothed curves, 95% credible bands, and
affected-region bars at the bottom of each panel.

```{r mfsusie-plot, fig.width = 8, fig.height = 5, fig.cap = "Joint fit across DNAm (T = 64) and RNA-seq (T = 32). PIP panel combines evidence from both outcomes; effect panels show TI-smoothed curves with 95% credible bands."}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

```{r cs}
fit$sets$cs
```

`coef(fit)` returns a list of length `M`. Element `m` is an `L x T_m`
matrix containing the per-effect curve for outcome `m`.

```{r coef}
B_hat <- coef(fit)
sapply(B_hat, dim)
```

## Choosing `prior_variance_scope`

- `"per_outcome"` (used above): one mixture-of-normals prior shared
  across every wavelet scale of a outcome. Faster; appropriate when
  the scales are exchangeable a priori.
- `"per_scale"` (default): one mixture per (outcome, scale),
  letting different scales adapt independently. More flexible at the
  cost of a larger M-step.

Switch between them by changing one argument; the rest of the API and
the fit object structure are identical.

## When to reach for mfsusie over fsusie

Use `mfsusie()` when:

- A variable plausibly affects multiple distinct phenotypes (different
  tissues, different molecular layers).
- The phenotypes have heterogeneous shapes (some scalar, some
  functional, some at different sampling resolutions). mfsusie does
  not require a shared `T`.
- You want a single credible set list combining all the evidence,
  rather than ad-hoc colocalization across separate fsusie fits.

Use `fsusie()` when there is exactly one response (functional or
scalar) and no reason to model joint effects across outcomes.
