---
title: "Joint multi-outcome fine-mapping"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Joint multi-outcome fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)

# Colour-rendering pipeline for `summary()` output.
#
# `crayon::has_color()` returns FALSE under knitr, so `summary(configs)`
# would emit plain text and lose its visual encoding (green = active,
# yellow = ambiguous, etc.). Force-enable crayon so the summary always
# emits ANSI SGR escapes, then wire up `fansi::set_knit_hooks` (when
# available) to convert those escapes into HTML <span> styling. The
# fallback path (no fansi) keeps the plain-text output -- still readable,
# just monochrome.
if (requireNamespace("fansi", quietly = TRUE)) {
  options(crayon.enabled = TRUE)
  fansi::set_knit_hooks(knitr::knit_hooks)
}
```

`mfsusie()` extends `fsusie()` from one functional response to
many. Each outcome `m` is a matrix `n x T_m`; outcomes are
typically functional (DNAm tracks, RNA-seq coverage, ATAC-seq
profiles) and `T_m` may differ across outcomes. The fit returns
one credible set list and one PIP vector per variable, combining
evidence from every outcome. Scalar outcomes (`T_m = 1`) are
supported through the same interface but `susieR::susie()` is
typically the better tool for that case; the strength of
`mfsusie()` is joint fine-mapping across multiple functional
layers.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Multi-outcome example

`multiomic_example` is a four-outcome simulated cis-QTL example.
We use the two functional outcomes here: DNAm with `T = 64`
irregular CpGs and RNA-seq with `T = 32` exon-body positions.
Two causal SNPs are shared across both outcomes; per-outcome
shapes and signs differ. Genotypes use the
`susieR::N3finemapping` LD scaffold for realistic correlation.

```{r load-data}
data(multiomic_example)
Y_list   <- multiomic_example$Y_list[c("DNAm", "RNA-seq")]
pos_list <- multiomic_example$pos_list[c("DNAm", "RNA-seq")]
sapply(Y_list, ncol)     # T_m per outcome
multiomic_example$causal_snps
```

## Fit

```{r fit}
fit <- mfsusie(multiomic_example$X, Y_list, pos = pos_list)
fit$niter; fit$converged
length(fit$sets$cs)
fit$pip[multiomic_example$causal_snps]
```

# Post-hoc per-outcome causal-configuration assessment

For each credible set, ask which outcomes share the causal
variant. `susie_post_outcome_configuration(fit, by = "outcome",
method = ...)` runs one of two complementary post-hoc analyses
on the joint mfsusie fit; pick the one you want via `method` and
call again to get the other:

- `method = "susiex"` (default) -- combinatorial enumeration:
  posterior probability of each of the `2^M` binary
  "which-outcomes-share-the-causal" patterns, plus a per-outcome
  marginal probability of being active. One list element per CS
  index shared across all outcomes (so for `M` outcomes that
  share `K` CSes, `length(configs$susiex) == K`).
- `method = "coloc_pairwise"` -- standard H0/H1/H2/H3/H4 coloc
  posteriors for every outcome pair x CS pair (one data.frame).

`summary()` tidies either result into a colour-coded table:
green for active / shared, yellow for ambiguous, dim for
below-threshold. By default it filters to rows that show signal
(pass `signal_only = FALSE` to see everything).

```{r post-config-susiex}
configs <- susieR::susie_post_outcome_configuration(fit, by = "outcome",
                                                    method = "susiex")
summary(configs)
```

For the `coloc_pairwise` analysis the table has one row per
(outcome pair) x (CS pair), which on a 4-outcome fit with 3 shared
CSes means `choose(4, 2) * 3 * 3 = 54` rows. We pass
`signal_only = TRUE` (the default) so H0-dominant rows drop out
and we see only the pairs that carry colocalisation evidence:

```{r post-config-coloc}
pairwise <- susieR::susie_post_outcome_configuration(fit, by = "outcome",
                                                    method = "coloc_pairwise")
summary(pairwise)
```

Index the raw objects directly when you need full per-CS detail:

```{r post-config-raw}
configs$susiex[[1L]]$marginal_prob # per-outcome marginal P(active)
configs$susiex[[1L]]$config_prob   # 2^M configuration probabilities
head(pairwise$coloc_pairwise)      # raw coloc data.frame
```

## Inspecting the fit

`mfsusie_plot()` shows everything in one call: PIPs in the
top-left tile (CS-colored), then one effect-curve panel per
outcome with TI-smoothed curves, 95% credible bands, and
affected-region bars at the bottom of each panel.

```{r mfsusie-smooth}
fit_s <- mf_post_smooth(fit)
```

When the fit has both `M > 1` outcomes and several credible
sets, `facet_cs = "stack"` lays out one effect panel per
`(outcome, CS)` pair stacked vertically below the PIP panel,
which keeps the per-CS curves at full width. The default
`facet_cs = "auto"` falls back to a per-outcome overlay grid
when `M > 1`; switch to `"stack"` when you want a separated
view per CS.

`mfsusie_plot_dimensions()` returns a `(width, height)`
recommendation in inches sized to the number of panels the
fit will produce. Pass it through to the chunk so each cell
keeps a legible amount of vertical space.

```{r mfsusie-plot-dims}
plot_dims <- mfsusie_plot_dimensions(fit_s, facet_cs = "stack")
plot_dims
```

```{r mfsusie-plot, fig.width = plot_dims$width, fig.height = plot_dims$height}
mfsusie_plot(fit_s, facet_cs = "stack",
             effect_variables = multiomic_example$causal_snps)
```

```{r cs}
fit$sets$cs
```

`coef(fit)` returns a list of length `M`. Element `m` is an `L x T_m`
matrix containing the per-effect curve for outcome `m`.

```{r coef}
B_hat <- coef(fit)
sapply(B_hat, dim)
```

## Choosing `prior_variance_scope`

- `"per_outcome"` (default; used above): one mixture-of-normals
  prior shared across every wavelet scale of an outcome.
  Appropriate when scales are exchangeable a priori, and roughly
  `S_m`-fold cheaper per IBSS iter (one mixsqp call per outcome
  instead of one per (outcome, scale)).
- `"per_scale"`: one mixture per (outcome, scale), letting
  different scales adapt independently. More flexible at the cost
  of an `S_m`-fold larger M-step; useful when shape varies
  systematically with scale.
- `"per_scale_normal"`: a 2-parameter point-normal
  `(pi_0, sigma)` per (outcome, scale), fit in
  `mf_em_point_normal()` by direct MLE on the lead variant
  under the current `alpha[l, ]`. Skips the `mixsqp` call and
  tends to be faster on multi-outcome workloads.

Switch between them by changing one argument; the rest of the API and
the fit object structure are identical.

## Wavelet-domain preprocessing

Two arguments control preprocessing of the per-outcome wavelet
matrix before the IBSS loop:

- `wavelet_magnitude_cutoff` (default `0`): wavelet columns with
  `median(|column|) <= wavelet_magnitude_cutoff` are masked, contributing
  `(Bhat = 0, Shat = 1)` to the SER step. Use a small positive
  value when the response has sparse non-zero entries (e.g.,
  ATAC-seq read pileups).
- `wavelet_qnorm` (default `TRUE`): applies a column-wise
  rank-based normal quantile transform to the wavelet matrix.
  Useful when the per-outcome residuals are heavy-tailed or
  visibly non-Gaussian. Set to `FALSE` to skip the transform when
  the wavelet coefficients are already approximately Gaussian.

Both flags route through `mf_adjust_for_covariates()` as well.

## When to reach for mfSuSiE over fSuSiE+COLOC

Use `mfsusie()` when:

- A variable plausibly affects multiple distinct phenotypes (different
  tissues, different molecular layers).
- The phenotypes have heterogeneous shapes (some scalar, some
  functional, some at different sampling resolutions). mfSuSiE does
  not require a shared `T`.
- You want a single credible set list combining all the evidence,
  rather than ad-hoc colocalization across separate fsusie fits.
## Multi-trait data with missing samples

In multi-tissue or multi-cell-type studies some samples are not profiled
for every outcome. Set the corresponding rows of `Y[[m]]` to `NA` before
calling `mfsusie()`; the package automatically restricts every
regression and variance update to the observed rows for that trait.

Both `wavelet_qnorm = TRUE` (the default) and `wavelet_qnorm = FALSE`
handle missing rows correctly.

```{r missing-data-setup}
set.seed(42)
n <- 80; p <- 60; T_m <- 8; M <- 3
X_miss <- matrix(rbinom(n * p, 2, 0.3), n, p)
X_miss <- scale(X_miss)

# Plant one causal SNP shared across all three traits.
causal_j <- 5L
signal <- as.numeric(X_miss[, causal_j]) * 2
Y_miss <- lapply(seq_len(M), function(m)
  matrix(signal, n, T_m) + matrix(rnorm(n * T_m, sd = 0.5), n, T_m))

# Trait 2 is missing 20 % of rows; trait 3 is missing 30 %.
Y_miss[[2]][sample(n, 16), ] <- NA
Y_miss[[3]][sample(n, 24), ] <- NA

# Per-trait observed sample sizes after masking.
sapply(Y_miss, function(y) sum(complete.cases(y)))
```

### `wavelet_qnorm = FALSE`

```{r missing-noqnorm}
t_noq <- system.time(
  fit_noq <- mfsusie(X_miss, Y_miss,
                     wavelet_qnorm = FALSE)
)
cat(sprintf("Runtime: %.1f s   niter: %d   CS: %d\n",
            t_noq["elapsed"], fit_noq$niter, length(fit_noq$sets$cs)))
fit_noq$pip[causal_j]                    # causal PIP
```

### `wavelet_qnorm = TRUE` (default)

```{r missing-qnorm}
t_q <- system.time(
  fit_q <- mfsusie(X_miss, Y_miss)
)
cat(sprintf("Runtime: %.1f s   niter: %d   CS: %d\n",
            t_q["elapsed"], fit_q$niter, length(fit_q$sets$cs)))
fit_q$pip[causal_j]
```

### Runtime summary

```{r missing-runtime-table}
data.frame(
  path    = c("qnorm=FALSE", "qnorm=TRUE"),
  runtime = c(t_noq["elapsed"], t_q["elapsed"]),
  niter   = c(fit_noq$niter,    fit_q$niter),
  n_cs    = c(length(fit_noq$sets$cs), length(fit_q$sets$cs)),
  causal_pip = c(fit_noq$pip[causal_j], fit_q$pip[causal_j])
)
```

## Session info

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
