---
title: "Practical patterns for long-running fits"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Practical patterns for long-running fits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

Joint multi-outcome fits over many ATAC-seq conditions, dense
position grids, or large `p` (number of predictors) can run beyond
the wall time of a typical cluster job. This vignette covers
practical patterns for handling those cases.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Diagnose convergence before increasing the budget

The fit object records the per-iteration ELBO trajectory, the
iteration count, and a converged flag. If the ELBO is still rising
when `niter == max_iter`, the run terminated by budget.

```{r run-short}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(120)]
n <- nrow(X); p <- ncol(X)
T_func <- c(32L, 32L)
beta1 <- numeric(p); beta1[c(37, 88)] <- c(1, -0.6)

Y_list <- lapply(T_func, function(T_m) {
  X %*% matrix(rep(beta1, T_m), nrow = p) +
    matrix(rnorm(n * T_m, sd = 0.4), nrow = n)
})

fit_short <- mfsusie(X, Y_list, L = 15, L_greedy = 5, max_iter = 5,
                     prior_variance_scope = "per_outcome",
                     verbose = TRUE)
fit_short$niter
fit_short$converged
tail(fit_short$elbo, 5)
```

If `converged` is `FALSE` and the ELBO is still climbing in the last
few iterations, the fix is more iterations.

```{r run-full}
fit_full <- mfsusie(X, Y_list, L = 15, L_greedy = 5, max_iter = 100,
                    prior_variance_scope = "per_outcome",
                    verbose = TRUE)
fit_full$niter
fit_full$converged
```

## Save and restore an in-progress fit

Save the fit object after each cluster job. Re-running with the
identical seed and full iteration budget reproduces the same
trajectory deterministically; mfsusie's IBSS dispatch is purely
deterministic given the same `(X, Y, L, prior, seed)`.

```{r save-restore, eval = FALSE}
saveRDS(fit_short, "fit_progress.rds")
# ...later, on a new job...
fit_short <- readRDS("fit_progress.rds")
```

## Mid-run restart from a saved partial fit

`mfsusie()` accepts a `model_init` argument carrying a previously
returned fit. The IBSS loop seeds itself from the supplied
`alpha`, `mu`, `mu2`, `V`, `pi_V`, `G_prior`, `sigma2`, and
`fitted` rather than the cold-start zero state, so a job that
hit a per-call wall-time cap can resume from where it left off.

```{r warm-restart, eval = FALSE}
# First job: cap iterations and save the partial fit.
fit_partial <- mfsusie(X, Y_list, L = 15, L_greedy = 5,
                       max_iter = 4,
                       prior_variance_scope = "per_outcome",
                       verbose = TRUE)
saveRDS(fit_partial, "fit_progress.rds")

# Later, on a new job: resume.
fit_partial <- readRDS("fit_progress.rds")
fit_full    <- mfsusie(X, Y_list, L = 15, L_greedy = 5,
                       max_iter = 100,
                       model_init = fit_partial,
                       prior_variance_scope = "per_outcome",
                       verbose = TRUE)
```

The resumed run typically converges in one or two additional
iterations when the partial fit was already close to the fixed
point. The supplied `model_init` must have the same `L` as the
new call, or fewer (in which case the warm-loaded effects are
expanded with zero-state slots up to the requested `L`, the
same growth pattern used by `L_greedy` between rounds).

For diagnostic monitoring inside a long run, consider:

```{r monitor, eval = FALSE}
fit <- mfsusie(X, Y_list, L = 15, L_greedy = 5,
               max_iter = 100, tol = 1e-3,
               prior_variance_scope = "per_outcome",
               verbose = TRUE,                # logs per-iter ELBO
               track_fit = TRUE)              # records full trajectory
```

`track_fit = TRUE` keeps the per-iteration model snapshots inside
`fit$trace`, which is useful for debugging slow convergence at the
cost of memory.

## When the bottleneck is `T_basis` rather than iterations

A fit on a single outcome with `T_m = 4096` can be slow at the
wavelet step regardless of iteration count. Two levers:

- `max_padded_log2` caps the wavelet grid size. The default is `10`
  (i.e., `T_basis <= 1024`). Drop it for a sanity check.
- Drop `coverage` from `0.95` to a coarser `0.9` to reduce the size
  of credible-set search.

For very large `T_m`, post-hoc smoothing helpers (smash, TI, HMM)
are exposed through a separate `mf_post_smooth()` API rather than
folded into the IBSS loop. The core fit returns the wavelet-space
posterior; the smoothing module is a follow-up PR.
