---
title: "Joint multi-outcome fine-mapping"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Joint multi-outcome fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

`mfsusie()` extends `fsusie()` from one functional response to
many. Each outcome `m` is a matrix `n x T_m`; outcomes are
typically functional (DNAm tracks, RNA-seq coverage, ATAC-seq
profiles) and `T_m` may differ across outcomes. The fit returns
one credible set list and one PIP vector per variable, combining
evidence from every outcome. Scalar outcomes (`T_m = 1`) are
supported through the same interface but `susieR::susie()` is
typically the better tool for that case; the strength of
`mfsusie()` is joint fine-mapping across multiple functional
layers.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Multi-outcome example

`multiomic_example` is a four-outcome simulated cis-QTL example.
We use the two functional outcomes here: DNAm with `T = 64`
irregular CpGs and RNA-seq with `T = 32` exon-body positions.
Two causal SNPs are shared across both outcomes; per-outcome
shapes and signs differ. Genotypes use the
`susieR::N3finemapping` LD scaffold for realistic correlation.

```{r load-data}
data(multiomic_example)
Y_list   <- multiomic_example$Y_list[c("DNAm", "RNA-seq")]
pos_list <- multiomic_example$pos_list[c("DNAm", "RNA-seq")]
sapply(Y_list, ncol)     # T_m per outcome
multiomic_example$causal_snps
```

## Fit

```{r fit}
fit <- mfsusie(multiomic_example$X, Y_list, pos = pos_list,
               L = 15, L_greedy = 5,
               prior_variance_scope = "per_outcome",
               verbose = TRUE)
fit$niter; fit$converged
length(fit$sets$cs)
fit$pip[multiomic_example$causal_snps]
```

## Inspecting the fit

`mfsusie_plot()` shows everything in one call: PIPs in the
top-left tile (CS-colored), then one effect-curve panel per
outcome with TI-smoothed curves, 95% credible bands, and
affected-region bars at the bottom of each panel.

```{r mfsusie-smooth}
fit_s <- mf_post_smooth(fit, method = "TI")
```

When the fit has both `M > 1` outcomes and several credible
sets, `facet_cs = "stack"` lays out one effect panel per
`(outcome, CS)` pair stacked vertically below the PIP panel,
which keeps the per-CS curves at full width. The default
`facet_cs = "auto"` falls back to a per-outcome overlay grid
when `M > 1`; switch to `"stack"` when you want a separated
view per CS.

`mfsusie_plot_dimensions()` returns a `(width, height)`
recommendation in inches sized to the number of panels the
fit will produce. Pass it through to the chunk so each cell
keeps a legible amount of vertical space.

```{r mfsusie-plot-dims}
plot_dims <- mfsusie_plot_dimensions(fit_s, facet_cs = "stack")
plot_dims
```

```{r mfsusie-plot, fig.width = plot_dims$width, fig.height = plot_dims$height}
mfsusie_plot(fit_s, facet_cs = "stack")
```

```{r cs}
fit$sets$cs
```

`coef(fit)` returns a list of length `M`. Element `m` is an `L x T_m`
matrix containing the per-effect curve for outcome `m`.

```{r coef}
B_hat <- coef(fit)
sapply(B_hat, dim)
```

## Choosing `prior_variance_scope`

- `"per_outcome"` (used above): one mixture-of-normals prior shared
  across every wavelet scale of a outcome. Faster; appropriate when
  the scales are exchangeable a priori.
- `"per_scale"` (default): one mixture per (outcome, scale),
  letting different scales adapt independently. More flexible at the
  cost of a larger M-step.

Switch between them by changing one argument; the rest of the API and
the fit object structure are identical.

## Wavelet-domain preprocessing

Two arguments control preprocessing of the per-outcome wavelet
matrix before the IBSS loop:

- `low_count_filter` (default `0`): wavelet columns with
  `median(|column|) <= low_count_filter` are masked, contributing
  `(Bhat = 0, Shat = 1)` to the SER step. Use a small positive
  value when the response has sparse non-zero entries (e.g.,
  ATAC-seq read pileups).
- `quantile_norm` (default `FALSE`): when `TRUE`, applies a
  column-wise rank-based normal quantile transform to the
  wavelet matrix. Use when the per-outcome residuals are
  heavy-tailed or visibly non-Gaussian.

Both flags route through `mf_adjust_for_covariates()` as well.

## When to reach for mfSuSiE over fSuSiE+COLOC

Use `mfsusie()` when:

- A variable plausibly affects multiple distinct phenotypes (different
  tissues, different molecular layers).
- The phenotypes have heterogeneous shapes (some scalar, some
  functional, some at different sampling resolutions). mfSuSiE does
  not require a shared `T`.
- You want a single credible set list combining all the evidence,
  rather than ad-hoc colocalization across separate fsusie fits.
## Session info

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
