---
title: "Joint multi-outcome fine-mapping"
author: "Gao Wang and contributors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Joint multi-outcome fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 5, fig.height = 3, fig.align = "center", dpi = 120
)
```

`mfsusie()` extends `fsusie()` from one response to many. Each
outcome `m` is a matrix `n x T_m`; outcomes can be functional
(`T_m > 1`) or scalar (`T_m = 1`), and `T_m` may differ across
outcomes. The fit returns one credible set list and one PIP vector
per variable combining evidence from every outcome.

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

## Bundled multi-omic fixture

`multiomic_example` is a four-outcome cis-QTL fixture: DNA
methylation (T = 64, irregular CpGs), RNA-seq (T = 32, exon
body), and two scalar QTLs (eQTL, pQTL). Two causal SNPs are
shared across all four outcomes with per-outcome shapes and
signs. Genotypes use the `susieR::N3finemapping` LD scaffold for
realistic correlation.

```{r load-data}
data(multiomic_example)
str(multiomic_example, max.level = 1)
sapply(multiomic_example$Y_list, ncol)   # T_m per outcome
multiomic_example$causal_snps
```

## Fit

```{r fit}
fit <- mfsusie(multiomic_example$X, multiomic_example$Y_list,
               pos = multiomic_example$pos_list,
               L = 5,
               prior_variance_scope = "per_outcome",
               verbose = FALSE)
fit$niter; fit$converged
length(fit$sets$cs)
fit$pip[multiomic_example$causal_snps]
```

## Inspecting the fit

`mfsusie_plot()` shows everything in one call: PIPs in the
top-left tile (CS-colored), then one effect-curve panel per
outcome. Functional outcomes show position-space effect curves
and credible bands; scalar outcomes (`T_m = 1`) show a per-CS dot
plot. The legend on each panel labels CSs by index.

```{r mfsusie-plot, fig.width = 8, fig.height = 7, fig.cap = "Joint fit across DNAm (T = 64), RNA-seq (T = 32), and two scalar QTLs. PIP panel (top-left) combines evidence from all four outcomes; functional panels show TI-smoothed effect curves with 95% credible bands; scalar panels show per-CS effect points."}
fit_s <- mf_post_smooth(fit, method = "TI",
                        wavelet_filter = 1L,
                        wavelet_family = "DaubExPhase")
mfsusie_plot(fit_s)
```

```{r cs}
fit$sets$cs
```

`coef(fit)` returns a list of length `M`. Element `m` is an `L x T_m`
matrix containing the per-effect curve for outcome `m`.

```{r coef}
B_hat <- coef(fit)
sapply(B_hat, dim)
```

## Choosing `prior_variance_scope`

- `"per_outcome"` (used above): one mixture-of-normals prior shared
  across every wavelet scale of a outcome. Faster; appropriate when
  the scales are exchangeable a priori.
- `"per_scale"` (default): one mixture per (outcome, scale),
  letting different scales adapt independently. More flexible at the
  cost of a larger M-step.

Switch between them by changing one argument; the rest of the API and
the fit object structure are identical.

## When to reach for mfsusie over fsusie

Use `mfsusie()` when:

- A variable plausibly affects multiple distinct phenotypes (different
  tissues, different molecular layers).
- The phenotypes have heterogeneous shapes (some scalar, some
  functional, some at different sampling resolutions). mfsusie does
  not require a shared `T`.
- You want a single credible set list combining all the evidence,
  rather than ad-hoc colocalization across separate fsusie fits.

Use `fsusie()` when there is exactly one response (functional or
scalar) and no reason to model joint effects across outcomes.
