---
title: "Smoothing effect curves and credible bands"
author: "Gao Wang and contributors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Smoothing effect curves and credible bands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, fig.align = "center", dpi = 120
)
```

## Why post-process

`mfsusie()` and `fsusie()` fit the model in the wavelet domain. The
per-effect posterior is summarised by `coef(fit)`, which inverts the
wavelet representation back to position space. The raw inverse curve
is correct but visually wiggly: small fine-scale wavelet coefficients
contribute pointwise noise that obscures the underlying shape, and
the fit returns no pointwise credible band by default.

The post-processor `mf_post_smooth(fit)` applies translation-invariant
wavelet thresholding (TIWT) to the per-effect wavelet posterior mean,
shrinking small fine-scale coefficients to zero, and computes
pointwise credible bands from the wavelet posterior variance. The
resulting fit gains two slots:

- `fit$effect_curves[[m]][[l]]` — smoothed position-space curve for
  outcome `m`, effect `l`.
- `fit$credible_bands[[m]][[l]]` — `T_basis x 2` matrix
  `[lower, upper]`.

`fsusie_plot()` and `mfsusie_plot()` automatically draw the smoothed
curves + bands when those slots are populated; otherwise they fall
back to the raw inverse-DWT curves.

## A small example

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

Simulate one functional outcome with a smooth bell-shaped effect at
each of two causal variables:

```{r simulate}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X); T_m <- 64

positions <- seq_len(T_m)
shape <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 8)^2))
beta <- matrix(0, p, T_m)
beta[42, ] <-  1.5 * shape
beta[97, ] <- -1.0 * shape
Y <- X %*% beta + matrix(rnorm(n * T_m, sd = 0.5), n)

fit <- fsusie(Y, X, pos = positions, L = 3, verbose = FALSE)
```

## Before smoothing

`fsusie_plot()` on the unsmoothed fit shows curves recovered by
`coef()` (raw inverse DWT). They track the true bell-shape but the
fine wavelet scales add visible jitter, and there are no credible
bands.

```{r before, fig.cap = "Raw effect curves (no smoothing). Top: PIPs. Bottom: per-effect inverse-DWT curve."}
fsusie_plot(fit)
```

## After smoothing

```{r smooth}
fit_s <- mf_post_smooth(fit)
```

The same plot call now uses the smoothed curves and adds 95% pointwise
credible bands as transparent ribbons:

```{r after, fig.cap = "After mf_post_smooth(): TIWT-smoothed curves with 95% pointwise credible bands."}
fsusie_plot(fit_s)
```

The shape is preserved, fine-scale jitter is suppressed, and the
ribbon makes it obvious where the curve is significantly different
from zero (band excludes 0).

## Tuning the smoother

`mf_post_smooth()` exposes two knobs:

- `level` — credible-band coverage. Default `0.95`. Drop to `0.90`
  for narrower bands; raise to `0.99` for wider.
- `threshold_factor` — multiplier on the universal SureShrink
  threshold `sigma * sqrt(2 log T)`. `1` is standard. Smaller values
  keep more fine-scale wavelet energy (less smoothing); larger
  values smooth more aggressively.

```{r tuning, fig.cap = "Same fit smoothed with threshold_factor = 0.5 (less smoothing)."}
fit_light <- mf_post_smooth(fit, threshold_factor = 0.5)
fsusie_plot(fit_light)
```

## Multi-outcome usage

`mf_post_smooth()` works on `mfsusie()` fits the same way; the
returned fit has `$effect_curves[[m]]` and `$credible_bands[[m]]`
populated for every functional outcome. Scalar outcomes
(`T_m = 1`) are passed through (no smoothing needed) and get a
band derived from the wavelet posterior SD.

```{r mfsusie}
T_per <- c(64L, 32L)
Y_list <- lapply(T_per, function(T_m) {
  X %*% matrix(rep(beta[, 1] * 0 + 0.6 * shape[seq_len(T_m)], p),
               nrow = p, byrow = TRUE) +
    matrix(rnorm(n * T_m, sd = 0.4), n)
})
# Make beta non-trivial again per outcome.
Y_list[[1]] <- X %*% beta + matrix(rnorm(n * T_m, sd = 0.5), n)
beta2 <- matrix(0, p, 32L)
beta2[42, ] <- 1.0 * exp(-((seq_len(32) - 16)^2) / (2 * 4^2))
Y_list[[2]] <- X %*% beta2 + matrix(rnorm(n * 32L, sd = 0.4), n)

fit_m <- mfsusie(X, Y_list, L = 3, verbose = FALSE)
fit_m_s <- mf_post_smooth(fit_m)
```

```{r mfsusie-plot, fig.cap = "Multi-outcome fit, smoothed. Top-left: PIPs across all outcomes. Other panels: per-outcome smoothed effect curves with 95% credible bands."}
mfsusie_plot(fit_m_s)
```

## Method

The current `method = "tiwt"` performs scalewise soft-thresholding
of the wavelet posterior mean:

1. For each effect `l` and outcome `m`, compute the wavelet-domain
   posterior mean `mean_w = alpha[l, ] %*% mu[[l]][[m]]` and the
   posterior second moment via `mu2`. The pointwise variance is
   `var_w = alpha[l, ] %*% mu2[[l]][[m]] - mean_w^2`.
2. Per wavelet scale `s`, soft-threshold `mean_w[scale_index[[s]]]`
   at `factor * sd_s * sqrt(2 log T)`, where `sd_s` is the mean
   posterior SD on that scale.
3. Invert the thresholded coefficients via `mf_invert_dwt()` to
   obtain the smoothed position-space curve.
4. Compute the pointwise SD by inverting `sqrt(var_w)` (Parseval
   approximation on the orthonormal basis) and form the credible
   band as `mean_pos +/- z(level) * sd_pos`.

This is a v1 smoother. Cycle-spinning translation-invariant
denoising (the full TI estimator from the fSuSiE paper) and
empirical-Bayes wavelet shrinkage (`smashr`) are planned
extensions; both will be exposed via the same
`method = "..."` argument when added.
