---
title: "Smoothing effect curves and credible bands"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Smoothing effect curves and credible bands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

## Why post-process

`mfsusie()` and `fsusie()` fit the model in the wavelet domain. The
per-effect posterior is summarized by `coef(fit)`, which inverts the
wavelet representation back to position space. The raw inverse curve
is correct but visually wiggly: small fine-scale wavelet coefficients
contribute pointwise noise that obscures the underlying shape, and
the fit returns no pointwise credible band by default.

The post-processor `mf_post_smooth(fit, method = ...)` runs the
chosen smoother on the per-effect wavelet posterior mean and
stores the result on the fit at `fit$smoothed[[method]]`. The
slot carries:

- `fit$smoothed[[method]]$effect_curves[[m]][[l]]` — smoothed
  position-space curve for outcome `m`, effect `l`.
- `fit$smoothed[[method]]$credible_bands[[m]][[l]]` —
  `T_basis x 2` matrix `[lower, upper]`.
- `fit$smoothed$HMM$lfsr_curves[[m]][[l]]` — per-position local
  false sign rate (HMM only).

The fit accumulates per-method results: every call to
`mf_post_smooth(fit, method = ...)` adds (or, with
`overwrite_previous = TRUE`, replaces) one entry in
`fit$smoothed`. The raw inverse-DWT curves remain accessible via
`coef(fit)`; pass `coef(fit, smooth_method = "TI")` to get the
TI-smoothed curves on the same `list[M]` of `L x T_basis` matrix
shape.

`mfsusie_plot()` reads from `fit$smoothed` and picks a smoother
in the priority order `TI > smash > HMM > scalewise` when the
user does not name one explicitly. When more than one smoother
has been applied, the plot emits a hint listing the alternatives
so the user can compare. After `method = "HMM"` smoothing the
per-position lfsr curves are also overlaid on a secondary axis
with a threshold line at `lfsr_threshold = 0.05`. Affected
regions (where each CS's credible band excludes zero) are marked
with thick bars at the bottom of each effect panel.

## A small example

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

Simulate one functional outcome with a smooth bell-shaped effect at
each of two causal variables:

```{r simulate}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X); T_m <- 64

positions <- seq_len(T_m)
shape <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 8)^2))
beta <- matrix(0, p, T_m)
beta[42, ] <-  1.5 * shape
beta[97, ] <- -1.0 * shape
Y <- X %*% beta + matrix(rnorm(n * T_m, sd = 0.5), n)

fit <- fsusie(Y, X, pos = positions)
```

## Before smoothing

`mfsusie_plot()` on the unsmoothed fit shows curves recovered from
`coef()` (raw inverse DWT). The curves track the true bell shape,
but fine wavelet scales contribute high-frequency noise, and the
fit returns no credible bands by default.

```{r before-dims}
plot_dims_b <- mfsusie_plot_dimensions(fit)
```

```{r before, fig.width = plot_dims_b$width, fig.height = plot_dims_b$height}
mfsusie_plot(fit)
```

## After smoothing

```{r smooth}
fit_s <- mf_post_smooth(fit)            # default method = "TI"
names(fit_s$smoothed)                   # which smoothings are on the fit
```

The same plot call now reads `fit_s$smoothed$TI` and adds 95%
pointwise credible bands as transparent ribbons:

```{r after-dims}
plot_dims_a <- mfsusie_plot_dimensions(fit_s)
```

```{r after, fig.width = plot_dims_a$width, fig.height = plot_dims_a$height}
mfsusie_plot(fit_s)
```

The shape is preserved, high-frequency noise is suppressed, and
the ribbon marks the positions where the band excludes zero.

## Stacking multiple smoothers on one fit

Every call to `mf_post_smooth(fit, method = ...)` adds an entry
to `fit$smoothed` rather than overwriting the previous result.
Comparing methods is a few lines:

```{r stack}
fit_s <- mf_post_smooth(fit_s, method = "HMM")
fit_s <- mf_post_smooth(fit_s, method = "scalewise")
names(fit_s$smoothed)
```

`coef()` exposes the same multi-method state via the
`smooth_method` argument:

```{r coef-multi}
str(coef(fit_s),                          max.level = 1L)  # raw inverse DWT
str(coef(fit_s, smooth_method = "TI"),    max.level = 1L)
str(coef(fit_s, smooth_method = "HMM"),   max.level = 1L)
```

Because `coef(fit)` defaults to the raw inverse DWT, smoothing
is always an explicit opt-in. Pass `smooth_method = "<name>"`
when you want the post-processed curves.

## Choosing a smoother

`mf_post_smooth(method = ...)` supports four methods:

- `"TI"` (default) — translation-invariant wavelet denoising via
  cycle spinning. Best when the per-position band is the visual
  focus.
- `"scalewise"` — per-scale soft-thresholding of the
  wavelet posterior mean at
  `threshold_factor * sd_s * sqrt(2 log T)`, followed by inverse
  DWT. Cheapest, no extra dependencies. Lowers fine-scale
  jitter while preserving the overall shape.
- `"HMM"` — hidden Markov model on per-position regression
  estimates with `ashr` mixture refinement. Adds a per-position
  local false sign rate (`lfsr_curves`); `mfsusie_plot()`
  overlays it on a secondary axis with the threshold line.
- `"smash"` — `smashr::smash.gaus` empirical-Bayes wavelet
  shrinkage on the per-position regression estimate. Requires
  the `smashr` package (Suggests).

All four smoothers can coexist on the same fit. `mfsusie_plot()`
picks one by priority (TI > smash > HMM > scalewise) and emits a
hint listing the others; pass `smooth_method = "<name>"` to plot
a specific one.

```{r tuning-scalewise-dims}
fit_scalewise <- mf_post_smooth(fit, method = "scalewise",
                                threshold_factor = 0.5)
plot_dims_sc <- mfsusie_plot_dimensions(fit_scalewise,
                                        smooth_method = "scalewise")
```

```{r tuning-scalewise, fig.width = plot_dims_sc$width, fig.height = plot_dims_sc$height}
mfsusie_plot(fit_scalewise, smooth_method = "scalewise")
```

```{r tuning-hmm-dims}
fit_hmm <- mf_post_smooth(fit, method = "HMM")
plot_dims_hm <- mfsusie_plot_dimensions(fit_hmm)
```

```{r tuning-hmm, fig.width = plot_dims_hm$width, fig.height = plot_dims_hm$height}
mfsusie_plot(fit_hmm)   # picks HMM (the only smoother on the fit)
```

```{r tuning-smash-defaults, include = FALSE}
# Fallback values so the next chunk's fig.* expressions resolve
# even when smashr is unavailable (the chunk body is gated by
# `eval`, but knitr always evaluates fig.width / fig.height).
plot_dims_sm <- list(width = 8, height = 6)
```

```{r tuning-smash, eval = requireNamespace("smashr", quietly = TRUE)}
fit_smash <- mf_post_smooth(fit, method = "smash")
plot_dims_sm <- mfsusie_plot_dimensions(fit_smash)
```

```{r tuning-smash-show, eval = requireNamespace("smashr", quietly = TRUE), fig.width = plot_dims_sm$width, fig.height = plot_dims_sm$height}
mfsusie_plot(fit_smash)
```

When several smoothers are stacked, plotting picks the
highest-priority one and lists the others as a hint:

```{r tuning-stacked-dims}
fit_multi <- mf_post_smooth(fit)
fit_multi <- mf_post_smooth(fit_multi, method = "HMM")
fit_multi <- mf_post_smooth(fit_multi, method = "scalewise")
plot_dims_ml <- mfsusie_plot_dimensions(fit_multi)
```

```{r tuning-stacked, fig.width = plot_dims_ml$width, fig.height = plot_dims_ml$height}
mfsusie_plot(fit_multi)                              # picks TI; hint prints
mfsusie_plot(fit_multi, smooth_method = "HMM")       # plot HMM directly
```

## Multi-outcome usage

`mf_post_smooth()` works on `mfsusie()` fits the same way; the
chosen entry of `fit$smoothed[[method]]` carries
`effect_curves[[m]]` and `credible_bands[[m]]` for every
functional outcome. Scalar outcomes (`T_m = 1`) are passed
through (no smoothing needed) and get a band derived from the
wavelet posterior SD.

```{r mfsusie}
# Two functional outcomes at different sampling resolutions
# (T = 64 and T = 32). Same two causal variables, different
# per-outcome bell shapes.
T_per <- c(64L, 32L)
shape64 <- exp(-((seq_len(64) - 32)^2) / (2 * 8^2))
shape32 <- exp(-((seq_len(32) - 16)^2) / (2 * 4^2))
beta1 <- matrix(0, p, 64L); beta1[42, ] <- 1.5 * shape64
                            beta1[97, ] <- -1.0 * shape64
beta2 <- matrix(0, p, 32L); beta2[42, ] <- 1.0 * shape32
                            beta2[97, ] <- -0.6 * shape32
Y_list <- list(
  X %*% beta1 + matrix(rnorm(n * 64, sd = 0.5), n),
  X %*% beta2 + matrix(rnorm(n * 32, sd = 0.4), n)
)

fit_m   <- mfsusie(X, Y_list)
fit_m_s <- mf_post_smooth(fit_m)
```

```{r mfsusie-plot-dims}
plot_dims_ms <- mfsusie_plot_dimensions(fit_m_s)
```

```{r mfsusie-plot, fig.width = plot_dims_ms$width, fig.height = plot_dims_ms$height}
mfsusie_plot(fit_m_s)
```

## Method

Each method takes the per-effect wavelet-domain posterior
`(mean_w, var_w) = (alpha[l, ] %*% mu[[l]][[m]],
alpha[l, ] %*% mu2[[l]][[m]] - mean_w^2)` and produces a smoothed
position-space curve plus a pointwise credible band:

- `"TI"` — cycle spinning: at every shift, run a stationary DWT
  with squared filter coefficients to compute pointwise variance,
  apply ash to the wavelet coefficients per scale, and average
  the inverse reconstructions across shifts. Credible bands come
  from the average-of-basis variance kernel.
- `"scalewise"` — soft-threshold `mean_w[scale_index[[s]]]` per
  wavelet scale `s` at `threshold_factor * sd_s * sqrt(2 log T)`,
  then invert via `mf_invert_dwt()`. The pointwise variance is
  the position-space variance from the wavelet posterior:
  `Var(pos[t]) = sum_k W^T_{tk}^2 * var_w[k]` where `W^T` is
  the inverse-DWT matrix. The band is `mean_pos +/- z(level)
  * sqrt(Var(pos))`.
- `"HMM"` — discrete-state hidden Markov model on per-position
  regression estimates with hub-and-spoke transitions and ash
  refinement at each EM iteration. The smoothed curve is the
  posterior mean. The pointwise SD is computed via the law of
  total variance over the HMM state mixture:
  `Var(pos[t]) = sum_k prob[t, k] * (sd_kt^2 + mean_kt^2)
  - (sum_k prob[t, k] * mean_kt)^2`, where `prob[t, k]` is the
  forward-backward state probability and `(mean_kt, sd_kt)` are
  the per-state ash posterior moments. Band as
  `mean_pos +/- z(level) * sqrt(Var(pos))`.
- `"smash"` — `smashr::smash.gaus` empirical-Bayes wavelet
  shrinkage on the per-position regression estimate of the
  effect against the lead variable. The credible band uses
  `smash`'s `post.var = TRUE` posterior variance.

## Session info

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
