---
title: "Smoothing effect curves and credible bands"
author: "Gao Wang, Anjing Liu and William Denault"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Smoothing effect curves and credible bands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 6, fig.align = "center", dpi = 110,
  out.width = "100%"
)
```

## Why post-process

`mfsusie()` and `fsusie()` fit the model in the wavelet domain. The
per-effect posterior is summarised by `coef(fit)`, which inverts the
wavelet representation back to position space. The raw inverse curve
is correct but visually wiggly: small fine-scale wavelet coefficients
contribute pointwise noise that obscures the underlying shape, and
the fit returns no pointwise credible band by default.

The post-processor `mf_post_smooth(fit)` applies translation-invariant
wavelet thresholding (TIWT) to the per-effect wavelet posterior mean,
shrinking small fine-scale coefficients to zero, and computes
pointwise credible bands from the wavelet posterior variance. The
resulting fit gains two slots:

- `fit$effect_curves[[m]][[l]]` — smoothed position-space curve for
  outcome `m`, effect `l`.
- `fit$credible_bands[[m]][[l]]` — `T_basis x 2` matrix
  `[lower, upper]`.

`mfsusie_plot()` automatically draws the smoothed curves and
credible bands when those slots are populated; otherwise it falls
back to the raw inverse-DWT curves. After `method = "HMM"`
smoothing the per-position lfsr curves are also overlaid on a
secondary axis with a threshold line at `lfsr_threshold = 0.05`.
Affected regions (where each CS's credible band excludes zero)
are marked with thick bars at the bottom of each effect panel.

## A small example

```{r setup, message = FALSE}
library(mfsusieR)
library(susieR)
set.seed(1)
```

Simulate one functional outcome with a smooth bell-shaped effect at
each of two causal variables:

```{r simulate}
data(N3finemapping)
X <- N3finemapping$X[, seq_len(150)]
n <- nrow(X); p <- ncol(X); T_m <- 64

positions <- seq_len(T_m)
shape <- exp(-((positions - T_m / 2)^2) / (2 * (T_m / 8)^2))
beta <- matrix(0, p, T_m)
beta[42, ] <-  1.5 * shape
beta[97, ] <- -1.0 * shape
Y <- X %*% beta + matrix(rnorm(n * T_m, sd = 0.5), n)

fit <- fsusie(Y, X, pos = positions, L = 15, L_greedy = 5, verbose = TRUE)
```

## Before smoothing

`mfsusie_plot()` on the unsmoothed fit shows curves recovered from
`coef()` (raw inverse DWT). The curves track the true bell shape,
but fine wavelet scales contribute high-frequency noise, and the
fit returns no credible bands by default.

```{r before}
mfsusie_plot(fit)
```

## After smoothing

```{r smooth}
fit_s <- mf_post_smooth(fit)
```

The same plot call now uses the smoothed curves and adds 95% pointwise
credible bands as transparent ribbons:

```{r after}
mfsusie_plot(fit_s)
```

The shape is preserved, high-frequency noise is suppressed, and
the ribbon marks the positions where the band excludes zero.

## Choosing a smoother

`mf_post_smooth(method = ...)` supports three methods:

- `"scalewise"` (default) — per-scale soft-thresholding of the
  wavelet posterior mean at `threshold_factor * sd_s * sqrt(2 log T)`,
  followed by inverse DWT. Cheapest, no extra dependencies. Lowers
  fine-scale jitter while preserving the overall shape.
- `"TI"` — translation-invariant wavelet denoising via cycle
  spinning. Bit-identical to `fsusieR::univariate_TI_regression`
  on the orthonormal wavelet path. Best when the per-position
  band is the visual focus.
- `"HMM"` — hidden Markov model on per-position regression
  estimates with `ashr` mixture refinement. Adds a per-position
  local false sign rate (`lfsr_curves`); `mfsusie_plot()` overlays
  it on a secondary axis with the threshold line.

```{r tuning-scalewise}
fit_light <- mf_post_smooth(fit, method = "scalewise",
                            threshold_factor = 0.5)
mfsusie_plot(fit_light)
```

```{r tuning-hmm}
fit_hmm <- mf_post_smooth(fit, method = "HMM")
mfsusie_plot(fit_hmm)
```

## Multi-outcome usage

`mf_post_smooth()` works on `mfsusie()` fits the same way; the
returned fit has `$effect_curves[[m]]` and `$credible_bands[[m]]`
populated for every functional outcome. Scalar outcomes
(`T_m = 1`) are passed through (no smoothing needed) and get a
band derived from the wavelet posterior SD.

```{r mfsusie}
# Two functional outcomes at different sampling resolutions
# (T = 64 and T = 32). Same two causal variables, different
# per-outcome bell shapes.
T_per <- c(64L, 32L)
shape64 <- exp(-((seq_len(64) - 32)^2) / (2 * 8^2))
shape32 <- exp(-((seq_len(32) - 16)^2) / (2 * 4^2))
beta1 <- matrix(0, p, 64L); beta1[42, ] <- 1.5 * shape64
                            beta1[97, ] <- -1.0 * shape64
beta2 <- matrix(0, p, 32L); beta2[42, ] <- 1.0 * shape32
                            beta2[97, ] <- -0.6 * shape32
Y_list <- list(
  X %*% beta1 + matrix(rnorm(n * 64, sd = 0.5), n),
  X %*% beta2 + matrix(rnorm(n * 32, sd = 0.4), n)
)

fit_m   <- mfsusie(X, Y_list, L = 15, L_greedy = 5, verbose = TRUE)
fit_m_s <- mf_post_smooth(fit_m, method = "TI",
                          wavelet_filter = 1L,
                          wavelet_family = "DaubExPhase")
```

```{r mfsusie-plot, fig.width = 8, fig.height = 6}
mfsusie_plot(fit_m_s)
```

## Method

Each method takes the per-effect wavelet-domain posterior
`(mean_w, var_w) = (alpha[l, ] %*% mu[[l]][[m]],
alpha[l, ] %*% mu2[[l]][[m]] - mean_w^2)` and produces a smoothed
position-space curve plus a pointwise credible band:

- `"scalewise"` — soft-threshold `mean_w[scale_index[[s]]]` per
  wavelet scale `s` at `threshold_factor * sd_s * sqrt(2 log T)`,
  then invert via `mf_invert_dwt()`. The pointwise SD inverts
  `sqrt(var_w)` (Parseval on the orthonormal basis) and the band
  is `mean_pos +/- z(level) * sd_pos`.
- `"TI"` — cycle spinning: at every shift, run a stationary DWT
  with squared filter coefficients to compute pointwise variance,
  apply ash to the wavelet coefficients per scale, and average
  the inverse reconstructions across shifts. Credible bands come
  from the average-of-basis variance kernel.
- `"HMM"` — discrete-state hidden Markov model on per-position
  regression estimates with hub-and-spoke transitions and ash
  refinement at each EM iteration. The smoothed curve is the
  posterior mean; the band uses `abs(eff) / qnorm(1 - lfsr)` as
  a pointwise SD.
