---
title: "Practical data applications"
author: "Anjing Liu, Gao Wang, William Denault"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Practical data applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 7, fig.height = 4,
  dpi = 110
)
```

## Motivation

This vignette walks through three preprocessing and inference
options on `mfsusie()` that matter on real-world functional
fine-mapping data:

- `low_count_filter`: mask wavelet columns whose
  absolute-value median is at or below a user-supplied
  threshold. Columns that carry no signal contribute noise to
  the prior init and the SER step; masking them removes that
  noise.
- `quantile_norm`: apply a column-wise rank-based normal
  quantile transform to the wavelet matrix. Heavy-tailed
  wavelet coefficients (a common signature of sparse-count
  assays) violate the Normal-likelihood assumption underlying
  the SuSiE Bayes factor; the transform repairs the marginal.
- `small_sample_correction`: replace the per-variable
  Wakefield Normal marginal Bayes factor with the Johnson 2005
  scaled Student-t marginal, with degrees of freedom
  `df = n - 1`. Useful when `n` is small.

The motivating dataset is a chromatin-accessibility scQTL panel
by Anjing Liu and William Denault: 169 cis-regions, 84
individuals, 6 brain cell types per region (Ast, Exc, Inh, Mic,
Oli, OPC), 10240 binned coverage values per cell-type after
window expansion. The genotype and individual-level read counts
are not redistributable, so this vignette uses a small
simulated dataset shaped after the same characteristics
(`n = 40`, two outcomes, sparse-coverage columns in one
outcome, heavy-tailed contamination in the other) to exercise
the same code paths on data the reader can run locally.

```{r load, message = FALSE}
library(mfsusieR)
data(practical_fits)
practical_fits$cell_types
```

`practical_fits` carries five fits on the same simulated
dataset (`n = 40`, `p = 25`, `M = 2`, `T_m = 32`, signal SNP at
index 7), one per configuration:

```{r fit-summary}
summarize_fit <- function(fit) {
  data.frame(
    niter        = fit$niter,
    converged    = fit$converged,
    n_credible   = length(fit$sets$cs),
    pip_signal   = round(fit$pip[practical_fits$signal_snp], 3),
    pip_max_null = round(max(fit$pip[-practical_fits$signal_snp]), 3),
    pip_sum_null = round(sum(fit$pip[-practical_fits$signal_snp]), 3)
  )
}
do.call(rbind,
        lapply(c("default", "johnson", "lowcount", "qn", "combined"),
               function(nm) cbind(config = nm,
                                  summarize_fit(practical_fits[[nm]]))))
```

The default Wakefield kernel puts `PIP = 1.0` on the signal
SNP. The corrections, applied one at a time, return slightly
lower PIP values; this is the methodologically honest answer
under each correction's assumption set. Below we walk through
the three options in turn.

## Low-count filtering

`low_count_filter = c` masks wavelet-domain columns with
`median(|column|) <= c`. The masked columns contribute
`(Bhat = 0, Shat = 1)` at every IBSS iteration and are
excluded from the ash sampling pool that initializes the
mixture-of-normals prior. The threshold is in the same units as
the wavelet coefficients (so the right value depends on the
scale of the input).

In the source ATAC-seq dataset, cell types with small total
read counts have many bins where the per-individual log1p
coverage is essentially zero. Those bins do not contribute
information about the SNP-effect curve; passing them through
unmasked inflates the SER residual variance and pulls the
mixture prior toward a degenerate spike at zero.

In the simulated dataset, outcome 1 carries a block of 8 bins
at very low SNR. Setting `low_count_filter = 0.02` masks those
bins; the per-variable PIP at the signal SNP is unchanged on
this clean simulation, but the SER residual variance is no
longer inflated by the masked columns.

```{r usage-low-count, eval = FALSE}
fit <- mfsusie(X, Y_list,
               low_count_filter = 0.02,
               L = 8)
```

```{r plot-low-count, fig.cap = "PIP comparison: default vs low_count_filter."}
pip_default  <- practical_fits$default$pip
pip_lowcount <- practical_fits$lowcount$pip
plot(pip_default, pip_lowcount,
     xlim = c(0, 1), ylim = c(0, 1),
     pch = 16, cex = 0.9, col = "#1f78b4",
     xlab = "PIP (default)",
     ylab = "PIP (low_count_filter = 0.02)",
     main = "low_count_filter")
abline(0, 1, lty = 3, col = "grey50")
```

## Quantile normalization

`quantile_norm = TRUE` applies a column-wise rank-based normal
quantile transform to the per-outcome wavelet matrix before the
IBSS loop. After the transform every wavelet column is
marginally standard Gaussian, which restores the SuSiE Bayes
factor's distributional assumption.

In the source ATAC-seq dataset, log1p coverage in
sparse-coverage cell types has many zeros and a heavy right
tail. The corresponding wavelet coefficients inherit the
heavy-tailed marginal, and per-variable Bayes factors are
biased toward over-confidence at variants where the heavy
tails happen to align.

In the simulated dataset, outcome 2 carries 18% per-cell heavy
contamination. Setting `quantile_norm = TRUE` Gaussianizes the
wavelet coefficients and the SER step receives a marginal
matching the SuSiE assumption. On this dataset
`quantile_norm = TRUE` reduces PIP at the signal SNP from
`1.000` to `0.903`. The reduction reflects the more honest
uncertainty estimate once heavy-tail outliers are no longer
inflating per-variable Bayes factors.

```{r usage-qn, eval = FALSE}
fit <- mfsusie(X, Y_list,
               quantile_norm = TRUE,
               L = 8)
```

```{r plot-qn, fig.cap = "PIP comparison: default vs quantile_norm."}
pip_qn <- practical_fits$qn$pip
plot(pip_default, pip_qn,
     xlim = c(0, 1), ylim = c(0, 1),
     pch = 16, cex = 0.9, col = "#1f78b4",
     xlab = "PIP (default)",
     ylab = "PIP (quantile_norm = TRUE)",
     main = "quantile_norm")
abline(0, 1, lty = 3, col = "grey50")
```

`quantile_norm` is most useful on outcomes with heavy-tailed
wavelet coefficients and non-trivial signal density. On
outcomes that are mostly zero (e.g. cell types with very low
coverage), the rank transform turns the zeros into uniform
Gaussian noise and can dilute weak signals; pair it with
`low_count_filter` so the no-information columns are masked
before the rank transform sees them.

## Small-sample correction

`small_sample_correction = TRUE` replaces the per-variable
Wakefield Normal marginal Bayes factor with the Johnson 2005
scaled Student-t marginal. The two kernels differ only in the
density used in the per-component mixture:

$$
\text{Wakefield: } \mathrm{dnorm}\bigl(\hat{\beta}_j;\; 0,\;
   \sqrt{\sigma_k^2 + \hat{s}_j^2}\bigr),
$$

$$
\text{Johnson-t: } \mathrm{dst}\bigl(\hat{\beta}_j;\;
   \tau = (\sigma_k^2 + \hat{s}_j^2)^{-1},\; \nu = n - 1\bigr).
$$

The Student-t marginal arises from the Johnson 2005 JRSSB
construction (an objective Bayes factor for a t test); for the
multi-functional case, the same marginal is exposed in the
legacy `mvf.susie.alpha::multfsusie(..., cor_small = TRUE)`
and `fsusieR::susiF(..., cor_small = TRUE)` switches. For
small `n` the Student-t marginal places more mass in the tails
than the Normal, which dampens the per-variable Bayes factor at
variants whose `\hat{\beta} / \hat{s}` is large by chance.
Posterior moments given inclusion are unchanged; only the
variable-selection probabilities are corrected.

The mfsusieR Johnson-t kernel and the upstream
`fsusieR::log_BF(..., df = n - 1)` reference agree at machine
precision (one ULP) on the same inputs.

```{r usage-johnson, eval = FALSE}
fit <- mfsusie(X, Y_list,
               small_sample_correction = TRUE,
               L = 8)
```

```{r plot-johnson, fig.cap = "PIP comparison: default vs small_sample_correction."}
pip_johnson <- practical_fits$johnson$pip
plot(pip_default, pip_johnson,
     xlim = c(0, 1), ylim = c(0, 1),
     pch = 16, cex = 0.9, col = "#1f78b4",
     xlab = "PIP (default Wakefield)",
     ylab = "PIP (small_sample_correction = TRUE)",
     main = "small_sample_correction")
abline(0, 1, lty = 3, col = "grey50")
```

On this simulated dataset the Johnson-t correction shaves
`PIP[signal SNP]` from `1.000` to `0.994`. The shift is small
because the simulation is clean enough that the Wakefield
marginal is already nearly correct; on the source ATAC-seq
panel the difference is larger because coverage and signal are
both noisier.

## Choosing among the three options

The three options act on different points in the pipeline and
combine without conflict:

| Option | Acts on | When to use |
|---|---|---|
| `low_count_filter` | wavelet columns with `median(|.|) <= c` | sparse-coverage assays where some cell types have many near-zero bins |
| `quantile_norm` | wavelet matrix per outcome | heavy-tailed wavelet coefficients (e.g. log1p of sparse counts) |
| `small_sample_correction` | per-variable SER Bayes factor | small `n` (here `n = 40`); rule-of-thumb `n <= 100` |

A typical small-cohort sparse-count fit might combine all
three:

```{r combined, eval = FALSE}
fit <- mfsusie(X, Y_list,
               low_count_filter        = 0.02,
               quantile_norm           = TRUE,
               small_sample_correction = TRUE,
               L = 8)
```

```{r plot-combined, fig.cap = "PIP comparison: default vs all three combined."}
pip_combined <- practical_fits$combined$pip
plot(pip_default, pip_combined,
     xlim = c(0, 1), ylim = c(0, 1),
     pch = 16, cex = 0.9, col = "#1f78b4",
     xlab = "PIP (default)",
     ylab = "PIP (all three combined)",
     main = "all three combined")
abline(0, 1, lty = 3, col = "grey50")
```

On this simulation, combining all three corrections drops the
signal SNP's PIP further (from `1.000` to `0.333`). Stack the
corrections with care: each one widens the variable-selection
prior in a different way, and on a small clean dataset the
combined effect can over-correct. On the messier source
dataset, where the Wakefield default is over-confident, the
combination produces more conservative credible sets.

## Source dataset

The motivating ATAC-seq panel is described in Anjing Liu and
William Denault's MultifSuSiE manuscript. The small sample size
(`n = 84`) and the sparse, heavy-tailed coverage make this
exactly the regime where the three options matter. The panel is
not redistributable, so the simulation in this vignette is
seeded for reproducibility and the same code paths are exercised.
